---
title: "Study factors that could correlate with pseudotime in GTEx data"
author: "Clara Benoit-Pilven"
date: "`r Sys.Date()`"
output: html_document
---

```{r library, echo=FALSE}
suppressPackageStartupMessages(library(edgeR))
suppressPackageStartupMessages(library(limma))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(SummarizedExperiment))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tibble))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(readxl))
suppressPackageStartupMessages(library(naniar))
```

```{r setup, echo=FALSE}
knitr::opts_knit$set(root.dir = '~/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/')
```

# Load data

## Load pseudotime from original analysis

Load the pseudotime 

```{r load_pseudotime, echo=FALSE}
load("data/4-GTEx/1-GTEx_analysis/brain_forebrain_v8_dream_counts_voom_phenotypes_pseudotime_interaction.RData")
```

## Load full phenotype file

Load full phenotype file

```{r load_phenotype, echo=FALSE}
phenotype_full <- read.table("data/4-GTEx/1-GTEx_analysis/GTEx_Analysis_v8_Annotations_SampleAttributesDS.txt", 
                            sep = "\t", header = TRUE, quote = NULL)
# keep only RNA-seq brain samples
brain_region <- c("Brain - Amygdala", "Brain - Anterior cingulate cortex (BA24)", "Brain - Caudate (basal ganglia)", "Brain - Frontal Cortex (BA9)", "Brain - Hippocampus", "Brain - Hypothalamus", "Brain - Nucleus accumbens (basal ganglia)", "Brain - Putamen (basal ganglia)")
phenotype_brain <- phenotype_full[which(phenotype_full$SMTS == "Brain" & phenotype_full$SMAFRZE == "RNASEQ" & phenotype_full$SMTSD %in% brain_region), ]
```

Format it keeping keeping the following information:
* Sample id (SAMPID)
* Brain region (SMTSD)
* Sequencing center (SMCENTER)
* RIN (SMRIN)
* Ischemic time (SMTSISCH)
* Batch (SMNABTCH)

```{r format_phenotype, echo=FALSE}
selected_variable <- c("SAMPID", "SMTSD", "SMRIN", "SMTSISCH", "SMCENTER", "SMNABTCH")
phenotype_brain <- phenotype_brain[which(colnames(phenotype_brain) %in% selected_variable)]
names(phenotype_brain) <- c("Sample_id", "Center", "RIN", "BrainRegion", "IschemicTime", "Batch")
phenotype_brain$Sample_id <- gsub("-", ".", phenotype_brain$Sample_id)
```


## Merge pseudotime with phenotype

```{r merge, echo=FALSE}
metadata_phenotype <- merge(metadata, phenotype_brain, by.x = "SAMPLE", by.y = "Sample_id")
```


# Correlation between pseudotime and technical factors

## For original pseudotime

### Pseudotime vs brain region

Boxplot of the pseudotime per brain region.

```{r boxplot_brainRegion_pseudotime, echo=FALSE}
metadata_phenotype$BrainRegion <- as.factor(metadata_phenotype$BrainRegion)
p <- ggplot(metadata_phenotype, aes(x = BrainRegion, y = PSEUDOTIME))
p <- p + geom_boxplot()
p <- p + theme_minimal()
p <- p + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
p <- p + xlab("Forebrain region") + ylab("Pseudotime")
p
# save the plot
ggsave("correlation_pseudotimeVSbrainregion.png",
       path = "images/4-GTEx/1-GTEx_analysis/pseudotime_analysis/",
       plot = p, width = 10, height = 10, units = "cm")
```
Test difference in mean between each group.

```{r test_brainRegion_pseudotime, echo=FALSE}
# test all brain region against all brain region without accounting for paired samples
compare_means(PSEUDOTIME~BrainRegion, data = metadata_phenotype, method = "wilcox.test", paired = FALSE)
# test frontal cortex against hyppocampus accounting for paired samples
metadata_phenotype_paired <- metadata_phenotype[which(metadata_phenotype$BrainRegion %in% c("Brain - Frontal Cortex (BA9)", "Brain - Hippocampus")), ]
nbr_samplesPerSubjid <- metadata_phenotype_paired %>% group_by(SUBJID) %>% summarise(number_samples = n())
subjid2keep <- nbr_samplesPerSubjid$SUBJID[which(nbr_samplesPerSubjid$number_samples == 2)]
metadata_phenotype_paired <- metadata_phenotype_paired[which(metadata_phenotype_paired$SUBJID %in% subjid2keep), ]
compare_means(PSEUDOTIME~BrainRegion, data = metadata_phenotype_paired, method = "wilcox.test", paired = TRUE)
# linear model
lmRes <- lm(PSEUDOTIME~BrainRegion, metadata_phenotype)
summary(lmRes)
```



### Pseudotime vs RIN

Look at the correlation between the pseudotime and the RIN.

```{r cor_RIN_pseudotime, echo=FALSE}
cor.test(metadata_phenotype$PSEUDOTIME, metadata_phenotype$RIN)
```

Also check how the RIN correlates with other variables such as sex, ischemic time, batch or svas.

Correlation with sex:
```{r cor_RIN_sex, echo=FALSE}
cor.test(metadata_phenotype$SEX, metadata_phenotype$RIN)
```
 
Correlation with ischemic time:
```{r cor_RIN_ischemicTime, echo=FALSE}
cor.test(metadata_phenotype$IschemicTime, metadata_phenotype$RIN)
```
 
Correlation with svas:
```{r cor_RIN_sva, echo=FALSE}
for (i in 1:sva.obj$n.sv){
  curSva <- paste0("X", i)
  print(sprintf("SVA: %s", curSva))
  resTest <- cor.test(metadata_phenotype[, curSva], metadata_phenotype$RIN)
  print(resTest)
}
```



### Pseudotime vs Ischemic Time

Look at the correlation between the pseudotime and ischemic time.

```{r cor_ischemicTime_pseudotime, echo=FALSE}
cor.test(metadata_phenotype$PSEUDOTIME, as.numeric(metadata_phenotype$IschemicTime))
```

### Pseudotime vs Batch

Look at the correlation between the pseudotime and the batch.

```{r corPlot_batch_pseudotime, echo=FALSE}
metadata_phenotype$Batch <- as.factor(metadata_phenotype$Batch)
p <- ggplot(metadata_phenotype, aes(x = Batch, y = PSEUDOTIME))
p <- p + geom_boxplot()
p <- p + theme_minimal()
p <- p + xlab("Batch") + ylab("Pseudotime")
p
```


### Pseudotime vs svas

Look at the correlation of the pseudotime with each svas.

```{r cor_pseudotime_sva, echo=FALSE}
for (i in 1:sva.obj$n.sv){
  curSva <- paste0("X", i)
  print(sprintf("SVA: %s", curSva))
  resTest <- cor.test(metadata_phenotype[, curSva], metadata_phenotype$PSEUDOTIME)
  print(resTest)
}
```



# Session Info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

