---
title: "Pseudotime analysis in adult forebrain"
author: "Clara Benoit-Pilven"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
---

```{r library, echo=FALSE}
suppressPackageStartupMessages(library(edgeR))
suppressPackageStartupMessages(library(limma))
suppressPackageStartupMessages(library(SummarizedExperiment))
suppressPackageStartupMessages(library(ade4))
suppressPackageStartupMessages(library(factoextra))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tidyr))
```

```{r setup, echo=FALSE}
knitr::opts_knit$set(root.dir = '~/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/')
```


# Load and prepare data

## Load the counts and the phenotype data

```{r load_data, echo=FALSE}
load("data/4-GTEx/1-GTEx_analysis/brain_forebrain_v8_dream_counts_phenotypes.RData")
```

## Load pseudotime results

```{r load_pseudotime, echo=FALSE}
load("data/4-GTEx/1-GTEx_analysis/GTEx_v8_forebrain_pseudotime_analysis.RData")
```

## Add age information

We  also need to load the table containing the age of the individuals.

```{r load_age, echo=FALSE}
phenoAge <- read.table("data/4-GTEx/1-GTEx_analysis/GTEx_Analysis_v8_Annotations_SubjectPhenotypesDS.txt",
                       sep = "\t", header = TRUE)
```

And then merge the age information in the phenotype data frame.

```{r merge_pheno, echo=FALSE}
metadata$sample <- rownames(metadata)
metadata <- merge(metadata, phenoAge[, c(1,3)], by = "SUBJID", all.x = TRUE)
names(metadata) <- c("SUBJID", "SEX", "DTHHRDY", "TISSUE", "pseudotime", "Sample", "AGE")
metadata$AGE <- as.factor(metadata$AGE)
```


# Plots

## Histogram

Plot the distribution of the pseudotime.

```{r distrib_pseudotime, echo=FALSE}
hist(metadata$pseudotime, main = "Histogram of pseudotime", xlab = "Pseudotime")
png("images/4-GTEx/1-GTEx_analysis/pseudotime_analysis/distrib_pseudotime.png",
    width = 900, height = 800)
hist(metadata$pseudotime, main = "Histogram of pseudotime", xlab = "Pseudotime")
dev.off() 
```

## PCA plots

Next, we plot the samples in a PCA plot coloring by pseudotime for both sexes together and for each sex separatly

```{r select_counts_PCA, echo=FALSE}
# select higly variable genes
variance <- rowVars(x$counts)
ntop <- min(2000, dim(x$counts)[1])
selectntop <- order(variance, decreasing=TRUE)[seq_len(ntop)]
selectedCounts_PCA <- x$counts[selectntop,]
```

```{r do_PCA, echo=FALSE}
pcaHVG <- dudi.pca(t(selectedCounts_PCA), scannf = FALSE, nf = 10)
varExplained <- pcaHVG$eig/sum(pcaHVG$eig)*100
pcaHVG.ind <- get_pca_ind(pcaHVG)
pcaHVG.var <- get_pca_var(pcaHVG)
```

```{r plot_PCA_pseudotime_color, echo=FALSE}
p <- ggplot(pcaHVG.ind$coord, aes(x = Dim.1, y = Dim.2))
p <- p + geom_point(aes(color = metadata$pseudotime))
p <- p + labs(title = "PCA" , 
              x = sprintf("PC%s (%s %%)", 1, round(varExplained[1])), 
              y = sprintf("PC%s (%s %%)", 2, round(varExplained[2])))
p <- p + theme_minimal()
p <- p + scale_color_gradient(low = "blueviolet", high = "gold", name = "pseudotime")
p
# save plot
ggsave("PCA_1vs2.png",
       path = "images/4-GTEx/1-GTEx_analysis/PCA/",
       plot = p, width = 12, height = 10, units = "cm", create.dir = TRUE)
```

```{r plot_PCA_pseudotime_color_2vs3, echo=FALSE}
p <- ggplot(pcaHVG.ind$coord, aes(x = Dim.2, y = Dim.3))
p <- p + geom_point(aes(color = metadata$pseudotime))
p <- p + labs(title = "PCA" , 
              x = sprintf("PC%s (%s %%)", 2, round(varExplained[2])), 
              y = sprintf("PC%s (%s %%)", 3, round(varExplained[3])))
p <- p + theme_minimal()
p <- p + scale_color_gradient(low = "blueviolet", high = "gold", name = "pseudotime")
p
# save plot
ggsave("PCA_2vs3.png",
       path = "images/4-GTEx/1-GTEx_analysis/PCA/",
       plot = p, width = 12, height = 10, units = "cm")
```


Faceting using the sex of the samples.

```{r plot_PCA_pseudotime_color_sex_facet, echo=FALSE}
dfPlot <- merge(pcaHVG.ind$coord, metadata, by.x = 0, by.y = "Sample")
p <- ggplot(dfPlot, aes(x = Dim.1, y = Dim.2))
p <- p + geom_point(aes(color = pseudotime))
p <- p + labs(title = "PCA" , 
              x = sprintf("PC%s (%s %%)", 1, round(varExplained[1])), 
              y = sprintf("PC%s (%s %%)", 2, round(varExplained[2])))
p <- p + theme_minimal()
p <- p + scale_color_gradient(low = "blueviolet", high = "gold", name = "pseudotime")
p <- p + facet_grid(. ~ dfPlot$SEX)
p
# save plot
ggsave("PCA_1vs2_sex.png",
       path = "images/4-GTEx/1-GTEx_analysis/PCA/",
       plot = p, width = 15, height = 10, units = "cm")
```

```{r plot_PCA_pseudotime_color_sex_facet_2vs3, echo=FALSE}
dfPlot <- merge(pcaHVG.ind$coord, metadata, by.x = 0, by.y = "Sample")
p <- ggplot(dfPlot, aes(x = Dim.2, y = Dim.3))
p <- p + geom_point(aes(color = pseudotime))
p <- p + labs(title = "PCA" , 
              x = sprintf("PC%s (%s %%)", 2, round(varExplained[2])), 
              y = sprintf("PC%s (%s %%)", 3, round(varExplained[3])))
p <- p + theme_minimal()
p <- p + scale_color_gradient(low = "blueviolet", high = "gold", name = "pseudotime")
p <- p + facet_grid(. ~ dfPlot$SEX)
p
# save plot
ggsave("PCA_2vs3_sex.png",
       path = "images/4-GTEx/1-GTEx_analysis/PCA/",
       plot = p, width = 15, height = 10, units = "cm")
```

Faceting using the tissue of origin of the samples.

```{r plot_PCA_pseudotime_color_tissue_facet, echo=FALSE}
dfPlot <- merge(pcaHVG.ind$coord, metadata, by.x = 0, by.y = "Sample")
p <- ggplot(dfPlot, aes(x = Dim.1, y = Dim.2))
p <- p + geom_point(aes(color = pseudotime))
p <- p + labs(title = "PCA" , 
              x = sprintf("PC%s (%s %%)", 1, round(varExplained[1])), 
              y = sprintf("PC%s (%s %%)", 2, round(varExplained[2])))
p <- p + theme_minimal()
p <- p + scale_color_gradient(low = "blueviolet", high = "gold", name = "pseudotime")
p <- p + facet_grid(. ~ dfPlot$TISSUE)
p
# save plot
ggsave("PCA_1vs2_tissue.png",
       path = "images/4-GTEx/1-GTEx_analysis/PCA/",
       plot = p, width = 30, height = 10, units = "cm")
```

Faceting using the cause of death of the individuals.

```{r plot_PCA_pseudotime_color_causeDeath_facet, echo=FALSE}
dfPlot <- merge(pcaHVG.ind$coord, metadata, by.x = 0, by.y = "Sample")
p <- ggplot(dfPlot, aes(x = Dim.1, y = Dim.2))
p <- p + geom_point(aes(color = pseudotime))
p <- p + labs(title = "PCA" , 
              x = sprintf("PC%s (%s %%)", 1, round(varExplained[1])), 
              y = sprintf("PC%s (%s %%)", 2, round(varExplained[2])))
p <- p + theme_minimal()
p <- p + scale_color_gradient(low = "blueviolet", high = "gold", name = "pseudotime")
p <- p + facet_grid(. ~ dfPlot$DTHHRDY)
p
# save plot
ggsave("PCA_1vs2_causeDeath.png",
       path = "images/4-GTEx/1-GTEx_analysis/PCA/",
       plot = p, width = 30, height = 10, units = "cm")
```

Do PCA plots colored by the different covariates available: sex, brain subregion, cause of death and age group.

```{r PCA_sex, echo=FALSE}
p <- ggplot(pcaHVG.ind$coord, aes(x = Dim.1, y = Dim.2))
p <- p + geom_point(aes(color = SEX))
p <- p + labs(title = "PCA" , 
              x = sprintf("PC%s (%s %%)", 1, round(varExplained[1])), 
              y = sprintf("PC%s (%s %%)", 2, round(varExplained[2])))
p <- p + theme_minimal()
p
# save plot
ggsave("PCA_sex_1vs2.png",
       path = "images/4-GTEx/1-GTEx_analysis/PCA/",
       plot = p, width = 12, height = 10, units = "cm")
```

```{r PCA_tissue, echo=FALSE}
p <- ggplot(pcaHVG.ind$coord, aes(x = Dim.1, y = Dim.2))
p <- p + geom_point(aes(color = TISSUE))
p <- p + labs(title = "PCA" , 
              x = sprintf("PC%s (%s %%)", 1, round(varExplained[1])), 
              y = sprintf("PC%s (%s %%)", 2, round(varExplained[2])))
p <- p + theme_minimal()
p
# save plot
ggsave("PCA_tissue_1vs2.png",
       path = "images/4-GTEx/1-GTEx_analysis/PCA/",
       plot = p, width = 12, height = 10, units = "cm")
```

```{r PCA_causeDeath, echo=FALSE}
p <- ggplot(pcaHVG.ind$coord, aes(x = Dim.1, y = Dim.2))
p <- p + geom_point(aes(color = DTHHRDY))
p <- p + labs(title = "PCA" , 
              x = sprintf("PC%s (%s %%)", 1, round(varExplained[1])), 
              y = sprintf("PC%s (%s %%)", 2, round(varExplained[2])))
p <- p + theme_minimal()
p
# save plot
ggsave("PCA_causeDeath_1vs2.png",
       path = "images/4-GTEx/1-GTEx_analysis/PCA/",
       plot = p, width = 12, height = 10, units = "cm")
```

```{r PCA_age, echo=FALSE}
p <- ggplot(pcaHVG.ind$coord, aes(x = Dim.1, y = Dim.2))
p <- p + geom_point(aes(color = metadata$AGE))
p <- p + labs(title = "PCA" , 
              x = sprintf("PC%s (%s %%)", 1, round(varExplained[1])), 
              y = sprintf("PC%s (%s %%)", 2, round(varExplained[2])))
p <- p + theme_minimal()
p
# save plot
ggsave("PCA_age_1vs2.png",
       path = "images/4-GTEx/1-GTEx_analysis/PCA/",
       plot = p, width = 12, height = 10, units = "cm")
```


# Test pseudotime consistency per individuals

Check that for each individual, the pseudotimes of the different samples are similar.

```{r check_pseudotime_individuals, echo=FALSE}
sdPseudotime_ind <- metadata %>% group_by(SUBJID) %>% summarise(sd = sd(pseudotime))
meanSdPseudotime <- mean(sdPseudotime_ind$sd, na.rm = TRUE)
# Is the sd as big if we permute the pseudotimes?
meanSdPseudotime_permut <- c()
for (i in 1:1000){
  permutated_pseudotime <- sample(metadata$pseudotime)
  tmp_df <- data.frame(SUBJID = metadata$SUBJID, pseudotime = permutated_pseudotime)
  tmp_sdPseudotime_ind <- tmp_df %>% group_by(SUBJID) %>% summarise(sd = sd(pseudotime))
  meanSdPseudotime_permut <- c(meanSdPseudotime_permut, mean(tmp_sdPseudotime_ind$sd, na.rm = TRUE))
}
meanSdPseudotime_permut <- as.data.frame(meanSdPseudotime_permut)
# pvalue
(1 + sum(meanSdPseudotime_permut<meanSdPseudotime))/1000
# Plot
p <- ggplot(meanSdPseudotime_permut, aes(x = meanSdPseudotime_permut))
p <- p + geom_histogram(binwidth = 0.01, fill = "grey", color = "black")
p <- p + geom_vline(xintercept = meanSdPseudotime,
                    color = "red")
p <- p + labs(x = "SD of the pseudotime per individuals")
p <- p + theme_minimal()
p
# save plot
ggsave("permutation_sd_pseudotime.png",
       path = "images/4-GTEx/1-GTEx_analysis/pseudotime_analysis/",
       plot = p, width = 15, height = 10, units = "cm", create.dir = TRUE)
```


# Correlation of pseudotime with age group

Convert the developmental stages to a continuous variable.

```{r convert2continuous, echo=FALSE}
metadata$AgeCont <- as.integer(metadata$AGE)
```

### For both sexes

Compute Kendall correlation between the pseudotime and the new continuous developmental stage.

```{r correlation_kendall, echo=FALSE}
corTestRes_Kendall <- cor.test(metadata$AgeCont, metadata$pseudotime, method = "kendall")
corTestRes_Kendall
```

And also do a scatter plot of the pseudotime and the continuous developmental stage.

```{r scatter_plot, echo=FALSE}
p <- ggplot(metadata, aes(AGE, pseudotime))
p <- p + geom_boxplot(fill = "#DCDCDC", outlier.shape = NA)
p <- p + theme_minimal() + xlab("Age groups")
p <- p + geom_jitter(aes(color = SEX), position=position_jitter(0.2), show.legend = FALSE)
p <- p + annotate("text", x = 1.5, y = 1.1, label = sprintf("tau = %.2f\n p-value = %.2e", corTestRes_Kendall$estimate, corTestRes_Kendall$p.value))
p <- p + scale_color_manual(values = c("1" = "darkblue", "2" = "darkred"))
p
# save plot
ggsave("boxplot_age_vs_pseudotime.png", 
       path = "images/4-GTEx/1-GTEx_analysis/pseudotime_analysis/",
       plot = p, width = 15, height = 10, units = "cm")
```

Test difference in pseudotime variance between males and females.

```{r test_variance_sex, echo=FALSE}
var.test(pseudotime ~ SEX, metadata, 
         alternative = "two.sided")
```

### For males only

Select only male samples.

```{r select_males, echo=FALSE}
metadata_males <- metadata[which(metadata$SEX == 1),]
```

Compute Kendall correlation between the pseudotime and the new continuous age groups.

```{r correlation_kendall_males, echo=FALSE}
corTestRes_Kendall <- cor.test(metadata$AgeCont, metadata$pseudotime, method = "kendall")
corTestRes_Kendall
```


And also do a scatter plot of the pseudotime and the continuous developmental stage.

```{r scatter_plot_males, echo=FALSE}
p <- ggplot(metadata_males, aes(AGE, pseudotime))
p <- p + geom_boxplot(fill = "#a1c1e6")
p <- p + theme_minimal()
p <- p + geom_jitter(position=position_jitter(0.2))
p <- p + annotate("text", x = 1.5, y = 1.1, label = sprintf("tau = %.2f\n p-value = %.2e", corTestRes_Kendall$estimate, corTestRes_Kendall$p.value))
p
# save plot
ggsave("boxplot_age_vs_pseudotime_males.png",
       path = "images/4-GTEx/1-GTEx_analysis/pseudotime_analysis/",
       plot = p, width = 15, height = 10, units = "cm")
```

### For females only

Select only female samples.

```{r select_females, echo=FALSE}
metadata_females <- metadata[which(metadata$SEX == 2),]
```

Compute Kendall correlation between the pseudotime and the new continuous age groups.

```{r correlation_kendall_females, echo=FALSE}
corTestRes_Kendall <- cor.test(metadata$AgeCont, metadata$pseudotime, method = "kendall")
corTestRes_Kendall
```

And also do a scatter plot of the pseudotime and the continuous developmental stage.

```{r scatter_plot_females, echo=FALSE}
p <- ggplot(metadata_females, aes(AGE, pseudotime))
p <- p + geom_boxplot(fill = "#e0a3aa")
p <- p + theme_minimal()
p <- p + geom_jitter(position=position_jitter(0.2))
p <- p + annotate("text", x = 1.5, y = 1.1, label = sprintf("tau = %.2f\n p-value = %.2e", corTestRes_Kendall$estimate, corTestRes_Kendall$p.value))
p
# save plot
ggsave("boxplot_devStages_vs_pseudotime_females.png",
       path = "images/4-GTEx/1-GTEx_analysis/pseudotime_analysis/",
       plot = p, width = 15, height = 10, units = "cm")
```


# Session Info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
