---
title: "Visualize GTEx DE results with age as covariate"
author: "Clara Benoit-Pilven"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
---

```{r load_libraries, echo=FALSE}
suppressPackageStartupMessages( library(edgeR) )
suppressPackageStartupMessages( library(limma) )
suppressPackageStartupMessages( library(SummarizedExperiment) )
suppressPackageStartupMessages( library(EnhancedVolcano) )
suppressPackageStartupMessages( library(tidyverse) )
suppressPackageStartupMessages( library(ggpubr) )
suppressPackageStartupMessages( library(matrixStats) )
suppressPackageStartupMessages( library(ComplexHeatmap) )
suppressPackageStartupMessages( library(variancePartition) )
```

```{r setup, echo=FALSE}
knitr::opts_knit$set(root.dir = '~/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/')
```

Parameters
```{r thresholds, echo=FALSE}
threshold_qvalue <- 0.01
```


# Load and prepare DE results

## Load DE results

```{r load_sexDE_results, echo=FALSE}
de_results <- read.table("data/4-GTEx/1-GTEx_analysis/brain_forebrain_v8_VoomDream_topTable_with_sva_ind_age_interaction.txt",
                         header = TRUE)
```

## Filter DE results

From the full list of sex-DE genes, remove XIST and the protein-coding genes on the Y-chromosome.

```{r remove_training_genes, echo=FALSE}
genes2remove <- c("XIST", as.character(de_results$gene_name[which(de_results$chr == "chrY" & de_results$gene_type == "protein_coding")]))
de_results_filtered <- de_results[which(!de_results$gene_name %in% genes2remove),]
```

Create a data frame containing only the results for the autosomes.

```{r autosomes, echo=FALSE}
chrNonAuto <- c("chrY", "chrX", "chrM")
de_results_autosomes <- de_results_filtered[which(!de_results_filtered$chr %in% chrNonAuto),]
```

# Statistics

## Number of DE genes

Number of sex-DE significant genes:

```{r nbr_sexDE, echo=FALSE}
nbr_sexDE_signif <- sum(de_results_filtered$qvalue < threshold_qvalue & abs(de_results_filtered$logFC) > 0)
percent_sexDE_signif <- nbr_sexDE_signif/length(de_results_filtered$gene_name)*100
print(sprintf("Number of sex-DE genes = %s (representing %.3f %% of the genes)", nbr_sexDE_signif, percent_sexDE_signif))
# male-biased genes
nbr_maleBiased_signif <- length(de_results_filtered$gene_name[which(de_results_filtered$logFC < 0 & de_results_filtered$qvalue < threshold_qvalue)])
percent_maleBiased_signif <- nbr_maleBiased_signif/nbr_sexDE_signif*100
print(sprintf("Number of male-biased genes = %s (%.3f %% of sex-DE genes)", nbr_maleBiased_signif, percent_maleBiased_signif))
# female_biased genes
nbr_femaleBiased_signif <- length(de_results_filtered$gene_name[which(de_results_filtered$logFC > 0 & de_results_filtered$qvalue < threshold_qvalue)])
percent_femaleBiased_signif <- nbr_femaleBiased_signif/nbr_sexDE_signif*100
print(sprintf("Number of female-biased genes = %s (%.3f %% of sex-DE genes)", nbr_femaleBiased_signif, percent_femaleBiased_signif))
```

```{r nbr_sexDE_auto, echo=FALSE}
nbr_sexDE_auto_signif <- sum(de_results_autosomes$qvalue < threshold_qvalue & abs(de_results_autosomes$logFC) > 0)
percent_sexDE_auto_signif <- nbr_sexDE_auto_signif/length(de_results_autosomes$gene_name)*100
print(sprintf("Number of autosomal sex-DE genes = %s (representing %.3f %% of the genes)", nbr_sexDE_auto_signif, percent_sexDE_auto_signif))
# male-biased genes
nbr_maleBiased_auto_signif <- length(de_results_autosomes$gene_name[which(de_results_autosomes$logFC < 0 & de_results_autosomes$qvalue < threshold_qvalue)])
percent_maleBiased_auto_signif <- nbr_maleBiased_auto_signif/nbr_sexDE_auto_signif*100
print(sprintf("Number of autosomal male-biased genes = %s (%.3f %% of sex-DE genes)", nbr_maleBiased_auto_signif, percent_maleBiased_auto_signif))
# female_biased genes
nbr_femaleBiased_auto_signif <- length(de_results_autosomes$gene_name[which(de_results_autosomes$logFC > 0 & de_results_autosomes$qvalue < threshold_qvalue)])
percent_femaleBiased_auto_signif <- nbr_femaleBiased_auto_signif/nbr_sexDE_auto_signif*100
print(sprintf("Number of autosomal female-biased genes = %s (%.3f %% of sex-DE genes)", nbr_femaleBiased_auto_signif, percent_femaleBiased_auto_signif))
```

## Comparison with original sex-DE genes using pseudotime as covariate

Load sex-DE genes from original analysis.

```{r load_sexDE_original, echo=FALSE}
de_results_original <- read.table("data/4-GTEx/1-GTEx_analysis/brain_forebrain_v8_VoomDream_topTable_with_sva_ind_pseudotime_interaction.txt",
                                  header = TRUE)
# filter results
genes2remove <- c("XIST", as.character(de_results_original$gene_name[which(de_results_original$chr == "chrY" & de_results_original$gene_type == "protein_coding")]))
de_results_original_filtered <- de_results_original[which(!de_results_original$gene_name %in% genes2remove),]
```

Check how many sex-DE genes from this new analysis were also found sex-DE in the original one.

```{r overlap_original_sexDE, echo=FALSE}
overlap_genes <- de_results_filtered[which(de_results_filtered$gene_id[which(de_results_filtered$qvalue < threshold_qvalue)] %in% de_results_original_filtered$gene_id[which(de_results_original_filtered$qvalue < threshold_qvalue)]), ]
length(overlap_genes$gene_id)
```

Only 12/23 (52%) are also found in the analysis using the age bins. All these 12 genes are on the sex chromosomes.

# Plots

## Barplot representing the number of DE genes

Create a barplot of each of the 3 analysis representing the number of DE genes (up and down regulated separatly).

```{r function_barplot, echo=FALSE}
barplot <- function(data, legend = TRUE, nrowLegend = 4, myTitle = "Nbr sex-DE genes",
                    colors = c("Male-other" = "#053061", "Male-sex" = "#99b3ff", 
                               "Female-other" = "#B2182B", "Female-sex" = "#f3a5ae"),
                    labels = c("Female-other" = "Female autosomal genes", "Female-sex" = "Female sex chr. genes",
                               "Male-other" = "Male autosomal genes", "Male-sex" = "Male sex chr. genes")){
  p <- ggplot(data, aes(x = variable, y = nbr_genes, fill = coloring))
  p <- p + geom_bar(stat = "identity")
  p <- p + scale_fill_manual(values = colors, labels = labels)
  p <- p + theme_bw() + labs(x = "", y = "")
  p <- p + ggtitle(myTitle)
  p <- p + theme(legend.position="bottom", legend.title = element_blank()) 
  p <- p + guides(fill = guide_legend(nrow=nrowLegend, byrow=TRUE))
  if (!legend){
    p <- p +  guides(fill = "none")
  }
  return(p)
}
```

Sex-DE genes

```{r barplot_sexDE, echo=FALSE}
# prepare data
sexChr <- c("chrY", "chrX")
barplot_data_sexDE <- data.frame(variable = c("Female", "Female", "Male", "Male"), 
                           chr = c("other", "sex", "other", "sex"), 
                           nbr_genes = c(length(de_results_autosomes$gene_name[which(de_results_autosomes$qvalue < threshold_qvalue & de_results_autosomes$logFC > 0)]),
                                         length(de_results_filtered$gene_name[which((de_results_filtered$chr %in% sexChr) & de_results_filtered$qvalue < threshold_qvalue & de_results_filtered$logFC > 0)]),
                                         length(de_results_autosomes$gene_name[which(de_results_autosomes$qvalue < threshold_qvalue & de_results_autosomes$logFC < -0)]),
                                         length(de_results_filtered$gene_name[which((de_results_filtered$chr %in% sexChr) & de_results_filtered$qvalue < threshold_qvalue & de_results_filtered$logFC < 0)]))
)

barplot_data_sexDE$coloring <- paste(barplot_data_sexDE$variable, barplot_data_sexDE$chr, sep = "-")
barplot_data_sexDE <- barplot_data_sexDE[order(barplot_data_sexDE$nbr_genes, decreasing = TRUE),] # order data
barplot_data_sexDE$coloring <- factor(barplot_data_sexDE$coloring, levels = c("Male-sex", "Female-sex", "Male-other" , "Female-other"))
barplot_data_sexDE$variable <- factor(barplot_data_sexDE$variable, levels = c("Male", "Female"))

# barplot
p <- barplot(barplot_data_sexDE)
p
# save plot
namePlot <- "barplot_nbr_sexDE_genes.png"
ggsave(plot = p, namePlot, path = "images/4-GTEx/1-GTEx_analysis/age_analysis/barplot/",
       width = 8, height = 17, units = "cm", create.dir = TRUE)
```


## Volcanoplot

Sex-DE genes.

```{r volcanoplot_sexDE_auto, echo=FALSE}
keyvals <- rep('grey75', nrow(de_results_autosomes))
names(keyvals) <- rep('NS', nrow(de_results_autosomes))

keyvals[which(de_results_autosomes$logFC < 0 & de_results_autosomes$qvalue < threshold_qvalue)] <- 'darkblue'
names(keyvals)[which(de_results_autosomes$logFC  < 0 & de_results_autosomes$qvalue < threshold_qvalue)] <- 'Signif. males'

keyvals[which(de_results_autosomes$logFC > 0 & de_results_autosomes$qvalue < threshold_qvalue)] <- 'darkred'
names(keyvals)[which(de_results_autosomes$logFC > 0 & de_results_autosomes$qvalue < threshold_qvalue)] <- 'Signif. females'

unique(keyvals)
unique(names(keyvals))

p <- EnhancedVolcano(de_results_autosomes,
                     lab = as.character(de_results_autosomes$gene_name),
                     x = 'logFC',
                     y = 'qvalue',
                     xlim = c(-1.5,1.5),
                     FCcutoff = 2,
                     pCutoff = 10E-15,
                     cutoffLineType = 'blank',
                     colCustom = keyvals,
                     legendPosition = "none",
                     ylab = expression('-Log'[10]*' q-value'),
                     subtitle = "sex-DE analysis in HDBR",
                     caption = "autosomal genes")
p

# save plot
namePlot <- 'volcanoplot_colored_HDBR_dream_forebrain_pseudotime_sva_ind_interaction_sexDE_auto.png'
ggsave(namePlot, path = "images/4-GTEx/1-GTEx_analysis/age_analysis/volcanoplot/",
       width = 12, height = 15, units = "cm", create.dir = TRUE)
```


## Variance partition plot


Load voom normalized counts and phenotypes.

```{r load_data, echo=FALSE}
load("data/4-GTEx/1-GTEx_analysis/brain_forebrain_v8_dream_counts_voom_phenotypes_age_interaction.RData")
```

Compute variance attributable to each variable

```{r varpart, echo=FALSE}
form <- reformulate(c("SEX", "(1 | AGE)", "SEX:AGE", "(1 | SUBJID)", paste0("X", 1:sva.obj$n.sv)))
varPart <- fitExtractVarPartModel(y, form, metadata)
# sort variables (i.e. columns) by median fraction of variance explained
vp <- sortCols( varPart )
```

```{r mean_varexplained, echo=FALSE}
colMeans(vp)
```


Violin plot of contribution of each variable to total variance

```{r violin_plot, echo=FALSE}
p <- plotVarPart( vp )
p <- p + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
p
ggsave("violinPlot_variancePartition.png", p,
       path = "images/4-GTEx/1-GTEx_analysis/age_analysis/variancePartition/",
       width = 20, height = 12, units = "cm", create.dir = TRUE)
```


# Session Info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
