---
title: "Visualize GTEx DE results"
author: "Clara Benoit-Pilven"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
---

```{r load_libraries, echo=FALSE}
suppressPackageStartupMessages( library(edgeR) )
suppressPackageStartupMessages( library(limma) )
suppressPackageStartupMessages( library(SummarizedExperiment) )
suppressPackageStartupMessages( library(EnhancedVolcano) )
suppressPackageStartupMessages( library(tidyverse) )
suppressPackageStartupMessages( library(ggpubr) )
suppressPackageStartupMessages( library(matrixStats) )
suppressPackageStartupMessages( library(ComplexHeatmap) )
suppressPackageStartupMessages( library(variancePartition) )
```

```{r setup, echo=FALSE}
knitr::opts_knit$set(root.dir = '~/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/')
```

Parameters
```{r thresholds, echo=FALSE}
threshold_qvalue <- 0.01
```


# Load and prepare DE results

## Load DE results

```{r load_sexDE_results, echo=FALSE}
# load pseudotime data
load("data/4-GTEx/1-GTEx_analysis/GTEx_v8_forebrain_pseudotime_analysis.RData")
de_results <- read.table("data/4-GTEx/1-GTEx_analysis/brain_forebrain_v8_VoomDream_topTable_with_sva_ind_pseudotime_interaction.txt",
                         header = TRUE)
de_results_interaction <- read.table("data/4-GTEx/1-GTEx_analysis/brain_forebrain_v8_VoomDream_topTable_with_sva_ind_pseudotime_interaction_term.txt", 
                                     header = TRUE)
de_results_pseudotime <- read.table("data/4-GTEx/1-GTEx_analysis/brain_forebrain_v8_VoomDream_topTable_with_sva_ind_pseudotime_interaction_pseudotime.txt", 
                                           header = TRUE)
```

## Filter DE results

From the full list of sex-DE genes, remove XIST and the protein-coding genes on the Y-chromosome.

```{r remove_training_genes, echo=FALSE}
genes2remove <- c("XIST", as.character(de_results$gene_name[which(de_results$chr == "chrY" & de_results$gene_type == "protein_coding")]))
de_results_filtered <- de_results[which(!de_results$gene_name %in% genes2remove),]
genes2remove <- c("XIST", as.character(de_results_interaction$gene_name[which(de_results_interaction$chr == "chrY" & de_results_interaction$gene_type == "protein_coding")]))
de_results_interaction_filtered <- de_results_interaction[which(!de_results_interaction$gene_name %in% genes2remove),]
genes2remove <- c("XIST", as.character(de_results_pseudotime$gene_name[which(de_results_pseudotime$chr == "chrY" & de_results_pseudotime$gene_type == "protein_coding")]))
de_results_pseudotime_filtered <- de_results_pseudotime[which(!de_results_pseudotime$gene_name %in% genes2remove),]
```

From the pseudotime and interaction-DE genes remove the genes used to compute the pseudotime.

```{r remove_pseudotime_genes, echo=FALSE}
genes2remove <- rownames(selectedCounts)
de_results_interaction_filtered <- de_results_interaction_filtered[which(!de_results_interaction_filtered$gene_name %in% genes2remove),]
de_results_pseudotime_filtered <- de_results_pseudotime_filtered[which(!de_results_pseudotime_filtered$gene_name %in% genes2remove),]
```

Create a data frame containing only the results for the autosomes.

```{r autosomes, echo=FALSE}
chrNonAuto <- c("chrY", "chrX", "chrM")
de_results_autosomes <- de_results_filtered[which(!de_results_filtered$chr %in% chrNonAuto),]
de_results_interaction_autosomes <- de_results_interaction_filtered[which(!de_results_interaction_filtered$chr %in% chrNonAuto),]
de_results_pseudotime_autosomes <- de_results_pseudotime_filtered[which(!de_results_pseudotime_filtered$chr %in% chrNonAuto),]
```

# Statistics

## Number of DE genes

Number of sex-DE significant genes:

```{r nbr_sexDE, echo=FALSE}
nbr_sexDE_signif <- sum(de_results_filtered$qvalue < threshold_qvalue & abs(de_results_filtered$logFC) > 0)
percent_sexDE_signif <- nbr_sexDE_signif/length(de_results_filtered$gene_name)*100
print(sprintf("Number of sex-DE genes = %s (representing %.3f %% of the genes)", nbr_sexDE_signif, percent_sexDE_signif))
# male-biased genes
nbr_maleBiased_signif <- length(de_results_filtered$gene_name[which(de_results_filtered$logFC < 0 & de_results_filtered$qvalue < threshold_qvalue)])
percent_maleBiased_signif <- nbr_maleBiased_signif/nbr_sexDE_signif*100
print(sprintf("Number of male-biased genes = %s (%.3f %% of sex-DE genes)", nbr_maleBiased_signif, percent_maleBiased_signif))
# female_biased genes
nbr_femaleBiased_signif <- length(de_results_filtered$gene_name[which(de_results_filtered$logFC > 0 & de_results_filtered$qvalue < threshold_qvalue)])
percent_femaleBiased_signif <- nbr_femaleBiased_signif/nbr_sexDE_signif*100
print(sprintf("Number of female-biased genes = %s (%.3f %% of sex-DE genes)", nbr_femaleBiased_signif, percent_femaleBiased_signif))
```

```{r nbr_sexDE_auto, echo=FALSE}
nbr_sexDE_auto_signif <- sum(de_results_autosomes$qvalue < threshold_qvalue & abs(de_results_autosomes$logFC) > 0)
percent_sexDE_auto_signif <- nbr_sexDE_auto_signif/length(de_results_autosomes$gene_name)*100
print(sprintf("Number of autosomal sex-DE genes = %s (representing %.3f %% of the genes)", nbr_sexDE_auto_signif, percent_sexDE_auto_signif))
# male-biased genes
nbr_maleBiased_auto_signif <- length(de_results_autosomes$gene_name[which(de_results_autosomes$logFC < 0 & de_results_autosomes$qvalue < threshold_qvalue)])
percent_maleBiased_auto_signif <- nbr_maleBiased_auto_signif/nbr_sexDE_auto_signif*100
print(sprintf("Number of autosomal male-biased genes = %s (%.3f %% of sex-DE genes)", nbr_maleBiased_auto_signif, percent_maleBiased_auto_signif))
# female_biased genes
nbr_femaleBiased_auto_signif <- length(de_results_autosomes$gene_name[which(de_results_autosomes$logFC > 0 & de_results_autosomes$qvalue < threshold_qvalue)])
percent_femaleBiased_auto_signif <- nbr_femaleBiased_auto_signif/nbr_sexDE_auto_signif*100
print(sprintf("Number of autosomal female-biased genes = %s (%.3f %% of sex-DE genes)", nbr_femaleBiased_auto_signif, percent_femaleBiased_auto_signif))
```

Number of interaction-DE significant genes:

```{r nbr_interactionDE, echo=FALSE}
nbr_sexDE_interaction_signif <- sum(de_results_interaction_filtered$qvalue < threshold_qvalue & abs(de_results_interaction_filtered$logFC) > 0)
percent_sexDE_interaction_signif <- nbr_sexDE_interaction_signif/length(de_results_interaction_filtered$gene_name)*100
print(sprintf("Number of interaction-DE genes = %s (representing %.3f %% of the genes)", nbr_sexDE_interaction_signif, percent_sexDE_interaction_signif))
# male-biased genes
nbr_maleBiased_interaction_signif <- length(de_results_interaction_filtered$gene_name[which(de_results_interaction_filtered$logFC < 0 & de_results_interaction_filtered$qvalue < threshold_qvalue)])
percent_maleBiased_interaction_signif <- nbr_maleBiased_interaction_signif/nbr_sexDE_interaction_signif*100
print(sprintf("Number of down interaction genes = %s (%.3f %% of sex-DE genes)", nbr_maleBiased_interaction_signif, percent_maleBiased_interaction_signif))
# female_biased genes
nbr_femaleBiased_interaction_signif <- length(de_results_interaction_filtered$gene_name[which(de_results_interaction_filtered$logFC > 0 & de_results_interaction_filtered$qvalue < threshold_qvalue)])
percent_femaleBiased_interaction_signif <- nbr_femaleBiased_interaction_signif/nbr_sexDE_interaction_signif*100
print(sprintf("Number of up interaction genes = %s (%.3f %% of sex-DE genes)", nbr_femaleBiased_interaction_signif, percent_femaleBiased_interaction_signif))
```

```{r nbr_interactionDE_auto, echo=FALSE}
nbr_sexDE_interaction_auto_signif <- sum(de_results_interaction_autosomes$qvalue < threshold_qvalue & abs(de_results_interaction_autosomes$logFC) > 0)
percent_sexDE_interaction_auto_signif <- nbr_sexDE_interaction_auto_signif/length(de_results_interaction_autosomes$gene_name)*100
print(sprintf("Number of autosomal interaction-DE genes = %s (representing %.3f %% of the genes)", nbr_sexDE_interaction_auto_signif, percent_sexDE_interaction_auto_signif))
# male-biased genes
nbr_maleBiased_interaction_auto_signif <- length(de_results_interaction_autosomes$gene_name[which(de_results_interaction_autosomes$logFC < 0 & de_results_interaction_autosomes$qvalue < threshold_qvalue)])
percent_maleBiased_interaction_auto_signif <- nbr_maleBiased_interaction_auto_signif/nbr_sexDE_interaction_auto_signif*100
print(sprintf("Number of autosomal down interaction genes = %s (%.3f %% of sex-DE genes)", nbr_maleBiased_interaction_auto_signif, percent_maleBiased_interaction_auto_signif))
# female_biased genes
nbr_femaleBiased_interaction_auto_signif <- length(de_results_interaction_autosomes$gene_name[which(de_results_interaction_autosomes$logFC > 0 & de_results_interaction_autosomes$qvalue < threshold_qvalue)])
percent_femaleBiased_interaction_auto_signif <- nbr_femaleBiased_interaction_auto_signif/nbr_sexDE_interaction_auto_signif*100
print(sprintf("Number of autosomal up interaction genes = %s (%.3f %% of sex-DE genes)", nbr_femaleBiased_interaction_auto_signif, percent_femaleBiased_interaction_auto_signif))
```

Number of genes DE with pseudotime:

```{r nbr_pseudotimeDE, echo=FALSE}
nbr_sexDE_pseudotime_signif <- sum(de_results_pseudotime_filtered$qvalue < threshold_qvalue & abs(de_results_pseudotime_filtered$logFC) > 0)
percent_sexDE_pseudotime_signif <- nbr_sexDE_pseudotime_signif/length(de_results_pseudotime_filtered$gene_name)*100
print(sprintf("Number of pseudotime-DE genes = %s (representing %.3f %% of the genes)", nbr_sexDE_pseudotime_signif, percent_sexDE_pseudotime_signif))
# up-regulated genes
nbr_up_pseudotime_signif <- length(de_results_pseudotime_filtered$gene_name[which(de_results_pseudotime_filtered$logFC < 0 & de_results_pseudotime_filtered$qvalue < threshold_qvalue)])
percent_up_pseudotime_signif <- nbr_up_pseudotime_signif/nbr_sexDE_pseudotime_signif*100
print(sprintf("Number of up-regulated genes = %s (%.3f %% of pseudotime-DE genes)", nbr_up_pseudotime_signif, percent_up_pseudotime_signif))
# down-regulated genes
nbr_down_pseudotime_signif <- length(de_results_pseudotime_filtered$gene_name[which(de_results_pseudotime_filtered$logFC > 0 & de_results_pseudotime_filtered$qvalue < threshold_qvalue)])
percent_down_pseudotime_signif <- nbr_down_pseudotime_signif/nbr_sexDE_pseudotime_signif*100
print(sprintf("Number of down-regulated genes = %s (%.3f %% of pseudotime-DE genes)", nbr_down_pseudotime_signif, percent_down_pseudotime_signif))
```

```{r nbr_pseudotimeDE_auto, echo=FALSE}
nbr_sexDE_pseudotime_auto_signif <- sum(de_results_pseudotime_autosomes$qvalue < threshold_qvalue & abs(de_results_pseudotime_autosomes$logFC) > 0)
percent_sexDE_pseudotime_auto_signif <- nbr_sexDE_pseudotime_auto_signif/length(de_results_pseudotime_autosomes$gene_name)*100
print(sprintf("Number of autosomal pseudotime-DE genes = %s (representing %.3f %% of the genes)", nbr_sexDE_pseudotime_signif, percent_sexDE_pseudotime_signif))
```



# Plots

## Barplot representing the number of DE genes

Create a barplot of each of the 3 analysis representing the number of DE genes (up and down regulated separatly).

```{r function_barplot, echo=FALSE}
barplot <- function(data, legend = TRUE, nrowLegend = 4, myTitle = "Nbr sex-DE genes",
                    colors = c("Male-other" = "#053061", "Male-sex" = "#99b3ff", 
                               "Female-other" = "#B2182B", "Female-sex" = "#f3a5ae"),
                    labels = c("Female-other" = "Female autosomal genes", "Female-sex" = "Female sex chr. genes",
                               "Male-other" = "Male autosomal genes", "Male-sex" = "Male sex chr. genes")){
  p <- ggplot(data, aes(x = variable, y = nbr_genes, fill = coloring))
  p <- p + geom_bar(stat = "identity")
  p <- p + scale_fill_manual(values = colors, labels = labels)
  p <- p + theme_bw() + labs(x = "", y = "")
  p <- p + ggtitle(myTitle)
  p <- p + theme(legend.position="bottom", legend.title = element_blank()) 
  p <- p + guides(fill = guide_legend(nrow=nrowLegend, byrow=TRUE))
  if (!legend){
    p <- p +  guides(fill = "none")
  }
  return(p)
}
```

### Sex-DE genes

```{r barplot_sexDE, echo=FALSE}
# prepare data
sexChr <- c("chrY", "chrX")
barplot_data_sexDE <- data.frame(variable = c("Female", "Female", "Male", "Male"), 
                           chr = c("other", "sex", "other", "sex"), 
                           nbr_genes = c(length(de_results_autosomes$gene_name[which(de_results_autosomes$qvalue < threshold_qvalue & de_results_autosomes$logFC > 0)]),
                                         length(de_results_filtered$gene_name[which((de_results_filtered$chr %in% sexChr) & de_results_filtered$qvalue < threshold_qvalue & de_results_filtered$logFC > 0)]),
                                         length(de_results_autosomes$gene_name[which(de_results_autosomes$qvalue < threshold_qvalue & de_results_autosomes$logFC < -0)]),
                                         length(de_results_filtered$gene_name[which((de_results_filtered$chr %in% sexChr) & de_results_filtered$qvalue < threshold_qvalue & de_results_filtered$logFC < 0)]))
)

barplot_data_sexDE$coloring <- paste(barplot_data_sexDE$variable, barplot_data_sexDE$chr, sep = "-")
barplot_data_sexDE <- barplot_data_sexDE[order(barplot_data_sexDE$nbr_genes, decreasing = TRUE),] # order data
barplot_data_sexDE$coloring <- factor(barplot_data_sexDE$coloring, levels = c("Male-sex", "Female-sex", "Male-other" , "Female-other"))
barplot_data_sexDE$variable <- factor(barplot_data_sexDE$variable, levels = c("Male", "Female"))

# barplot
p <- barplot(barplot_data_sexDE)
p
# save plot
namePlot <- "barplot_nbr_sexDE_genes.png"
ggsave(plot = p, namePlot, path = "images/4-GTEx/1-GTEx_analysis/barplot/",
       width = 8, height = 17, units = "cm", create.dir = TRUE)
```

### Pseudotime-DE genes

```{r barplot_pseudotimeDE, echo=FALSE}
barplot_data_pseudoDE <- data.frame(variable = c("Up", "Down"),
                           nbr_genes = c(length(de_results_pseudotime_filtered$gene_name[which(de_results_pseudotime_filtered$qvalue < threshold_qvalue & de_results_pseudotime_filtered$logFC > 0)]),
                                         length(de_results_pseudotime_filtered$gene_name[which(de_results_pseudotime_filtered$qvalue < threshold_qvalue & de_results_pseudotime_filtered$logFC < 0)]))
)

barplot_data_pseudoDE$coloring <- barplot_data_pseudoDE$variable
barplot_data_pseudoDE <- barplot_data_pseudoDE[order(barplot_data_pseudoDE$nbr_genes, decreasing = TRUE),] # order data

# barplot
p <- barplot(barplot_data_pseudoDE, myTitle = "Nbr pseudotime-DE genes",
             colors = c("Up" = 'aquamarine4', "Down" = "chocolate"),
             labels = c("Up" = 'Upregulated', "Down" = "Downregulated"), 
             nrowLegend = 2)
p
# save plot
namePlot <- "barplot_nbr_pseudotimeDE_genes.png"
ggsave(plot = p, namePlot, path = "images/4-GTEx/1-GTEx_analysis/barplot/",
       width = 8, height = 17, units = "cm")
```

### Interaction-DE genes

```{r barplot_interactionDE, echo=FALSE}
barplot_data_intDE <- data.frame(variable = c("Up", "Down"),
                           nbr_genes = c(length(de_results_interaction_filtered$gene_name[which(de_results_interaction_filtered$qvalue < threshold_qvalue & de_results_interaction_filtered$logFC > 0)]),
                                         length(de_results_interaction_filtered$gene_name[which(de_results_interaction_filtered$qvalue < threshold_qvalue & de_results_interaction_filtered$logFC < 0)]))
)

barplot_data_intDE$coloring <- barplot_data_intDE$variable
barplot_data_intDE <- barplot_data_intDE[order(barplot_data_intDE$nbr_genes, decreasing = TRUE),] # order data

# barplot
p <- barplot(barplot_data_intDE, myTitle = "Nbr interaction-DE genes",
             colors = c("Up" = 'tomato', "Down" = "lightblue"),
             labels = c("Up" = 'More upregulated in females', "Down" = "More upregulated in males"), 
             nrowLegend = 2)
p
# save plot
namePlot <- "barplot_nbr_interactionDE_genes.png"
ggsave(plot = p, namePlot, path = "images/4-GTEx/1-GTEx_analysis/barplot/",
       width = 8, height = 16, units = "cm")
```


## Volcanoplot

### Sex-DE genes

```{r volcanoplot_sexDE_auto, echo=FALSE}
keyvals <- rep('grey75', nrow(de_results_autosomes))
names(keyvals) <- rep('NS', nrow(de_results_autosomes))

keyvals[which(de_results_autosomes$logFC < 0 & de_results_autosomes$qvalue < threshold_qvalue)] <- 'darkblue'
names(keyvals)[which(de_results_autosomes$logFC  < 0 & de_results_autosomes$qvalue < threshold_qvalue)] <- 'Signif. males'

keyvals[which(de_results_autosomes$logFC > 0 & de_results_autosomes$qvalue < threshold_qvalue)] <- 'darkred'
names(keyvals)[which(de_results_autosomes$logFC > 0 & de_results_autosomes$qvalue < threshold_qvalue)] <- 'Signif. females'

unique(keyvals)
unique(names(keyvals))

p <- EnhancedVolcano(de_results_autosomes,
                     lab = as.character(de_results_autosomes$gene_name),
                     x = 'logFC',
                     y = 'qvalue',
                     xlim = c(-1.5,1.5),
                     FCcutoff = 2,
                     pCutoff = 10E-15,
                     cutoffLineType = 'blank',
                     colCustom = keyvals,
                     legendPosition = "none",
                     ylab = expression('-Log'[10]*' q-value'),
                     subtitle = "sex-DE analysis in HDBR",
                     caption = "autosomal genes")
p

# save plot
namePlot <- 'volcanoplot_colored_HDBR_dream_forebrain_pseudotime_sva_ind_interaction_sexDE_auto.png'
ggsave(namePlot, path = "images/4-GTEx/1-GTEx_analysis/volcanoplot/",
       width = 12, height = 15, units = "cm", create.dir = TRUE)
```

### Pseudotime-DE genes

```{r volcanoplot_pseudotimeDE, echo=FALSE}
keyvals <- rep('grey75', nrow(de_results_pseudotime_filtered))
names(keyvals) <- rep('NS', nrow(de_results_pseudotime_filtered))

keyvals[which(de_results_pseudotime_filtered$logFC < 0 & de_results_pseudotime_filtered$qvalue < threshold_qvalue)] <- 'chocolate'
names(keyvals)[which(de_results_pseudotime_filtered$logFC  < 0 & de_results_pseudotime_filtered$qvalue < threshold_qvalue)] <- 'Signif. Down'

keyvals[which(de_results_pseudotime_filtered$logFC > 0 & de_results_pseudotime_filtered$qvalue < threshold_qvalue)] <- 'aquamarine4'
names(keyvals)[which(de_results_pseudotime_filtered$logFC > 0 & de_results_pseudotime_filtered$qvalue < threshold_qvalue)] <- 'Signif. Up'

unique(keyvals)
unique(names(keyvals))

p <- EnhancedVolcano(de_results_pseudotime_filtered,
                     lab = as.character(de_results_pseudotime_filtered$gene_name),
                     x = 'logFC',
                     y = 'qvalue',
                     FCcutoff = 10,
                     pCutoff = 10E-10,
                     cutoffLineType = 'blank',
                     colCustom = keyvals,
                     legendPosition = "none",
                     ylab = expression('-Log'[10]*' q-value'),
                     subtitle = "pseudotime-DE analysis in HDBR",
                     caption = "all genes")
p

# save plot
namePlot <- 'volcanoplot_colored_HDBR_dream_forebrain_pseudotime_sva_ind_interaction_pseudotimeDE.png'
ggsave(namePlot, path = "images/4-GTEx/1-GTEx_analysis/volcanoplot/",
       width = 12, height = 15, units = "cm")
```

### Interaction-DE genes

```{r volcanoplot_interactionDE_auto, echo=FALSE}
keyvals <- rep('grey75', nrow(de_results_interaction_autosomes))
names(keyvals) <- rep('NS', nrow(de_results_interaction_autosomes))

keyvals[which(de_results_interaction_autosomes$logFC < -0 & de_results_interaction_autosomes$qvalue < threshold_qvalue)] <- 'lightblue'
names(keyvals)[which(de_results_interaction_autosomes$logFC  < -0 & de_results_interaction_autosomes$qvalue < threshold_qvalue)] <- 'Signif. males'

keyvals[which(de_results_interaction_autosomes$logFC > 0 & de_results_interaction_autosomes$qvalue < threshold_qvalue)] <- 'tomato'
names(keyvals)[which(de_results_interaction_autosomes$logFC > 0 & de_results_interaction_autosomes$qvalue < threshold_qvalue)] <- 'Signif. females'

unique(keyvals)
unique(names(keyvals))

p <- EnhancedVolcano(de_results_interaction_autosomes,
                     lab = as.character(de_results_interaction_autosomes$gene_name),
                     x = 'logFC',
                     y = 'qvalue',
                     FCcutoff = 4,
                     pCutoff = 10E-20,
                     cutoffLineType = 'blank',
                     colCustom = keyvals,
                     legendPosition = "none",
                     ylab = expression('-Log'[10]*' q-value'),
                     subtitle = "interaction-DE analysis in HDBR",
                     caption = "autosomal genes")
p

# save plot
namePlot <- 'volcanoplot_colored_HDBR_dream_forebrain_pseudotime_sva_ind_interaction_interactionDE_auto.png'
ggsave(namePlot, path = "images/4-GTEx/1-GTEx_analysis/volcanoplot/",
       width = 12, height = 15, units = "cm")
```


## Compare pseudotime-DE and sex-DE genes

### Upset plot

Merge sexDE and pseudotimeDE data to use it with ComplexUpSet package.

```{r merge_data, echo=FALSE}
merged_data <- merge(de_results_filtered[, c(1:5,11)], de_results_pseudotime_filtered[, c(1:5,11)], #all.x = TRUE,
                     by = c("gene_id", "gene_name", "gene_type", "chr"), suffixes = c("_sexDE", "_pseudotimeDE"))
merged_data$sexDE <- ifelse(merged_data$qvalue_sexDE < threshold_qvalue, TRUE, FALSE)
merged_data$femaleDE <- ifelse(merged_data$qvalue_sexDE < threshold_qvalue & merged_data$logFC_sexDE > 0, TRUE, FALSE)
merged_data$maleDE <- ifelse(merged_data$qvalue_sexDE < threshold_qvalue & merged_data$logFC_sexDE < 0, TRUE, FALSE)
merged_data$pseudotimeDE <- ifelse(merged_data$qvalue_pseudotimeDE < threshold_qvalue, TRUE, FALSE)
merged_data$upDE <- ifelse(merged_data$qvalue_pseudotimeDE < threshold_qvalue & merged_data$logFC_pseudotimeDE > 0, TRUE, FALSE)
merged_data$downDE <- ifelse(merged_data$qvalue_pseudotimeDE < threshold_qvalue & merged_data$logFC_pseudotimeDE < 0, TRUE, FALSE)
```

Also add information about autosome/sex-chromosome genes.

```{r add_chr, echo=FALSE}
merged_data$type <- ifelse(merged_data$chr %in% c("chrX", "chrY"), "sex chr.", "autosomes")
merged_data$sexChr <- ifelse(merged_data$chr %in% c("chrX", "chrY"), "TRUE", "FALSE")
merged_data$type <- factor(merged_data$type, levels = c("sex chr.", "autosomes"))
```

Rename columns to improve upset plot.

```{r rename_col, echo=FALSE}
names(merged_data)[c(10, 11, 13, 14)] <- c("female", "male", "up", "down")
```

Format the data for the upset plot from the ComplexHeatmap package.

```{r format_data_upset, echo=FALSE}
combination_matrix <- make_comb_mat(merged_data[c(10, 11, 13, 14)])
combination_matrix <- combination_matrix[comb_degree(combination_matrix) > 0]
```

Upset plot

```{r upset_plot, echo=FALSE}
colorVal <- list(group = c("up" = "#16967D", "down" = "#D35400", "male" = "#053061", "female" = "#B2182B"))
p <- UpSet(combination_matrix, set_order = c("up", "down", "male", "female"), 
           comb_order = order(comb_size(combination_matrix), decreasing = TRUE),
           row_names_side = "left",
           top_annotation = upset_top_annotation(combination_matrix, 
                                                 show_annotation_name = FALSE,
                                                 add_numbers = TRUE,
                                                 numbers_rot = 0,
                                                 numbers_gp = gpar(fontsize = 8),
                                                 height = unit(7, "cm")),
           left_annotation = rowAnnotation(group = c("female", "male", "down", "up"), 
                                            col = colorVal,
                                            show_annotation_name = FALSE,
                                            show_legend = FALSE)
)
p
# save plot
png("images/4-GTEx/1-GTEx_analysis/UpsetPlot/complexUpsetPlot_between_sexDE_pseudotimeDE_allGenes.png",
    width = 10, height = 10, units = "cm", res = 300)
p
dev.off()
```


### Test significance of overlap between pseudotime-DE and sex-DE genes

#### Overall overlap

Number of sex-DE genes also pseudotime-DE.

```{r nbr_sexDE_pseudoDE, echo=FALSE}
de_results_signif <- de_results_filtered[which(de_results_filtered$qvalue < threshold_qvalue),]
de_results_pseudotime_signif <- de_results_pseudotime_filtered[which(de_results_pseudotime_filtered$qvalue < threshold_qvalue),]
nbr_signif_overlap <- dim(de_results_signif[which(de_results_signif$gene_id %in% de_results_pseudotime_signif$gene_id),])[1]
nbr_total_overlap <- dim(de_results_signif[which(de_results_signif$gene_id %in% de_results_pseudotime_filtered$gene_id),])[1]
print(sprintf("Number of sex-DE genes also pseudotime-DE genes = %s / %s (representing %.3f %% of the genes)", nbr_signif_overlap, nbr_total_overlap, nbr_signif_overlap/nbr_total_overlap*100))
```

Are sex-DE genes enriched in pseudotime-DE genes?
```{r hypergeometric_test_function, echo=FALSE}
hypergeometric_test <- function(listGeneset, universe){
  listGenesetSize <- lapply(listGeneset, length)
  overlapSize <- length(purrr::reduce(listGeneset, intersect))
  universeSize <- length(universe)
  print(c(overlapSize-1, listGenesetSize[[1]], universeSize-listGenesetSize[[1]], listGenesetSize[[2]]))
  pval <- phyper(overlapSize-1, listGenesetSize[[1]], universeSize-listGenesetSize[[1]], listGenesetSize[[2]], lower.tail = FALSE, log.p = FALSE)
  return(pval)
}
```
```{r enrich_sexDE_pseudoDE, echo=FALSE}
# test for over-representation of the sex-DE genes in pseudotime-DE genes
listGeneSet <- list(sexDE = de_results_signif$gene_id, pseudotimeDE = de_results_pseudotime_signif$gene_id)
hypergeometric_test(listGeneSet, de_results_filtered$gene_id)
```

#### Specific overlap between male-biased genes and pseudotime down-regulated genes

```{r nbr_maleBiased_pseudoDownDE, echo=FALSE}
signifGenes_male <- de_results_filtered[which(de_results_filtered$qvalue < threshold_qvalue & de_results_filtered$logFC < 0),]
signifGenes_pseudotime_down <- de_results_pseudotime_filtered[which(de_results_pseudotime_filtered$qvalue < threshold_qvalue & de_results_pseudotime_filtered$logFC < 0),]
nbr_signif_overlap_male_down <- dim(signifGenes_male[which(signifGenes_male$gene_id %in% signifGenes_pseudotime_down$gene_id),])[1]
nbr_total_overlap_male_down <- dim(signifGenes_male[which(signifGenes_male$gene_id %in% de_results_pseudotime_filtered$gene_id),])[1]
print(sprintf("Number of male-biased genes also pseudotime down-regulated genes = %s / %s (representing %.3f %% of the genes)", nbr_signif_overlap_male_down, nbr_total_overlap_male_down, nbr_signif_overlap_male_down/nbr_total_overlap_male_down*100))
```

```{r hypergeometric_test_maleBiased, echo=FALSE}
# test for over-representation of the male-biased genes in pseudotime down-regulated genes
phyper(nbr_signif_overlap_male_down, length(signifGenes_male$gene_id), length(de_results_filtered$gene_id)-length(signifGenes_male$gene_id), length(signifGenes_pseudotime_down$gene_id), 
       lower.tail = FALSE, log.p = FALSE)
```


#### Specific overlap between female-biased genes and pseudotime up-regulated genes

```{r nbr_femaleBiased_pseudoUpDE, echo=FALSE}
signifGenes_female <- de_results_filtered[which(de_results_filtered$qvalue < threshold_qvalue & de_results_filtered$logFC > 0),]
signifGenes_pseudotime_up <- de_results_pseudotime_filtered[which(de_results_pseudotime_filtered$qvalue < threshold_qvalue & de_results_pseudotime_filtered$logFC > 0),]
nbr_signif_overlap_female_up <- dim(signifGenes_female[which(signifGenes_female$gene_id %in% signifGenes_pseudotime_up$gene_id),])[1]
nbr_total_overlap_female_up <- dim(signifGenes_female[which(signifGenes_female$gene_id %in% de_results_pseudotime_filtered$gene_id),])[1]
print(sprintf("Number of female-biased genes also pseudotime up-regulated genes = %s / %s (representing %.3f %% of the genes)", nbr_signif_overlap_female_up, nbr_total_overlap_female_up, nbr_signif_overlap_female_up/nbr_total_overlap_female_up*100))
```

```{r hypergeometric_test_femaleBiased, echo=FALSE}
# test for over-representation of the female-biased genes in pseudotime up-regulated genes
phyper(nbr_signif_overlap_female_up, length(signifGenes_female$gene_id), length(de_results_filtered$gene_id)-length(signifGenes_female$gene_id), length(signifGenes_pseudotime_up$gene_id), 
       lower.tail = FALSE, log.p = FALSE)
```


## Variance partition plot


Load voom normalized counts and phenotypes.

```{r load_data, echo=FALSE}
load("data/4-GTEx/1-GTEx_analysis/brain_forebrain_v8_dream_counts_voom_phenotypes_pseudotime_interaction.RData")
```

Compute variance attributable to each variable

```{r varpart, echo=FALSE}
form <- reformulate(c("SEX", "PSEUDOTIME", "SEX:PSEUDOTIME", "(1 | SUBJID)", paste0("X", 1:sva.obj$n.sv)))
varPart <- fitExtractVarPartModel(y, form, metadata)
# sort variables (i.e. columns) by median fraction of variance explained
vp <- sortCols( varPart )
```

```{r mean_varexplained, echo=FALSE}
colMeans(vp)
```


Violin plot of contribution of each variable to total variance

```{r violin_plot, echo=FALSE}
p <- plotVarPart( vp )
p <- p + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
p
ggsave("violinPlot_variancePartition.png", p,
       path = "images/4-GTEx/1-GTEx_analysis/variancePartition/",
       width = 20, height = 12, units = "cm", create.dir = TRUE)
```


# Session Info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
