---
title: 'Scripts to create the main figures: Figure 4'
author: "Clara Benoit-Pilven"
date: "`r Sys.Date()`"
output: html_document
---

```{r load_library, echo=FALSE}
suppressPackageStartupMessages( library(tidyverse) )
suppressPackageStartupMessages( library(edgeR) )
suppressPackageStartupMessages( library(limma) )
suppressPackageStartupMessages( library(gridExtra) )
suppressPackageStartupMessages( library(ggpubr) )
suppressPackageStartupMessages( library(ggpp) )
suppressPackageStartupMessages( library(cowplot) )
suppressPackageStartupMessages( library(ggh4x) )
```

```{r setup, echo=FALSE}
knitr::opts_knit$set(root.dir = '~/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/')
```


This figure shows the enrichment results with the different methods (exome, GWAS and metabrain network) for schizophreania (SCZ).

# Load and format metabrain data

## Load data

```{r load_metabrain_data, echo=FALSE}
metabrain_enrichRes <- read.table("data/5-postprocessing/6-Network/results_hyper/enrichment_results.txt", header = TRUE, sep = "\t")
```

## Format data

Select only the data of interest:
- bayes results (prenatal-specific and shared) for male- and female-biased genes
- corrected for level of expression
- cortex
- SCZ

```{r select_metabrain_data, echo=FALSE}
metabrain_enrichRes_selected <- metabrain_enrichRes[which(metabrain_enrichRes$analysis %in% c("prenatal_specific", "shared") &
                                                            metabrain_enrichRes$direction != "both" &
                                                            metabrain_enrichRes$ctrlExpr &
                                                            metabrain_enrichRes$geneList == "Schizophrenia.MetaBrainCortex"), ]
```

Format data for the plot.

```{r format_metabrain_data, echo=FALSE}
# add * for signif enrichment or depletion (with a threshold of 0.05/8*8=0.0078)
metabrain_enrichRes_selected$signif <- ifelse(metabrain_enrichRes_selected$relative_enrich > 1, 
                                              ifelse(metabrain_enrichRes_selected$pval_enrich<0.0078 & metabrain_enrichRes_selected$permut_pval < 0.05, "*", ""),
                                              ifelse(metabrain_enrichRes_selected$pval_deple<0.0078 & metabrain_enrichRes_selected$permut_pval < 0.05, "*", ""))
# order the different direction of effect
metabrain_enrichRes_selected$direction <- factor(metabrain_enrichRes_selected$direction,
                                                 levels = c("female", "male"))
# split network names into disease and metaBrain Dataset
metabrain_enrichRes_selected2 <- separate_wider_delim(data = metabrain_enrichRes_selected, cols = "geneList", delim = ".", names = c("disease", "metaBrainData")) 
# rename metaBrainData
metabrain_enrichRes_selected2$metaBrainData <- "MetaBrain\nCortex"
# rename the 2 analyses
metabrain_enrichRes_selected2$analysis <- factor(metabrain_enrichRes_selected2$analysis,
                                                 levels = c("prenatal_specific", "shared"),
                                                 labels = c("Prenatal\nspecific", "Shared"))

# extract and transform the results for the random set of genes
metabrain_enrichRes_random_selected2 <- metabrain_enrichRes_selected2[, c("disease", "metaBrainData", "analysis", "direction", "pval_enrich_random", "pval_deple_random", "relative_enrich_random")] %>% separate_longer_delim(cols = c(pval_enrich_random, pval_deple_random, relative_enrich_random), delim = ";")
metabrain_enrichRes_random_selected2$pval_enrich_random <- as.numeric(metabrain_enrichRes_random_selected2$pval_enrich_random)
metabrain_enrichRes_random_selected2$pval_deple_random <- as.numeric(metabrain_enrichRes_random_selected2$pval_deple_random)
metabrain_enrichRes_random_selected2$relative_enrich_random <- as.numeric(metabrain_enrichRes_random_selected2$relative_enrich_random)
```



# Load and format exome diseases enrichment results

## Load data

```{r load_exome_data, echo=FALSE}
exome_results <- read.table("data/5-postprocessing/4-DiseaseEnrichment/results_disease_hyper/enrichment_results.txt", 
                            header = TRUE, sep = "\t")
```

## Format data

Select only the data of interest:
- bayes results (prenatal-specific and shared) for male- and female-biased genes
- corrected for level of expression
- SCZ

```{r select_exome_data, echo=FALSE}
exome_results_selected <- exome_results[which(exome_results$analysis %in% c("prenatal_specific", "shared") & 
                                                exome_results$direction != "both" & 
                                                exome_results$ctrlExpr &
                                                exome_results$geneList == "scz"), ]
```

Format data for the plot.

```{r format_exome_data, echo=FALSE}
# add * for signif enrichment or depletion (with a threshold of 0.05/9*8=0.00069)
exome_results_selected$signif <- ifelse(exome_results_selected$relative_enrich > 1, 
                                        ifelse(exome_results_selected$pval_enrich<0.00069 & exome_results_selected$permut_pval < 0.05, "*", ""),
                                        ifelse(exome_results_selected$pval_deple<0.00069 & exome_results_selected$permut_pval < 0.05, "*", ""))
# order the different direction of effect
exome_results_selected$direction <- factor(exome_results_selected$direction,
                                           levels = c("female", "male"))
# rename the 2 analyses
exome_results_selected$analysis <- factor(exome_results_selected$analysis,
                                          levels = c("prenatal_specific", "shared"),
                                          labels = c("Prenatal\nspecific", "Shared"))

# extract and transform the results for the random set of genes
exome_results_random_selected <- exome_results_selected[, c("geneList", "analysis", "direction", "pval_enrich_random", "pval_deple_random", "relative_enrich_random")] %>% separate_longer_delim(cols = c(pval_enrich_random, pval_deple_random, relative_enrich_random), delim = ";")
exome_results_random_selected$pval_enrich_random <- as.numeric(exome_results_random_selected$pval_enrich_random)
exome_results_random_selected$pval_deple_random <- as.numeric(exome_results_random_selected$pval_deple_random)
exome_results_random_selected$relative_enrich_random <- as.numeric(exome_results_random_selected$relative_enrich_random)
```


# Load and format MAGMA results

## Load data

```{r load_MAGMA_data, echo=FALSE}
magma_results <- read.table("data/5-postprocessing/5-MAGMA/Magma_allLists_res.txt", header = TRUE)
names(magma_results) <- c("Comparison", "Trait", "Ngenes", "Beta", "Beta_std", "SE", "pval", "Dataset")
```

## Format data

Select only the data of interest:
- bayes results (prenatal-specific and shared) for male- and female-biased genes
- SCZ


```{r select_MAGMA_data, echo=FALSE}
magma_results_selected <- magma_results[which(magma_results$Dataset %in% c("prenatal_specific_femaleBiased", "prenatal_specific_maleBiased", "shared_femaleBiased", "shared_maleBiased") &
                                                magma_results$Trait == "SCZ"),]
```

Format the data for the plot.

```{r format_MAGMA_data, echo=FALSE}
magma_results_selected$Comparison <-magma_results_selected$Dataset
magma_results_selected <- separate(magma_results_selected, Dataset, sep = '([shared|specific]_)', into = c("Analysis", "Direction")) ## Strange behavior: the last character is removed. Why??
magma_results_selected$Direction <- factor(magma_results_selected$Direction, 
                                           levels = c("femaleBiased", "maleBiased"), 
                                           labels = c("femaleBiased" = "female", "maleBiased" = "male"))
magma_results_selected$Analysis <- factor(magma_results_selected$Analysis, levels = c("prenatal_specifi", "share"),
                                          labels = c("Prenatal\nspecific", "Shared"))
# add signif
magma_results_selected$signif <- ifelse(magma_results_selected$pval<0.0069, "*", "")
```


# Plot

## Plot Metabrain results

```{r plot_metaBrain_data, echo=FALSE}
# Colour strips
strip <- strip_themed(background_x = elem_list_rect(fill = c("#9729d6", "#d5952a"), color = c("#9729d6", "#d5952a")))
# plot
p <- ggplot(metabrain_enrichRes_random_selected2, aes(x = direction, y = relative_enrich_random, fill = direction, alpha = 0.4))
p <- p + geom_violin(linewidth = 0.1)
p <- p + geom_hline(yintercept = 1)
p <- p + facet_grid2(cols = vars(analysis), scales = "free_y", drop = TRUE, strip = strip, switch="x")
p <- p + scale_fill_manual(values = c("female" = "#ae4d4d", "male" = "#4d4dae"),
                           name = NULL)
p <- p + geom_point(data = metabrain_enrichRes_selected2, aes(x = direction, y = relative_enrich, color = direction), 
                    position = position_dodge(0.9), size = 2, alpha = 1)
p <- p + geom_text(data = metabrain_enrichRes_selected2, aes(x = direction, y = ifelse(relative_enrich>1, relative_enrich + 0.05, relative_enrich - 0.2), label = signif),
                   position = position_dodge(0.9), size = 5, alpha = 1)
p <- p + scale_color_manual(values = c("female" = "darkred", "male" = "darkblue"),
                            name = NULL)
p <- p + theme_minimal()
p <- p + xlab(NULL) + ylab("Relative enrichment")
p <- p + guides(alpha="none") 
p <- p + ggtitle("SCZ MetaBrain\nnetwork")
p <- p + theme(text = element_text(size=14),
               strip.text.x = element_text(size = 0),
               plot.title = element_text(hjust = 0.5),
               title = element_text(size = 12),
               strip.clip = "off",
               axis.text.x = element_blank())
p_metabrain <- p + coord_cartesian(xlim = c(0,1.75))
p_metabrain
```


## Plot exome results

```{r plot_bayes_ctrlExpr, echo=FALSE}
# Colour strips
strip <- strip_themed(background_x = elem_list_rect(fill = c("#9729d6", "#d5952a"), color = c("#9729d6", "#d5952a")))
# plot
p <- ggplot(exome_results_random_selected, aes(x = direction, y = relative_enrich_random, fill = direction, alpha = 0.4))
p <- p + geom_violin(linewidth = 0.1)
p <- p + geom_hline(yintercept = 1)
p <- p + scale_fill_manual(values = c("female" = "#ae4d4d", "male" = "#4d4dae"))
p <- p + geom_point(data = exome_results_selected, aes(x = direction, y = relative_enrich, color = direction), 
                    position = position_dodge(0.9), size = 2, alpha = 1)
p <- p + geom_text(data = exome_results_selected, aes(x = direction, y = ifelse(relative_enrich>1, relative_enrich + 0.05, relative_enrich - 0.2), label = signif),
                   position = position_dodge(0.9), size = 5, alpha = 1)
p <- p + scale_color_manual(values = c("female" = "darkred", "male" = "darkblue"))
p <- p + facet_grid2(cols = vars(analysis), scales = "free", space = "free", strip = strip, switch="x")
p <- p + theme_minimal()
p <- p + xlab(NULL) + ylab("Relative enrichent")
p <- p + guides(alpha="none", fill="none", color="none") 
p <- p + ggtitle("SCZ exome-seq\nstudy")
p <- p + theme(text = element_text(size=14),
               strip.text.x = element_text(size = 0),
               plot.title = element_text(hjust = 0.5),
               title = element_text(size = 12),
               strip.clip = "off",
               axis.text.x = element_blank())
p_exome <- p + coord_cartesian(xlim = c(0,3))
p_exome
```

## Plot MAGMA results 

```{r plot_MAGMA_results, echo=FALSE}
# Colour strips
strip <- strip_themed(background_x = elem_list_rect(fill = c("#9729d6", "#d5952a"), color = c("#9729d6", "#d5952a")))
# plot
p <- ggplot(magma_results_selected, aes(y = Beta, x = Direction, ymin = Beta-SE, ymax = Beta+SE, 
                                   color = Direction))
p <- p + geom_hline(yintercept = 0, color = "darkgrey")
p <- p + geom_pointrange(position = position_dodge(width = 0.75))
p <- p + scale_color_manual(values = c("male" = 'darkblue', "female" = 'darkred'), 
                            name = "Direction of effect")
p <- p + facet_grid2(cols = vars(Analysis), scales = "free_y", space = "free_y", drop = TRUE, strip = strip, switch="x")
# add significance to plot
p <- p + geom_text(aes(y = Beta+SE+0.05, x = Direction, label = signif), size = 5, position = position_dodge(width = 0.2))
p <- p + theme_minimal() + xlab(NULL)
p <- p + guides(color="none") 
p <- p + ggtitle("SCZ GWAS\n(MAGMA)")
p <- p + theme(text = element_text(size=14),
               strip.text.x = element_text(size = 0),
               plot.title = element_text(hjust = 0.5),
               title = element_text(size = 12),
               strip.clip = "off",
               axis.text.x = element_blank())
p_magma <- p + coord_cartesian(ylim = c(-0.1, 0.2))
p_magma
```


## Arrange the 3 plots together

```{r figure4_mergedplot, echo=FALSE}
ggsave("4_enrichment_analyses.png", 
       path = "images/figures/",
       arrangeGrob(p_exome, p_magma, p_metabrain, ncol = 3, widths=c(2,2,3.3)),
       width = 18, height = 8, units = "cm")
ggsave("4_enrichment_analyses.svg", 
       path = "images/figures/",
       arrangeGrob(p_exome, p_magma, p_metabrain, ncol = 3, widths=c(2,2,3.3)),
       width = 18, height = 8, units = "cm")
```



