---
title: "Differential analysis in prenatal forebrain using developmental stages as covariates"
author: "Clara Benoit-Pilven"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
---

```{r load_libraries, echo=FALSE}
suppressPackageStartupMessages( library(edgeR) )
suppressPackageStartupMessages( library(limma) )
suppressPackageStartupMessages( library(qvalue) )
suppressPackageStartupMessages( library(sva) )
suppressPackageStartupMessages( library(doParallel) )
suppressPackageStartupMessages( library(variancePartition) )
suppressPackageStartupMessages( library(ggplot2) )
```

```{r parallel_processing_param, echo=FALSE}
# Specify parallel processing parameters
# this is used implicitly by dream() to run in parallel
param <- SnowParam(4, "SOCK", progressbar=TRUE)
register(param)
```

```{r setup, echo=FALSE}
knitr::opts_knit$set(root.dir = '~/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/')
```


# Prepare data

## Load the counts and the phenotype data

```{r load_data, echo=FALSE}
load("data/1-preprocessing/counts_forebrain_phenotypes.RData")
```

## Prepare metadata

```{r prepare_metadata, echo=FALSE}
Sex <- as.factor(phenotypes$assigned_sex)
DevStage <- phenotypes$newDevStage
Individual <- as.factor(phenotypes$ind)
# metadata
metadata <- data.frame(Sex = Sex, DevStage = DevStage, Individual = Individual)
rownames(metadata) <- phenotypes$experiment_name
# design for voomDream (the variable to be tested must be a fixed effect)
form <- ~ Sex + (1|DevStage) + (1|Individual)
# design for sva computation
design <- model.matrix(~Sex+DevStage)
```

## Prepare counts for DE analysis

Estimate weights using linear mixed model of dream

```{r voom_norm, echo=FALSE}
png('images/3-DE/analysis_devStage/trendplot/trendplot_VoomDream_forebrain_dev_region_ind.png')
vobjDream <- voomWithDreamWeights(x, form, metadata, plot = TRUE)
dev.off()
```

Estimate surrogates variables (svas) and add them to the model.

```{r estimate_sva, echo=FALSE}
n.sv <- num.sv(vobjDream$E, design, method = "be")
sva.obj <- sva(vobjDream$E, design, method = "irw", n.sv = n.sv)
# add svas to metadata
metadata <- cbind(metadata, sva.obj$sv)
names(metadata) <- c("Sex", "DevStage", "Individual", paste0("X", 1:n.sv))
# include svas in formula
form <- reformulate(c("Sex", "(1|DevStage)", "(1|Individual)", paste0("X", 1:n.sv)))
```

Redo voom normalization with new design matrix (containing svas).

```{r voom_norm_svas, echo=FALSE}
y <- voomWithDreamWeights(x, form, metadata, plot = TRUE)
```

Save the normalized counts and the phenotypes vectors in a RData file.

```{r save_normCounts, echo=FALSE}
save(y, phenotypes, form, sva.obj, metadata,
     file = "data/3-DE/analysis_devStage/normCounts_dream_dev_sva_ind_interaction_phenotypes.RData")
```

## Sex-Differential Expression (sex-DE) analysis

Fit the dream model on each gene.

```{r sexDE_analysis, echo=FALSE}
# By default, uses the Satterthwaite approximation for the hypothesis test
fitmm <- dream(y, form, metadata)
fitmm2 <- topTable(fitmm, coef = 2, number = 30000, sort.by = "none") 
fitmm2$qvalue <- qvalue(fitmm2$P.Value)$qvalues
```

Compute the mean expression of each genes for male and female.

```{r compute_mean_expr, echo=FALSE}
fitmm2$Male <- rowMeans(y$E[which(y$genes$gene_id %in% fitmm2$gene_id), which(Sex==1)]) # for male samples
fitmm2$VarMale <- rowVars(y$E[which(y$genes$gene_id %in% fitmm2$gene_id), which(Sex==1)])
fitmm2$Female <- rowMeans(y$E[which(y$genes$gene_id %in% fitmm2$gene_id), which(Sex==2)]) # for female samples
fitmm2$VarFemale <- rowVars(y$E[which(y$genes$gene_id %in% fitmm2$gene_id), which(Sex==2)])
```

Save results.

```{r write_output, echo=FALSE}
write.table(fitmm2, 
            "data/3-DE/analysis_devStage/VoomDream_topTable_forebrain_dev_sva_ind.txt", 
            col.names = TRUE, row.names = FALSE, quote = FALSE)
```


## Variance partition plot

```{r compute_variance_explained, echo=FALSE}
# compute the variance explained by each variable in the model
form <- reformulate(c("(1|Sex)", "(1|DevStage)", "(1|Individual)", paste0("X", 1:n.sv)))
vp <- fitExtractVarPartModel(y, form, metadata)
```


```{r plot_variance_explained, echo=FALSE}
# variance partition violin plot
p <- plotVarPart(sortCols(vp))
p
ggsave("violinPlot_forebrain_VoomDream_dev_sva_ind.png", plot = p, 
       path = "images/3-DE/analysis_devStage/variancePartition/",
       width = 25, height = 12, units = "cm", create.dir = TRUE)
```



# Session info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
