---
title: "Comparison of DE analysis with pseudotime or developmental stages"
author: "Clara Benoit-Pilven"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
---

```{r load_libraries, echo=FALSE}
suppressPackageStartupMessages( library(edgeR) )
suppressPackageStartupMessages( library(limma) )
suppressPackageStartupMessages( library(tidyverse) )
suppressPackageStartupMessages( library(EnhancedVolcano) )
suppressPackageStartupMessages( library(ComplexHeatmap) )
suppressPackageStartupMessages( library(purrr) )
suppressPackageStartupMessages( library(qvalue) )
```

```{r setup, echo=FALSE}
knitr::opts_knit$set(root.dir = '~/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/')
workDir <- '~/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/'
```

```{r parameters, echo=FALSE}
qvalue_threshold <- 0.01
logFC_threshold <- 0
```



# Load data and functions

## Load sex-DE results

### With pseudotime

```{r load_pseudotime, echo=FALSE}
results_pseudo <- read.table("data/3-DE/VoomDream_topTable_forebrain_pseudotime_sva_ind_interaction.txt",
                           header = TRUE)
results_pseudo$ensemblID <- sapply(strsplit(as.character(results_pseudo$gene_id), "\\."), function(x){x[1]})
```

Prepare the data:
- remove the genes used to define the sex of the samples.
```{r prepare_pseudotime, echo=FALSE}
genes2remove <- c("XIST", as.character(results_pseudo$gene_name[which(results_pseudo$chr == "chrY" & results_pseudo$gene_type == "protein_coding")]))
results_pseudo <- results_pseudo[which(!results_pseudo$gene_name %in% genes2remove),]
```

Get sexDE results for autosomal genes only.
```{r prepare_pseudo_auto, echo=FALSE}
sexChr <- c("chrY", "chrX")
results_pseudo_auto <- results_pseudo[which(!results_pseudo$chr %in% sexChr),]
```

### With developmental stages

```{r load_devStage, echo=FALSE}
results_devStage <- read.table("data/3-DE/analysis_devStage/VoomDream_topTable_forebrain_dev_sva_ind.txt",
                               header = TRUE)
results_devStage$ensemblID <- sapply(strsplit(as.character(results_devStage$gene_id), "\\."), function(x){x[1]})
```

Prepare the data by removing the genes used to define the sex of the samples.
```{r prepare_devStage, echo=FALSE}
genes2remove <- c("XIST", as.character(results_devStage$gene_name[which(results_devStage$chr == "chrY" & results_devStage$gene_type == "protein_coding")]))
results_devStage <- results_devStage[which(!results_devStage$gene_name %in% genes2remove),]
```

Get sexDE results for autosomal genes only.
```{r prepare_devStage_auto, echo=FALSE}
sexChr <- c("chrY", "chrX")
results_devStage_auto <- results_devStage[which(!results_devStage$chr %in% sexChr),]
```

# Statistics of sex-DE results with developmental stages as covariates

Number of sex-DE significant genes:

```{r nbr_sexDE, echo=FALSE}
nbr_sexDE_signif <- sum(results_devStage$qvalue < qvalue_threshold)
percent_sexDE_signif <- nbr_sexDE_signif/length(results_devStage$gene_name)*100
print(sprintf("Number of sex-DE genes = %s (representing %.3f %% of the genes)", nbr_sexDE_signif, percent_sexDE_signif))
# male-biased genes
nbr_maleBiased_signif <- length(results_devStage$gene_name[which(results_devStage$logFC < 0 & results_devStage$qvalue < qvalue_threshold)])
percent_maleBiased_signif <- nbr_maleBiased_signif/nbr_sexDE_signif*100
print(sprintf("Number of male-biased genes = %s (%.3f %% of sex-DE genes)", nbr_maleBiased_signif, percent_maleBiased_signif))
# female_biased genes
nbr_femaleBiased_signif <- length(results_devStage$gene_name[which(results_devStage$logFC > 0 & results_devStage$qvalue < qvalue_threshold)])
percent_femaleBiased_signif <- nbr_femaleBiased_signif/nbr_sexDE_signif*100
print(sprintf("Number of female-biased genes = %s (%.3f %% of sex-DE genes)", nbr_femaleBiased_signif, percent_femaleBiased_signif))
```

# Visualization of sex-DE results with developmental stages as covariates

## Barplot

Function for the barplot.

```{r function_barplot, echo=FALSE}
barplot <- function(data, legend = TRUE, nrowLegend = 4, myTitle = "Nbr sex-DE genes",
                    colors = c("Male-other" = "#053061", "Male-sex" = "#99b3ff", 
                               "Female-other" = "#B2182B", "Female-sex" = "#f3a5ae"),
                    labels = c("Female-other" = "Female autosomal genes", "Female-sex" = "Female sex chr. genes",
                               "Male-other" = "Male autosomal genes", "Male-sex" = "Male sex chr. genes")){
  p <- ggplot(data, aes(x = variable, y = nbr_genes, fill = coloring))
  p <- p + geom_bar(stat = "identity")
  p <- p + scale_fill_manual(values = colors, labels = labels)
  p <- p + theme_bw() + labs(x = "", y = "")
  p <- p + ggtitle(myTitle)
  p <- p + theme(legend.position="bottom", legend.title = element_blank()) 
  p <- p + guides(fill = guide_legend(nrow=nrowLegend, byrow=TRUE))
  if (!legend){
    p <- p +  guides(fill = "none")
  }
  return(p)
}
```

Prepare data for barplot.

```{r prepare_data_barplot, echo=FALSE}
barplot_data_sexDE <- data.frame(variable = c("Female", "Female", "Male", "Male"), 
                           chr = c("other", "sex", "other", "sex"), 
                           nbr_genes = c(length(results_devStage_auto$gene_name[which(results_devStage_auto$qvalue < qvalue_threshold &
                                                                                        results_devStage_auto$logFC > 0)]),
                                         length(results_devStage$gene_name[which((results_devStage$chr %in% sexChr) &
                                                                                   results_devStage$qvalue < qvalue_threshold &
                                                                                   results_devStage$logFC > 0)]),
                                         length(results_devStage_auto$gene_name[which(results_devStage_auto$qvalue < qvalue_threshold &
                                                                                        results_devStage_auto$logFC < 0)]),
                                         length(results_devStage$gene_name[which((results_devStage$chr %in% sexChr) &
                                                                                   results_devStage$qvalue < qvalue_threshold &
                                                                                   results_devStage$logFC < 0)]))
)

barplot_data_sexDE$coloring <- paste(barplot_data_sexDE$variable, barplot_data_sexDE$chr, sep = "-")
barplot_data_sexDE <- barplot_data_sexDE[order(barplot_data_sexDE$nbr_genes, decreasing = TRUE),] # order data
barplot_data_sexDE$coloring <- factor(barplot_data_sexDE$coloring, levels = c("Male-sex", "Female-sex", "Male-other" , "Female-other"))
barplot_data_sexDE$variable <- factor(barplot_data_sexDE$variable, levels = c("Male", "Female"))
```

Barplot

```{r barplot, echo=FALSE}
p <- barplot(barplot_data_sexDE, legend = FALSE, myTitle = "")
p
# save plot
ggsave(plot = p, "barplot_nbr_sexDE_genes_noLegend.png", 
       path = "images/3-DE/analysis_devStage/barplot/", 
       width = 4, height = 6, units = "cm", create.dir = TRUE)
```



## Volcanoplot

```{r volcanoplot_sexDE, echo=FALSE}
keyvals <- rep('grey75', nrow(results_devStage))
names(keyvals) <- rep('NS', nrow(results_devStage))

keyvals[which(results_devStage$logFC < 0 & results_devStage$qvalue < qvalue_threshold)] <- 'darkblue'
names(keyvals)[which(results_devStage$logFC < 0 & results_devStage$qvalue < qvalue_threshold)] <- 'Signif. males'

keyvals[which(results_devStage$logFC > 0 & results_devStage$qvalue < qvalue_threshold)] <- 'darkred'
names(keyvals)[which(results_devStage$logFC > 0 & results_devStage$qvalue < qvalue_threshold)] <- 'Signif. females'

unique(keyvals)
unique(names(keyvals))

p <- EnhancedVolcano(results_devStage,
                     lab = as.character(results_devStage$gene_name),
                     x = 'logFC',
                     y = 'qvalue',
                     xlim = c(-2.5, 2.5),
                     ylim = c(0,20),
                     FCcutoff = 2,
                     pCutoff = 10E-15,
                     cutoffLineType = 'blank',
                     colCustom = keyvals,
                     legendPosition = "none",
                     ylab = expression('-Log'[10]*' q-value'),
                     subtitle = "sex-DE analysis in HDBR - developmental stages as covariates",
                     caption = "all genes")
p

# save plot
ggsave('volcanoplot_colored_sexDE.png', 
       path = "images/3-DE/analysis_devStage/volcanoplot/", 
       width = 12, height = 15, units = "cm", create.dir = TRUE)
```


# Compare the two analyses

## Overlap & upset plot

Merge of sex-DE results to use it with ComplexUpSet package.

```{r merge_data, echo=FALSE}
merged_data <- merge(results_pseudo, results_devStage, 
                     by = c("gene_id", "gene_name", "gene_type", "chr", "ensemblID"), suffixes = c("_pseudo", "_devStages"))

merged_data$femaleDE_pseudo <- ifelse(merged_data$qvalue_pseudo < qvalue_threshold & merged_data$logFC_pseudo > logFC_threshold, TRUE, FALSE)
merged_data$maleDE_pseudo <- ifelse(merged_data$qvalue_pseudo < qvalue_threshold & merged_data$logFC_pseudo < logFC_threshold, TRUE, FALSE)
merged_data$femaleDE_devStages <- ifelse(merged_data$qvalue_devStages < qvalue_threshold & merged_data$logFC_devStages > logFC_threshold, TRUE, FALSE)
merged_data$maleDE_devStages <- ifelse(merged_data$qvalue_devStages < qvalue_threshold & merged_data$logFC_devStages < logFC_threshold, TRUE, FALSE)
```

Rename columns to improve upset plot.

```{r rename_col, echo=FALSE}
names(merged_data)[c(28, 29, 30, 31)] <- c("female\n pseudotime analysis", "male\n pseudotime analysis", "female\n devStages analysis", "male\n devStages analysis")
```

Format the data for the upset plot from the ComplexHeatmap package.

```{r format_data_upset, echo=FALSE}
combination_matrix <- make_comb_mat(merged_data[c(28, 29, 30, 31)])
combination_matrix <- combination_matrix[comb_degree(combination_matrix) > 0]
```


```{r upset_plot, echo=FALSE}
colorVal <- list(group = c("female" = "#B2182B", "male" = "#053061"))
p <- UpSet(combination_matrix, set_order = c("female\n pseudotime analysis", "male\n pseudotime analysis", "female\n devStages analysis", "male\n devStages analysis"), 
           comb_order = order(comb_size(combination_matrix), decreasing = TRUE),
           row_names_gp = gpar(fontsize = 8),
           row_names_side = "left",
           top_annotation = upset_top_annotation(combination_matrix, 
                                                 show_annotation_name = FALSE,
                                                 add_numbers = TRUE,
                                                 numbers_rot = 0,
                                                 numbers_gp = gpar(fontsize = 8),
                                                 height = unit(4, "cm")),
           left_annotation = rowAnnotation(group = c("female", "male", "female", "male"), 
                                            col = colorVal,
                                            show_annotation_name = FALSE,
                                            show_legend = FALSE)
)
p
# save plot
# png("~/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/images/3-DE/UpsetPlot/complexUpsetPlot_between_sexDE_pseudotimeDE_allGenes.png", width = 10, height = 10, units = "cm", res = 300)
png(paste0(workDir, "images/3-DE/analysis_devStage/UpsetPlot/complexUpsetPlot_sexDE_between_pseudotime_devStages_analysis.png"),
    width = 8, height = 8, units = "cm", res = 300)
p
dev.off()
```


## Test overlap

```{r permutation_test_function, echo=FALSE}
permutation_test <- function(listGeneset, universe, nperm = 1000, binwidth = 10){
  observed <- length(purrr::reduce(listGeneset, intersect))
  # permutation
  simulated <- map_dbl(seq_len(nperm), function(x) {
    sim <- map(lengths(listGeneset), ~sample(universe, .x))
    sim <- length(purrr::reduce(sim, intersect))
    return(sim)
  })
  # histogram of simulated values
  p <- ggplot(as.data.frame(simulated), aes(simulated))
  p <- p + geom_histogram(binwidth = binwidth)
  p <- p + geom_vline(aes(xintercept = observed), color = "red")
  p <- p + theme_minimal()
  print(p)
  # pvalue
  pval <- (sum(simulated >= observed) + 1) / (nperm + 1)
  return(pval)
}
```

```{r permutation_test, echo=FALSE}
signifGenes_pseudo <- results_pseudo[which(results_pseudo$qvalue < qvalue_threshold & 
                                             abs(results_pseudo$logFC) > logFC_threshold), ]
signifGenes_devStage <- results_devStage[which(results_devStage$qvalue < qvalue_threshold & 
                                                 abs(results_devStage$logFC) > logFC_threshold), ]
listData <- list(pseudotime = signifGenes_pseudo$gene_name, devStage = signifGenes_devStage$gene_name)

permutation_test(listData, merged_data$gene_name, binwidth = 1)
```

## Overlap with different q-value threshold

For the following q-value threshold (q>=0.1, q<0.1, q<0.05, q<0.01) applied to the analysis using the developmental stages as covariate, we look at how many genes are sex-DE or not in the analysis using the pseudotime as covariate.

```{r table_overlap_qvalue, echo=FALSE}
merged_data$sexDE_pseudotime <- ifelse(merged_data$qvalue_pseudo < qvalue_threshold, "Yes", "No")
merged_data$qvalue_cat_devStage <- ifelse(merged_data$qvalue_devStages < 0.01, "q<0.01",
                                            ifelse(merged_data$qvalue_devStages < 0.05, "q<0.05",
                                                   ifelse(merged_data$qvalue_devStages < 0.1, "q<0.1", "q>0.1")))
table(merged_data$sexDE_pseudotime, merged_data$qvalue_cat_devStage)
```

Do the same but looking at the p-value.

```{r table_overlap_pvalue, echo=FALSE}
merged_data$sexDE_pseudotime <- ifelse(merged_data$qvalue_pseudo < qvalue_threshold, "Yes", "No")
merged_data$pvalue_cat_devStage <- ifelse(merged_data$P.Value_devStages < 0.01, "p<0.01",
                                            ifelse(merged_data$P.Value_devStages < 0.05, "p<0.05",
                                                   ifelse(merged_data$P.Value_devStages < 0.1, "p<0.1", "p>0.1")))
table(merged_data$sexDE_pseudotime, merged_data$pvalue_cat_devStage)
```



## Replication rate

We compute the replication rate (pi1).

```{r function_pi1_score, echo=FALSE}
pi1_score <- function(data1, data2, columnMerge = "ensemblID"){
  # compute pi1 for second dataset compared to the first dataset
  selected_pvalue <- data2$P.Value[which(data2[, columnMerge] %in% data1[, columnMerge])]
  if (berryFunctions::is.error(qvalue::pi0est(p = selected_pvalue, lambda = 0.5, pi0.method = "smoother")) ) {
    pi1 <- qvalue::qvalue(p = selected_pvalue, pi0 = 1)$pi0
  }else{
    pi1 <- 1-qvalue::pi0est(p = selected_pvalue, lambda = 0.5, pi0.method = "smoother")$pi0
  }
  
  # continue only if the pi1 value was computed
  if (is.numeric(pi1)){
    return(pi1)
  } else{
    return("No pi1 analysis!")
  }
}
```

```{r function_pi1_normalized, echo=FALSE}
pi1_normalized <- function(data1, data2, columnMerge = "ensemblID", qvalue_threshold = 0.01, logFC_threshold = 0, verbose = TRUE){
  # get the significant genes in the first dataset
  data1_signif <- data1[which(data1$qvalue < qvalue_threshold & abs(data1$logFC) > logFC_threshold), ]
  
  # pi1 replication rate for the significant genes
  pi1_signif <- pi1_score(data1_signif, data2, columnMerge = "ensemblID")
  if (verbose){
    print(sprintf("pi1 for signif genes = %s", pi1_signif))
  }
  # pi1 replication rate for all genes
  pi1_all <- pi1_score(data1, data2, columnMerge = "ensemblID")
  if (verbose){
    print(sprintf("pi1 for all genes = %s", pi1_all))
  }
  
  # return the pi1 for significant genes normalized by the pi1 for all genes
  return(pi1_signif/pi1_all)
}
```

```{r pi1_score, echo=FALSE}
print("Replication of prenatal pseudotime sex-DE results in prenatal dev. stages sex-DE analysis:")
pi1_normalized(results_pseudo, results_devStage, qvalue_threshold = qvalue_threshold)
print("Replication of prenatal dev. stages sex-DE results in prenatal pseudotime sex-DE analysis:")
pi1_normalized(results_devStage, results_pseudo, qvalue_threshold = qvalue_threshold)
```


## Correlation of the logFC

Scatter plot of the logFC between the 2 analyses with information about the correlation (r and p-value of cor.test).

```{r function_scatter_plot_logFC, echo=FALSE}
scatter_logFC <- function(data1, data2, nameData1, nameData2, pointColor, columnMerge = "ensemblID"){
  # merge the 2 dataset
  de_results_merged <- merge(data1, data2, by = columnMerge, suffixes = c(".data1", ".data2"))
  # test correlation
  res <- cor.test(de_results_merged$logFC.data1, de_results_merged$logFC.data2)
  if (res$p.value == 0){
    annotation_text <- sprintf("r = %.3f\nP < 2.2e-16", res$estimate)
  } else{
    annotation_text <- sprintf("r = %.3f\nP = %1.2e", res$estimate, res$p.value)
  }
  yrng <- range(de_results_merged$logFC.data1)
  xrng <- range(de_results_merged$logFC.data2)
  # plot
  p <- ggplot(de_results_merged, aes(x = logFC.data1, y = logFC.data2))
  p <- p + geom_abline(intercept = 0, slope = 1, col = "black")
  p <- p + geom_hline(yintercept = 0, color = "grey")
  p <- p + geom_vline(xintercept = 0, color = "grey")
  p <- p + geom_point(color=pointColor, alpha = 0.3)
  p <- p + theme_bw() + labs(title = "")
  p <- p + theme(panel.grid = element_blank(), 
                 plot.title = element_text(hjust = 0.5),
                 plot.subtitle = element_text(hjust = 0.5))
  p <- p + xlab(paste0("logFC (", nameData1, ")")) + ylab(paste0("logFC (", nameData2, ")"))
  p <- p + annotate("text", xrng[1], yrng[2], hjust = 0, vjust = 1, 
                    label = annotation_text)
  return(p)
}
```


Correlation plot only of pseudotime sex-DE genes.

```{r corr_pseudo_sexDE, echo=FALSE}
p <- scatter_logFC(results_pseudo[which(results_pseudo$qvalue < qvalue_threshold & abs(results_pseudo$logFC) > logFC_threshold),], results_devStage, "pseudotime", "dev. stages", 
                      pointColor = "black")
p
# save plot
ggsave("correlation_logFC_sigPseudotime.png", plot = p, 
       path = "images/3-DE/analysis_devStage/comparison/", 
       width = 10, height = 11, units = "cm", create.dir = TRUE)
```

Correlation plot only of autosomal pseudotime sex-DE genes.

```{r corr_pseudo_sexDE_auto, echo=FALSE}
p <- scatter_logFC(results_pseudo_auto[which(results_pseudo_auto$qvalue < qvalue_threshold & abs(results_pseudo_auto$logFC) > logFC_threshold),], results_devStage, "pseudotime", "dev. stages", 
                      pointColor = "black")
p
# save plot
ggsave("correlation_logFC_sigPseudotime_auto.png", plot = p, 
       path = "images/3-DE/analysis_devStage/comparison/", 
       width = 10, height = 11, units = "cm", create.dir = TRUE)
```

## Effect size consistency

Test the consistency of the effect size.

```{r _function_consistency, echo=FALSE}
consistency_test <- function(data1, data2, columnMerge = "ensemblID", binomTestP = 0.5,
                             qvalueThreshold = 0.01, logFCThreshold = 0, typeComparison = "1vs2"){
  # merge the 2 datasets
  de_results_merged <- merge(data1, data2, by = columnMerge, suffixes = c(".data1", ".data2"))
  # Depending on the type of comparison choosen, get the observed consistency 
  de_results_merged_filtered <- data.frame()
  nbr_consistent <- 0
  consistency <- 0
  if (typeComparison == "1vs2"){
    # keep only the genes/transcripts that are significant in the first data
    de_results_merged_filtered <- de_results_merged[which(de_results_merged$qvalue.data1 < qvalueThreshold & abs(de_results_merged$logFC.data1) > logFCThreshold), ]
    print(sprintf("Number of genes compared = %s", dim(de_results_merged_filtered)[1]))
    # number of consistent genes/transcripts
    nbr_consistent <- sum(sign(de_results_merged_filtered$t.data1) == sign(de_results_merged_filtered$t.data2))
    consistency <- nbr_consistent/length(de_results_merged_filtered[, columnMerge])
  } else if (typeComparison == "intersection"){
    # keep only the genes/transcripts that are significant in the second data
    de_results_merged_filtered <- de_results_merged[which(de_results_merged$qvalue.data1 < qvalueThreshold & abs(de_results_merged$logFC.data1) > logFCThreshold &
                                                            de_results_merged$qvalue.data2 < qvalueThreshold & abs(de_results_merged$logFC.data2) > logFCThreshold), ]
    print(sprintf("Number of genes compared = %s", dim(de_results_merged_filtered)[1]))
    # number of consistent genes/transcripts
    nbr_consistent <- sum(sign(de_results_merged_filtered$t.data1) == sign(de_results_merged_filtered$t.data2))
    consistency <- nbr_consistent/length(de_results_merged_filtered[, columnMerge])
  } else if (typeComparison == "union"){
    # keep only the genes/transcripts that are significant in the second data
    de_results_merged_filtered <- de_results_merged[which((de_results_merged$qvalue.data1 < qvalueThreshold & abs(de_results_merged$logFC.data1) > logFCThreshold) |
                                                            (de_results_merged$qvalue.data2 < qvalueThreshold & abs(de_results_merged$logFC.data2) > logFCThreshold)), ]
    print(sprintf("Number of genes compared = %s", dim(de_results_merged_filtered)[1]))
    # number of consistent genes/transcripts
    nbr_consistent <- sum(sign(de_results_merged_filtered$t.data1) == sign(de_results_merged_filtered$t.data2))
    consistency <- nbr_consistent/length(de_results_merged_filtered[, columnMerge])
  }
  # compare this consistency score with the consistency of list of random genes
  cons_values <- rep(NA, 1000)
  # to do 1000 times
  for (i in 1:1000){
    # randomly select the same number of genes/transcripts in the data2 to test the overlap with the data1 signif genes
    selection <- sample(de_results_merged[, columnMerge], length(de_results_merged_filtered[, columnMerge]))
    nbr_consistent_th <- sum(sign(de_results_merged$logFC.data2[which(de_results_merged[, columnMerge] %in% selection)]) == sign(de_results_merged$logFC.data1[which(de_results_merged[, columnMerge] %in% selection)]))
    cons_values[i] <- nbr_consistent_th/length(de_results_merged_filtered[, columnMerge])
  }
  cons_values <- as.data.frame(cons_values)
  # plot the distribution of the predited consistency and the real value
  p <- ggplot(cons_values, aes(cons_values))
  p <- p + geom_histogram(binwidth=0.02, boundary = 0, fill="grey", color="black")
  p <- p + theme_bw() + xlab("consistency") + ylab("Frequency")
  p <- p + scale_x_continuous(limits = c(0, 1))
  p <- p + geom_vline(xintercept = consistency, color = "red")
  print(p)
  # test this consistency with a permutation test
  nsmaller <- sum(cons_values >= consistency)
  permutation_test <- round((nsmaller+1)/(1000+1), 3)
  # test this consistency with a binomial test
  result_test <- binom.test(nbr_consistent, length(de_results_merged_filtered[, columnMerge]), p = binomTestP)
  # print results
  print(sprintf("number of consistent gene = %s", nbr_consistent))
  print(sprintf("consistency = %s", consistency))
  print(sprintf("permutation test p-value = %s", permutation_test))
  print(sprintf("binomial test p-value = %s", result_test$p.value))
  return(consistency)
}
```


```{r consistency_score_pseudotimeVSdevStage, echo=FALSE}
print("Consistency of the prenatal pseudotime sex-DE results in prenatal dev. stages sex-DE analysis:")
consistency_all <- consistency_test(results_pseudo, results_devStage, qvalueThreshold = qvalue_threshold, logFCThreshold = 0)
```

## Test if pseudotime introduce bias in sex-DE genes

We hypothize that if the pseudotime introduce a bias in the sex-DE genes we find, we would see less directionality concordance between the pseudotime analysis and the developmental stage analysis for the DE genes classified as prenatal-specific compared to the shared ones.

To test this, we need to load the Bayesian results. Then we compute the logFC consistency for the 2 group of genes (prenatal-specific and shared) separately.

### Load bayes results

```{r load_bayes_results, echo=FALSE}
bayesResults <- read.table("data/4-GTEx/2-comparison/lineModel/lineModel_prenatalVSadult_6models.txt", header = TRUE)
# get prenatal-specific gene list
prenatalSpe_genes <- bayesResults$ensemblID[which(bayesResults$specificity == "PRENATAL")]
# get shared gene list
shared_genes <- bayesResults$ensemblID[which(bayesResults$specificity %in% c("SHARED_UNSPECIFIED", "SHARED0.5", "SHARED1", "SHARED2"))]
```

### Effect size consistency

Define the background as the genes that were not analyzed in the the bayesian analysis (not sex-DE in either data).

```{r define_bckground, echo=FALSE}
background_data <- merged_data[which(!merged_data$ensemblID %in% bayesResults$ensemblID),]
```

Consistency of the prenatal-specific genes between the 2 analyses.

```{r consistency_prenatalSpe, echo=FALSE}
merged_data_prenatalSpe <- merged_data[which(merged_data$ensemblID %in% prenatalSpe_genes), ]
print(sprintf("Number of genes compared = %s", dim(merged_data_prenatalSpe)[1]))
# number of consistent genes
nbr_consistent <- sum(sign(merged_data_prenatalSpe$t_pseudo) == sign(merged_data_prenatalSpe$t_devStages))
consistency <- nbr_consistent/length(merged_data_prenatalSpe$ensemblID)
# compare this consistency score with the consistency of list of random genes
cons_values <- rep(NA, 1000)
# to do 1000 times
for (i in 1:1000){
  # randomly select the same number of genes/transcripts in the data2 to test the overlap with the data1 signif genes
  selection <- sample(background_data$ensemblID, length(merged_data_prenatalSpe$ensemblID))
  nbr_consistent_th <- sum(sign(background_data$logFC_pseudo[which(background_data$ensemblID %in% selection)]) == sign(background_data$logFC_devStages[which(background_data$ensemblID %in% selection)]))
  cons_values[i] <- nbr_consistent_th/length(merged_data_prenatalSpe$ensemblID)
}
cons_values <- as.data.frame(cons_values)
# plot the distribution of the predited consistency and the real value
p <- ggplot(cons_values, aes(cons_values))
p <- p + geom_histogram(binwidth=0.02, boundary = 0, fill="grey", color="black")
p <- p + theme_bw() + xlab("consistency") + ylab("Frequency")
p <- p + scale_x_continuous(limits = c(0, 1))
p <- p + geom_vline(xintercept = consistency, color = "red")
print(p)
# test this consistency with a permutation test
nsmaller <- sum(cons_values >= consistency)
permutation_test <- round((nsmaller+1)/(1000+1), 3)
# test this consistency with a binomial test
result_test <- binom.test(nbr_consistent, length(merged_data_prenatalSpe$ensemblID), p = 0.5)
# print results
print(sprintf("number of consistent gene = %s", nbr_consistent))
print(sprintf("consistency = %s", consistency))
print(sprintf("permutation test p-value = %s", permutation_test))
print(sprintf("binomial test p-value = %s", result_test$p.value))
```


Consistency of the shared genes between the 2 analyses.

```{r consistency_shared, echo=FALSE}
merged_data_shared <- merged_data[which(merged_data$ensemblID%in% shared_genes), ]
print(sprintf("Number of genes compared = %s", dim(merged_data_shared)[1]))
# number of consistent genes
nbr_consistent <- sum(sign(merged_data_shared$t_pseudo) == sign(merged_data_shared$t_devStages))
consistency <- nbr_consistent/length(merged_data_shared$ensemblID)
# compare this consistency score with the consistency of list of random genes
cons_values <- rep(NA, 1000)
# to do 1000 times
for (i in 1:1000){
  # randomly select the same number of genes/transcripts in the data2 to test the overlap with the data1 signif genes
  selection <- sample(background_data$ensemblID, length(merged_data_shared$ensemblID))
  nbr_consistent_th <- sum(sign(background_data$logFC_pseudo[which(background_data$ensemblID %in% selection)]) == sign(background_data$logFC_devStages[which(background_data$ensemblID %in% selection)]))
  cons_values[i] <- nbr_consistent_th/length(merged_data_shared$ensemblID)
}
cons_values <- as.data.frame(cons_values)
# plot the distribution of the predited consistency and the real value
p <- ggplot(cons_values, aes(cons_values))
p <- p + geom_histogram(binwidth=0.02, boundary = 0, fill="grey", color="black")
p <- p + theme_bw() + xlab("consistency") + ylab("Frequency")
p <- p + scale_x_continuous(limits = c(0, 1))
p <- p + geom_vline(xintercept = consistency, color = "red")
print(p)
# test this consistency with a permutation test
nsmaller <- sum(cons_values >= consistency)
permutation_test <- round((nsmaller+1)/(1000+1), 3)
# test this consistency with a binomial test
result_test <- binom.test(nbr_consistent, length(merged_data_shared$ensemblID), p = 0.5)
# print results
print(sprintf("number of consistent gene = %s", nbr_consistent))
print(sprintf("consistency = %s", consistency))
print(sprintf("permutation test p-value = %s", permutation_test))
print(sprintf("binomial test p-value = %s", result_test$p.value))
```

The prenatal-specific sex-DE genes are highly consistent in their direction of effect (% consistency = 86.5%, permutation p-value = 0.001).
While the shared sex-DE genes are as consistent as a random set of non-sex-DE genes (% consistency = 67.0%, permutation p-value = 0.528).

### Replication rate

We also compute the replication rate by group of genes.

For prenatal-specific sex-DE genes:

```{r pi1_prenatakSpe, echo=FALSE}
pi1_score(results_pseudo[which(results_pseudo$ensemblID %in% prenatalSpe_genes), ], results_devStage)
```

The replication rate for the prenatal-specific sex-DE genes is 0.398 in the analysis using developmental stages as covariate. This is slightly lower than the overall pi1 for all prenatal sex-DE genes (pi1 = 0.421)

For shared sex-DE genes:

```{r pi1_shared, echo=FALSE}
pi1_score(results_pseudo[which(results_pseudo$ensemblID %in% shared_genes), ], results_devStage)
```

The replication rate for the shared sex-DE genes is 0.243 in the analysis using developmental stages as covariate. This is a bit higher than the overall pi1 for the full set of genes analyzed in the prenatal analysis (pi1 = 0.207).


# Session info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

