---
title: "sex-DE dream analysis on HDBR forebrain samples using pseudotime as covariate"
author: "Clara Benoit-Pilven"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
---

```{r load_libraries, echo=FALSE}
suppressPackageStartupMessages( library(edgeR) )
suppressPackageStartupMessages( library(limma) )
suppressPackageStartupMessages( library(qvalue) )
suppressPackageStartupMessages( library(sva) )
suppressPackageStartupMessages( library(doParallel) )
suppressPackageStartupMessages( library(variancePartition) )
```

```{r parallel_processing_param, echo=FALSE}
# Specify parallel processing parameters
# this is used implicitly by dream() to run in parallel
param <- SnowParam(4, "SOCK", progressbar=TRUE)
register(param)
```

```{r, setup, include=FALSE, echo=FALSE}
knitr::opts_knit$set(root.dir = '~/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/')
```

# Sex-DE analysis on genes with pseudotime covariate

## Load the counts and the phenotype data

```{r load_data, echo=FALSE}
load("data/1-preprocessing/counts_forebrain_phenotypes.RData")
load("data/2-pseudotime/forebrain_metadata_pseudotime.RData")
metadata <- metadata[,c(1,6,3)]
```

## Prepare design

```{r prepare_metadata, echo=FALSE}
# design for voomDream (the variable to be tested must be a fixed effect)
form <- ~ Sex + pseudotime + Sex:pseudotime + (1|Individual)
# design for sva computation
design <- model.matrix(~metadata$Sex + metadata$pseudotime + metadata$Sex:metadata$pseudotime)
```

## Prepare counts for DE analysis

Estimate weights using linear mixed model of dream

```{r voom_norm, echo=FALSE, warning=FALSE}
png('images/3-DE/trendplot/trendplot_VoomDream_forebrain_pseudotime_ind_interaction.png')
suppressWarnings( vobjDream <- voomWithDreamWeights(x, form, metadata, plot = TRUE) )
dev.off()
```

Estimate surrogates variables (svas) and add them to the model.

```{r estimate_sva, echo=FALSE}
n.sv <- num.sv(vobjDream$E, design, method = "be")
sva.obj <- sva(vobjDream$E, design, method = "irw", n.sv = n.sv)
# add svas to metadata
metadata <- cbind(metadata, sva.obj$sv)
names(metadata) <- c("Sex", "pseudotime", "Individual", paste0("X", 1:n.sv))
# include svas in formula
form <- reformulate(c("Sex", "pseudotime", "Sex:pseudotime", "(1|Individual)", paste0("X", 1:n.sv)))
```

Redo voom normalization with new design matrix (containing svas).

```{r voom_norm_svas, echo=FALSE, warning=FALSE}
suppressWarnings( y <- voomWithDreamWeights(x, form, metadata, plot = TRUE) )
```

Save the normalized counts and the phenotypes vectors in a RData file.

```{r save_normCounts, echo=FALSE}
save(y, phenotypes, form, sva.obj, metadata,
     file = "data/3-DE/normCounts_dream_pseudotime_sva_ind_interaction_phenotypes.RData")
```


## Sex-Differential Expression (sex-DE) analysis

Fit the dream model on each gene.

```{r sexDE_analysis, echo=FALSE}
# By default, uses the Satterthwaite approximation for the hypothesis test
suppressWarnings( fitmm <- dream(y, form, metadata) )
fitmm2 <- topTable(fitmm, coef = 2, number = 30000, sort.by = "none") 
fitmm2$qvalue <- qvalue(fitmm2$P.Value)$qvalues
```

Save the fitmm object.
```{r save_fitmm, echo=FALSE}
save(fitmm, file = "data/3-DE/fitmm_dream_pseudotime_sva_ind_interaction_phenotypes.RData")
```

Compute the mean expression of each genes for male and female.

```{r compute_mean_expr, echo=FALSE}
fitmm2$Male <- rowMeans(y$E[,which(metadata$Sex==1)]) # for male samples
fitmm2$VarMale <- rowVars(y$E[,which(metadata$Sex==1)])
fitmm2$Female <- rowMeans(y$E[,which(metadata$Sex==2)]) # for female samples
fitmm2$VarFemale <- rowVars(y$E[,which(metadata$Sex==2)])
```

Save results.

```{r write_output, echo=FALSE}
write.table(fitmm2, 
            "data/3-DE/VoomDream_topTable_forebrain_pseudotime_sva_ind_interaction.txt", 
            col.names = TRUE, row.names = FALSE, quote = FALSE)
```

Extract the interaction results.

```{r interaction_results, echo=FALSE}
colnames(fitmm)
fitmm_interaction <- topTable(fitmm, coef = "Sex2:pseudotime", number = 30000, sort.by = "none") 
fitmm_interaction$qvalue <- qvalue(fitmm_interaction$P.Value)$qvalues
```

Save the interaction results.

```{r save_interaction, echo=FALSE}
write.table(fitmm_interaction, 
            "data/3-DE/VoomDream_topTable_forebrain_pseudotime_sva_ind_interaction_term.txt", 
            col.names = TRUE, row.names = FALSE, quote = FALSE)
```

Extract the results of the pseudotime-DE genes.

```{r ipseudotime_DE_results, echo=FALSE}
fitmm_pseudotime <- topTable(fitmm, coef = "pseudotime", number = 30000, sort.by = "none") 
fitmm_pseudotime$qvalue <- qvalue(fitmm_pseudotime$P.Value)$qvalues
```

Save the interaction results.

```{r save_pseudotime, echo=FALSE}
write.table(fitmm_pseudotime, 
            "data/3-DE/VoomDream_topTable_forebrain_pseudotime_sva_ind_interaction_pseudotime.txt", 
            col.names = TRUE, row.names = FALSE, quote = FALSE)
```

# Session Info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

