---
title: "Factors that correlate with pseudotime"
author: "Clara Benoit-Pilven"
date: "`r Sys.Date()`"
output: html_document
---

```{r library, echo=FALSE}
suppressPackageStartupMessages(library(edgeR))
suppressPackageStartupMessages(library(limma))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(SummarizedExperiment))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tibble))
suppressPackageStartupMessages(library(readxl))
suppressPackageStartupMessages(library(naniar))
```

```{r setup, echo=FALSE}
knitr::opts_knit$set(root.dir = '~/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/')
```

# Load data

## Raw counts

Load raw counts outputed by Salmon.

```{r load_counts, echo=FALSE}
counts <- fread(file="data/1-preprocessing/counts_PRJEB14594.txt", sep = "\t", fill = TRUE)
# reformat
rownames(counts) <- counts$Name
counts <- setNames(data.frame(t(counts[,-1])), rownames(counts))
```

And compute a proxy for the library size using the total number of counts for each samples.

```{r compute_library_size, echo=FALSE}
totalCounts <- rowSums(counts, na.rm = TRUE)
names(totalCounts) <- names(counts)
```

Load the phenotype file to match the sample names with the ones used in the metadata.

```{r load_phenotypes, echo=FALSE}
samples <- read.table("data/1-preprocessing/PRJEB14954_samples_with_counts_info.txt", 
                      header = FALSE, sep = "\t", fill = TRUE)
names(samples) <- c("experiment_name", "ind", "dev_stage", "brain_region", "karyotype", "batch", "sample_name")
# order the df to match the counts df
samples <- samples[match(names(totalCounts), samples$sample_name),]
# rename the totalCounts names with the experiment_name to match the names used in the pseudotime metadata
names(totalCounts) <- samples$experiment_name
```

Sum the samples coming from the same sample (same experiment name).

```{r sum_technicalRep, echo=FALSE}
totalCounts_merged <- data.frame(sampleName = names(totalCounts), totalCounts = totalCounts) %>%
  group_by(sampleName) %>%
  summarise(totalCounts = sum(totalCounts), .groups = "drop")
```

Then, to save memory, we delete the raw counts from the R session.

```{r delete_counts_data, echo=FALSE}
rm(counts)
```


## Pseudotime from original analysis

Load the pseudotime 

```{r load_pseudotime, echo=FALSE}
load("data/2-pseudotime/forebrain_metadata_pseudotime.RData")
```


## Pseudotime from analysis using the developmental stages as covariate

Load the pseudotime

```{r load_pseudotime_devStage, echo=FALSE}
loadRData <- function(fileName){
  #loads an RData file, and returns it
  load(fileName)
  get(ls()[ls() != "fileName"])
}
metadata_devStage <- loadRData("data/2-pseudotime/analysis_withDevStage/forebrain_metadata_pseudotime_withDevStage.RData")
```


## SVA from pseudotime analysis

Load only sva object.

```{r load_sva_pseudotime, echo=FALSE}
loadObject <- function(fileName, objectName){
  #loads an RData file, and returns it
  load(fileName)
  get(ls()[ls() == objectName])
}
sva_pseudotime <- loadObject("data/3-DE/normCounts_dream_pseudotime_sva_ind_interaction_phenotypes.RData", "sva.obj")
```

## SVA from devStage analysis


```{r load_sva_devStage, echo=FALSE}
sva_devStage <- loadObject("data/3-DE/analysis_devStage/normCounts_dream_dev_sva_ind_interaction_phenotypes.RData", "sva.obj")
```


## Load other HDBR metadata

Read excel file with the metadata contained in the article. This file contains one additional information called "time sample in transit".

```{r load_hdbr_metadata, echo=FALSE}
hdbr_add_metadata <- read_excel("data/0-input/HDBR_AmendedSupplementaryTable1.xlsx",
                                 sheet = 1, skip = 6)
# rename columns
names(hdbr_add_metadata) <- c("Ind", "DevelopmentalStage", "Sample", "BrainRegion", "LeftRight", "Karyotype", "TimeInTransit")
# replace "UNKNOWN" values by NA
hdbr_add_metadata <- replace_with_na_all(hdbr_add_metadata, condition = ~.x == "UNKNOWN")
# convert TimeInTransit to a numeric
hdbr_add_metadata$TimeInTransit <- as.numeric(hdbr_add_metadata$TimeInTransit)
```

Merge this metadata with our original metadata and the metadata from the pseudotime with the devStage covariate analysis.

```{r merge_metadata, echo=FALSE}
metadata_merged <- merge(metadata, hdbr_add_metadata[, c(3,5,7)], by.x = 0, by.y = "Sample", all.x = TRUE)
metadata_devStage_merged <- merge(metadata_devStage, hdbr_add_metadata[, c(3,5,7)], by.x = 0, by.y = "Sample", all.x = TRUE)
```



# Correlation between pseudotime and technical factors

## For original pseudotime

### Pseudotime vs library size

Merge the library size with the metadata containing the pseudotime.

```{r merge_librarySize_pseudotime, echo=FALSE}
metadata_librarySize <- merge(metadata, totalCounts_merged, by.x = 0, by.y = 1)
names(metadata_librarySize)[1] <- "sampleName"
```

Look at the correlation between the pseudotime and the library size.

```{r cor_librarySize_pseudotime, echo=FALSE}
cor.test(metadata_librarySize$pseudotime, metadata_librarySize$totalCounts)
```

```{r plot_cor_librarySize_pseudotime, echo=FALSE}
p <- ggplot(metadata_librarySize, aes(x = pseudotime, y = totalCounts))
p <- p + geom_point()
p <- p + theme_minimal()
p <- p + scale_y_log10()
p <- p + ylab("Library size\nproxied by total count")
p
# save plot
ggsave("pseudotime_vs_librarySize.png", 
       path = "images/2-pseudotime/pseudotimeVStechnicalFactors/",
       plot = p, width = 15, height = 10, units = "cm")
```


### Pseudotime vs Time in transit

For this analysis some samples will be ignored as there is no information about the time in transit for the sample.

```{r cor_TimeInTransit_pseudotime_devStage, echo=FALSE}
cor.test(metadata_merged$pseudotime, metadata_merged$TimeInTransit)
```

```{r plot_cor_TimeInTransit_pseudotime_devStage, echo=FALSE}
p <- ggplot(metadata_merged, aes(x = pseudotime, y = TimeInTransit))
p <- p + geom_point()
p <- p + theme_minimal()
p <- p + ylab("Time in transit (in minutes)")
p
# save plot
ggsave("pseudotime_vs_timeInTransit.png", 
       path = "images/2-pseudotime/pseudotimeVStechnicalFactors/",
       plot = p, width = 15, height = 10, units = "cm")
```


### Pseudotime vs sva from original analysis

Look at the correlation of the 22 original svas with the pseudotime from the original analysis.

```{r cor_svaOriginal_vs_pseudotime, echo=FALSE}
for (i in 1:sva_pseudotime$n.sv){
  print(sprintf("Pseudotime original vs SVAs original %i", i))
  corTestRes <- cor.test(metadata$pseudotime, sva_pseudotime$sv[,i])
  print(corTestRes)
}
```

Only the second sva is significantly correlated with pseudotime (r=0.127 and p-value=0.037).


### Pseudotime vs sva from devStage analysis

Look at the correlation of the 22 svas from the devStage analysis with the pseudotime from the original analysis.

```{r cor_svaDevStage_vs_pseudotime, echo=FALSE}
for (i in 1:sva_devStage$n.sv){
  print(sprintf("Pseudotime original vs SVAs from devStage analysis %i", i))
  corTestRes <- cor.test(metadata$pseudotime, sva_devStage$sv[,i])
  print(corTestRes)
}
```


## For pseudotime with devStage covariate

### Pseudotime vs library size

Merge the library size with the metadata containing the pseudotime.

```{r merge_librarySize_pseudotimeDevStage, echo=FALSE}
metadata_devStage_librarySize <- merge(metadata_devStage, totalCounts_merged, by.x = 0, by.y = 1)
names(metadata_devStage_librarySize)[1] <- "sampleName"
```

Look at the correlation between the pseudotime and the library size.

```{r cor_librarySize_pseudotimeDevStage, echo=FALSE}
cor.test(metadata_devStage_librarySize$pseudotime, metadata_devStage_librarySize$totalCounts)
```


```{r plot_cor_librarySize_pseudotimeDevStage, echo=FALSE}
p <- ggplot(metadata_devStage_librarySize, aes(x = pseudotime, y = totalCounts))
p <- p + geom_point()
p <- p + theme_minimal()
p <- p + scale_y_log10()
p <- p + ylab("Library size\nproxied by total count") + xlab("Pseudotime (devStage cov.)")
p
# save plot
ggsave("pseudotimeDevStage_vs_librarySize.png", 
       path = "images/2-pseudotime/pseudotimeVStechnicalFactors/",
       plot = p, width = 15, height = 10, units = "cm")
```


### Pseudotime vs Time in transit

For this analysis some samples will be ignored as there is no information about the time in transit for the sample.

```{r cor_TimeInTransit_pseudotime, echo=FALSE}
cor.test(metadata_devStage_merged$pseudotime, metadata_devStage_merged$TimeInTransit)
```

```{r plot_cor_TimeInTransit_pseudotime, echo=FALSE}
p <- ggplot(metadata_devStage_merged, aes(x = pseudotime, y = TimeInTransit))
p <- p + geom_point()
p <- p + theme_minimal()
p <- p + ylab("Time in transit (in minutes)")
p
# save plot
ggsave("pseudotimeDevStage_vs_timeInTransit.png", 
       path = "images/2-pseudotime/pseudotimeVStechnicalFactors/",
       plot = p, width = 15, height = 10, units = "cm")
```


### Pseudotime vs sva from original analysis

Look at the correlation of the 22 original svas with the pseudotime from the original analysis.

```{r cor_svaOriginal_vs_pseudotimeDevStage, echo=FALSE}
for (i in 1:sva_pseudotime$n.sv){
  print(sprintf("Pseudotime devStage vs SVAs original %i", i))
  corTestRes <- cor.test(metadata_devStage$pseudotime, sva_pseudotime$sv[,i])
  print(corTestRes)
}
```

5 different SVAs from the original analysis have a significant correlation with the pseudotime with devStage as covariate.
The maximum Pearson's correlation is r=0.22.


### Pseudotime vs sva from devStage analysis

Look at the correlation of the 22 svas from the devStage analysis with the pseudotime from the analysis with the devStage as covariates..

```{r cor_svaDevStage_vs_pseudotimeDevStage, echo=FALSE}
for (i in 1:sva_devStage$n.sv){
  print(sprintf("Pseudotime original vs SVAs from devStage analysis %i", i))
  corTestRes <- cor.test(metadata_devStage$pseudotime, sva_devStage$sv[,i])
  print(corTestRes)
}
```


# Session Info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

