---
title: "Pseudotime analysis in prenatal forebrain"
author: "Clara Benoit-Pilven"
date: "5/12/2021"
output:
  html_document:
    df_print: paged
    toc: true
---

```{r library, echo=FALSE}
suppressPackageStartupMessages(library(edgeR))
suppressPackageStartupMessages(library(limma))
suppressPackageStartupMessages(library(SummarizedExperiment))
suppressPackageStartupMessages(library(ade4))
suppressPackageStartupMessages(library(factoextra))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tidyr))
#suppressPackageStartupMessages(library(forcats))
#suppressPackageStartupMessages(library(ltm))
#suppressPackageStartupMessages(library(Kendall))
```

```{r, setup, include=FALSE, echo=FALSE}
knitr::opts_knit$set(root.dir = '~/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/')
```

# Load and prepare data

## Load the counts and the phenotype data

```{r load_data, echo=FALSE}
load("data/1-preprocessing/counts_forebrain_phenotypes.RData")
```

## Load pseudotime results

```{r load_pseudotime, echo=FALSE}
load("data/2-pseudotime/forebrain_pseudotime_analysis.RData")
```


# Plots

## Histogram

Plot the distribution of the pseudotime.

```{r distrib_pseudotime, echo=FALSE}
hist(metadata$pseudotime, main = "Histogram of pseudotime", xlab = "Pseudotime")
png("images/2-pseudotime/distrib_pseudotime.png",
    width = 900, height = 800)
hist(metadata$pseudotime, main = "Histogram of pseudotime", xlab = "Pseudotime")
dev.off() 
```


## PCA plots

Next, we plot the samples in a PCA plot coloring by pseudotime for both sexes together and for each sex separatly


```{r select_counts_PCA, echo=FALSE}
# select higly variable genes
variance <- rowVars(x$counts)
ntop <- min(2000, dim(x$counts)[1])
selectntop <- order(variance, decreasing=TRUE)[seq_len(ntop)]
selectedCounts_PCA <- x$counts[selectntop,]
```

```{r do_PCA, echo=FALSE}
pcaHVG <- dudi.pca(t(selectedCounts_PCA), scannf = FALSE, nf = 10)
varExplained <- pcaHVG$eig/sum(pcaHVG$eig)*100
pcaHVG.ind <- get_pca_ind(pcaHVG)
pcaHVG.var <- get_pca_var(pcaHVG)
```

```{r plot_PCA_pseudotime_color, echo=FALSE}
p <- ggplot(pcaHVG.ind$coord, aes(x = Dim.1, y = Dim.2))
p <- p + geom_point(aes(color = metadata$pseudotime))
p <- p + labs(title = "PCA" , 
              x = sprintf("PC%s (%s %%)", 1, round(varExplained[1])), 
              y = sprintf("PC%s (%s %%)", 2, round(varExplained[2])))
p <- p + theme_minimal()
p <- p + scale_color_gradient(low = "blueviolet", high = "gold", name = "pseudotime")
p
# save plot
ggsave("PCA_1vs2.png",
       path = "images/2-pseudotime/PCA/",
       plot = p, width = 12, height = 10, units = "cm",
       create.dir = TRUE)
```

```{r plot_PCA_pseudotime_color_1vs3, echo=FALSE}
p <- ggplot(pcaHVG.ind$coord, aes(x = Dim.1, y = Dim.3))
p <- p + geom_point(aes(color = metadata$pseudotime))
p <- p + labs(title = "PCA" , 
              x = sprintf("PC%s (%s %%)", 1, round(varExplained[1])), 
              y = sprintf("PC%s (%s %%)", 3, round(varExplained[3])))
p <- p + theme_minimal()
p <- p + scale_color_gradient(low = "blueviolet", high = "gold", name = "pseudotime")
p
# save plot
ggsave("PCA_1vs3.png",
       path = "images/2-pseudotime/PCA/",
       plot = p, width = 12, height = 10, units = "cm")
```

Faceting using the sex of the samples.

```{r plot_PCA_pseudotime_color_sex_facet, echo=FALSE}
dfPlot <- merge(pcaHVG.ind$coord, metadata, by = 0)
p <- ggplot(dfPlot, aes(x = Dim.1, y = Dim.2))
p <- p + geom_point(aes(color = pseudotime))
p <- p + labs(title = "PCA" , 
              x = sprintf("PC%s (%s %%)", 1, round(varExplained[1])), 
              y = sprintf("PC%s (%s %%)", 2, round(varExplained[2])))
p <- p + theme_minimal()
p <- p + scale_color_gradient(low = "blueviolet", high = "gold", name = "pseudotime")
p <- p + facet_grid(. ~ metadata$Sex)
p
# save plot
ggsave("PCA_1vs2_sex.png",
       path = "images/2-pseudotime/PCA/",
       plot = p, width = 15, height = 10, units = "cm")
```

```{r plot_PCA_pseudotime_color_sex_facet_2vs3, echo=FALSE}
dfPlot <- merge(pcaHVG.ind$coord, metadata, by = 0)
p <- ggplot(dfPlot, aes(x = Dim.2, y = Dim.3))
p <- p + geom_point(aes(color = pseudotime))
p <- p + labs(title = "PCA" , 
              x = sprintf("PC%s (%s %%)", 2, round(varExplained[2])), 
              y = sprintf("PC%s (%s %%)", 3, round(varExplained[3])))
p <- p + theme_minimal()
p <- p + scale_color_gradient(low = "blueviolet", high = "gold", name = "pseudotime")
p <- p + facet_grid(. ~ metadata$Sex)
p
# save plot
ggsave("PCA_2vs3_sex.png",
       path = "images/2-pseudotime/PCA/",
       plot = p, width = 15, height = 10, units = "cm")
```

Plot different covariates on the PCA.

```{r PCA_sex, echo=FALSE}
p <- ggplot(pcaHVG.ind$coord, aes(x = Dim.1, y = Dim.2))
p <- p + geom_point(aes(color = metadata$Sex))
p <- p + labs(title = "PCA" , 
              x = sprintf("PC%s (%s %%)", 1, round(varExplained[1])), 
              y = sprintf("PC%s (%s %%)", 2, round(varExplained[2])))
p <- p + theme_minimal()
p <- p + scale_color_manual(breaks = c("Female" = "2", "Male" = "1"), values = c("#B2182B", "#074c9b"), name = "Sex")
p
# save plot
ggsave("PCA_sex_1vs2.png",
       path = "images/2-pseudotime/PCA/",
       plot = p, width = 12, height = 10, units = "cm")
```

```{r PCA_region, echo=FALSE}
p <- ggplot(pcaHVG.ind$coord, aes(x = Dim.1, y = Dim.2))
p <- p + geom_point(aes(color = metadata$ForebrainRegion))
p <- p + labs(title = "PCA" , 
              x = sprintf("PC%s (%s %%)", 1, round(varExplained[1])), 
              y = sprintf("PC%s (%s %%)", 2, round(varExplained[2])))
p <- p + theme_minimal()
p <- p + scale_color_discrete(name = "Forebrain Region")
p
# save plot
ggsave("PCA_region_1vs2.png",
       path = "images/2-pseudotime/PCA/",
       plot = p, width = 15, height = 10, units = "cm")
```

```{r PCA_devStage, echo=FALSE}
p <- ggplot(pcaHVG.ind$coord, aes(x = Dim.1, y = Dim.2))
p <- p + geom_point(aes(color = metadata$DevStage))
p <- p + labs(title = "" , 
              x = sprintf("PC%s (%s %%)", 1, round(varExplained[1])), 
              y = sprintf("PC%s (%s %%)", 2, round(varExplained[2])))
p <- p + theme_minimal()
p <- p + scale_color_discrete(name = "Developmental\nstage")
p <- p + theme(legend.position="bottom")
p <- p + guides(color=guide_legend(nrow=3,byrow=TRUE))
p
# save plot
ggsave("PCA_devStage_1vs2.png",
       path = "images/2-pseudotime/PCA/",
       plot = p, width = 12, height = 13, units = "cm")
```

```{r PCA_batch, echo=FALSE}
p <- ggplot(pcaHVG.ind$coord, aes(x = Dim.1, y = Dim.2))
p <- p + geom_point(aes(color = metadata$Batch, shape = metadata$Batch))
p <- p + labs(title = "PCA" , 
              x = sprintf("PC%s (%s %%)", 1, round(varExplained[1])), 
              y = sprintf("PC%s (%s %%)", 2, round(varExplained[2])))
p <- p + theme_minimal()
p <- p + scale_color_discrete(name = "Batch") + scale_shape_manual(values = c(1:25, 35:38, 49:55), name = "Batch")
p <- p + theme(legend.key.size = unit(0.3, 'cm'))
p
# save plot
ggsave("PCA_batch_1vs2.png",
       path = "images/2-pseudotime/PCA/",
       plot = p, width = 18, height = 10, units = "cm")
```


## Boxplot

Finally, we do boxplot of the pseudotime per developmental stages stratified by sex or not.

```{r boxplot, echo=FALSE}
p <- ggplot(metadata, aes(x = DevStage, y = pseudotime))
p <- p + geom_boxplot(fill = "#DCDCDC")
p <- p + geom_jitter(position=position_jitter(0.2))
p <- p + theme_minimal()
p
# save plot
ggsave("boxplot_devStages_vs_pseudotime.png",
       path = "images/2-pseudotime/",
       plot = p, width = 15, height = 10, units = "cm")
```

```{r boxplot_devstage_sex, echo=FALSE}
p <- ggplot(metadata, aes(x = DevStage, y = pseudotime))
p <- p + geom_boxplot(aes(fill = Sex))
p <- p + scale_fill_manual(breaks = c(2, 1), values = c("#B2182B", "#074c9b"), labels = c("Females", "Males"))
p <- p + theme_minimal()
p
# save plot
ggsave("boxplot_devStages_vs_pseudotime_sex.png",
       path = "images/2-pseudotime/",
       plot = p, width = 15, height = 10, units = "cm")
```

```{r boxplot_sex, echo=FALSE}
p <- ggplot(metadata, aes(x = Sex, y = pseudotime))
p <- p + geom_boxplot(aes(fill = Sex))
p <- p + scale_fill_manual(breaks = c(2, 1), values = c("#B2182B", "#074c9b"), labels = c("Females", "Males"))
p <- p + theme_minimal()
p <- p + stat_compare_means(method = "wilcox.test")
p
```

Test difference in pseudotime variance between males and females.

```{r test_variance_sex, echo=FALSE}
var.test(pseudotime ~ Sex, metadata, 
         alternative = "two.sided")
```


# Test pseudotime consistency per individuals

Check that for each individual, the pseudotimes of the different samples are similar.

```{r check_pseudotime_individuals, echo=FALSE}
sdPseudotime_ind <- metadata %>% group_by(Individual) %>% summarise(sd = sd(pseudotime))
meanSdPseudotime <- mean(sdPseudotime_ind$sd, na.rm = TRUE)
print(sprintf("Mean observed pseudotime standard deviation = %s", meanSdPseudotime))
# Is the sd as big if we permute the pseudotimes?
meanSdPseudotime_permut <- c()
for (i in 1:1000){
  permutated_pseudotime <- sample(metadata$pseudotime)
  tmp_df <- data.frame(Individual = metadata$Individual, pseudotime = permutated_pseudotime)
  tmp_sdPseudotime_ind <- tmp_df %>% group_by(Individual) %>% summarise(sd = sd(pseudotime))
  meanSdPseudotime_permut <- c(meanSdPseudotime_permut, mean(tmp_sdPseudotime_ind$sd, na.rm = TRUE))
}
meanSdPseudotime_permut <- as.data.frame(meanSdPseudotime_permut)
print(sprintf("Mean expected pseudotime standard deviation = %s", mean(meanSdPseudotime_permut$meanSdPseudotime_permut)))
# pvalue
(1 + sum(meanSdPseudotime_permut<meanSdPseudotime))/1000
# Plot
p <- ggplot(meanSdPseudotime_permut, aes(x = meanSdPseudotime_permut))
p <- p + geom_histogram(binwidth = 0.01, fill = "grey", color = "black")
p <- p + geom_vline(xintercept = meanSdPseudotime,
                    color = "red")
p <- p + labs(x = "SD of the pseudotime per individuals")
p <- p + theme_minimal()
p
# save plot
ggsave("permutation_sd_pseudotime.png",
       path = "images/2-pseudotime/",
       plot = p, width = 15, height = 10, units = "cm")
```


# Correlation of pseudotime with developmental stages

## With merged developmental stages

Convert the developmental stages to a continuous variable.

```{r convert2continuous, echo=FALSE}
metadata$DevStageCont <- as.integer(metadata$DevStage)
```

### For both sexes

Compute Kendall correlation between the pseudotime and the new continuous developmental stage.

```{r correlation_kendall, echo=FALSE}
corTestRes_Kendall <- cor.test(metadata$DevStageCont, metadata$pseudotime, method = "kendall")
corTestRes_Kendall
```

And also do a scatter plot of the pseudotime and the continuous developmental stage.

```{r scatter_plot, echo=FALSE}
p <- ggplot(metadata, aes(DevStage, pseudotime))
p <- p + geom_boxplot(fill = "#DCDCDC", outlier.shape = NA)
p <- p + theme_minimal() + xlab("Developmental stages")
p <- p + geom_jitter(aes(color = Sex), position=position_jitter(0.2), show.legend = FALSE)
p <- p + annotate("text", x = 1.5, y = 1.1, label = sprintf("tau = %.2f\n p-value = %.2e", corTestRes_Kendall$estimate, corTestRes_Kendall$p.value))
p <- p + scale_color_manual(values = c("1" = "darkblue", "2" = "darkred"))
p
# save plot
ggsave("boxplot_devStages_vs_pseudotime.png", 
       path = "images/2-pseudotime/",
       plot = p, width = 15, height = 10, units = "cm")
```


Linear regression between devStage and pseudotime.

```{r lm_devStage_pseudotime, echo=FALSE}
summary(lm(metadata$DevStageCont~metadata$pseudotime))
lm_eqn <- function(df){
    m <- lm(pseudotime ~ DevStageCont, df);
    eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2, 
         list(a = format(unname(coef(m)[1]), digits = 2),
              b = format(unname(coef(m)[2]), digits = 2),
             r2 = format(summary(m)$r.squared, digits = 3)))
    as.character(as.expression(eq));
}
```

Add this linear model to the boxplot.

```{r scatter_plot_lm, echo=FALSE}
p <- ggplot(metadata, aes(DevStage, pseudotime))
p <- p + geom_boxplot(fill = "#DCDCDC")
p <- p + theme_minimal()
p <- p + geom_jitter(position=position_jitter(0.2))
p <- p + geom_smooth(aes(DevStageCont, pseudotime), method = lm, formula = "y ~ x", color = "purple")
p <- p + annotate("text", x = 2, y = 1.1, label = lm_eqn(metadata), parse = TRUE)
p
# save plot
ggsave("boxplot_devStages_vs_pseudotime_lm.png",
       path = "images/2-pseudotime/",
       plot = p, width = 15, height = 10, units = "cm")
```

The slope is 0.14 -> an increase of 1PCW, increase the pseudotime by 0.14.


### For males only

Select only male samples.

```{r select_males, echo=FALSE}
metadata_males <- metadata[which(metadata$Sex == 1),]
```

Compute Kendall correlation between the pseudotime and the new continuous developmental stage.

```{r correlation_kendall_males, echo=FALSE}
corTestRes_Kendall <- cor.test(metadata$DevStageCont, metadata$pseudotime, method = "kendall")
corTestRes_Kendall
```


And also do a scatter plot of the pseudotime and the continuous developmental stage.

```{r scatter_plot_males, echo=FALSE}
p <- ggplot(metadata_males, aes(DevStage, pseudotime))
p <- p + geom_boxplot(fill = "#a1c1e6")
p <- p + theme_minimal()
p <- p + geom_jitter(position=position_jitter(0.2))
p <- p + annotate("text", x = 1.5, y = 1.1, label = sprintf("tau = %.2f\n p-value = %.2e", corTestRes_Kendall$estimate, corTestRes_Kendall$p.value))
p
# save plot
ggsave("boxplot_devStages_vs_pseudotime_males.png",
       path = "images/2-pseudotime/",
       plot = p, width = 15, height = 10, units = "cm")
```


### For females only

Select only female samples.

```{r select_females, echo=FALSE}
metadata_females <- metadata[which(metadata$Sex == 2),]
```

Compute Kendall correlation between the pseudotime and the new continuous developmental stage.

```{r correlation_kendall_females, echo=FALSE}
corTestRes_Kendall <- cor.test(metadata$DevStageCont, metadata$pseudotime, method = "kendall")
corTestRes_Kendall
```

And also do a scatter plot of the pseudotime and the continuous developmental stage.

```{r scatter_plot_females, echo=FALSE}
p <- ggplot(metadata_females, aes(DevStage, pseudotime))
p <- p + geom_boxplot(fill = "#e0a3aa")
p <- p + theme_minimal()
p <- p + geom_jitter(position=position_jitter(0.2))
p <- p + annotate("text", x = 1.5, y = 1.1, label = sprintf("tau = %.2f\n p-value = %.2e", corTestRes_Kendall$estimate, corTestRes_Kendall$p.value))
p
# save plot
ggsave("boxplot_devStages_vs_pseudotime_females.png",
       path = "images/2-pseudotime/",
       plot = p, width = 15, height = 10, units = "cm")
```


# Session Info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
