---
title: "GO analysis with ORA - comparison between sex-DE and pseudotime-DE"
author: "Clara Benoit-Pilven"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
---

```{r load_libraries, echo=FALSE}
suppressPackageStartupMessages( library(SummarizedExperiment) )
suppressPackageStartupMessages( library(dplyr) )
suppressPackageStartupMessages( library(fgsea) )
suppressPackageStartupMessages( library(data.table) )
suppressPackageStartupMessages( library(qvalue) )
suppressPackageStartupMessages( library(ggplot2) )
suppressPackageStartupMessages( library(RColorBrewer) )
suppressPackageStartupMessages( library(limma) )
suppressPackageStartupMessages( library(clusterProfiler) )
suppressPackageStartupMessages( library(enrichplot) )
suppressPackageStartupMessages( library(org.Hs.eg.db) )
suppressPackageStartupMessages( library(gridExtra) )
```

```{r, setup, include=FALSE, echo=FALSE}
knitr::opts_knit$set(root.dir = '~/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/')
```

Set parameters:

```{r parameters, echo=FALSE}
qvalueThreshold <- 0.01
logFCThreshold <- 0
```


# Functions

## Prepare gene list

```{r fct_prepare_geneList, echo=FALSE}
prepareGeneList <- function(data, direction = "UP", qvalueThreshold = 0.01, logFCThreshold = 0){
  geneList <- c()
  if (direction == "UP"){
    geneList <- data$gene_name[which(data$qvalue < qvalueThreshold & data$logFC > logFCThreshold)]
  } else if (direction == "DOWN"){
    geneList <- data$gene_name[which(data$qvalue < qvalueThreshold & data$logFC < logFCThreshold)]
  }
  return(geneList)
}
```


## Enrichment with compareCluster and enrichGO

```{r fct_compareCluster, echo=FALSE}
compare_GOenrich <- function(data1, data2, data1Names, data2Names, universe, ont, OrgDb = org.Hs.eg.db, 
                             keyType = "SYMBOL", minGSSize = 15, maxGSSize = 500, pvalueCutoff = 0.05, 
                             qvalueThreshold = 0.01, logFCThreshold = 0, simplify = TRUE, cutoffSimplify = 0.7){
  # prepare gene lists of up and down regulated genes 
  clusters <- list(prepareGeneList(data1, direction = "UP", qvalueThreshold = qvalueThreshold, logFCThreshold = logFCThreshold),
                   prepareGeneList(data1, direction = "DOWN", qvalueThreshold = qvalueThreshold, logFCThreshold = logFCThreshold),
                   prepareGeneList(data2, direction = "UP", qvalueThreshold = qvalueThreshold, logFCThreshold = logFCThreshold),
                   prepareGeneList(data2, direction = "DOWN", qvalueThreshold = qvalueThreshold, logFCThreshold = logFCThreshold))
  names(clusters) <- c(data1Names, data2Names)
  # do enrichment analysis
  resGO <- compareCluster(clusters, fun = "enrichGO", universe = universe, ont = ont, OrgDb = OrgDb, keyType = keyType, 
                          minGSSize = minGSSize, maxGSSize = maxGSSize, pvalueCutoff = pvalueCutoff)
  if (simplify){
    # simplify the GO terms
    resGO_simplify <- simplify(resGO, cutoff=cutoffSimplify, by="pvalue", select_fun=min)
    return(resGO_simplify)
  } else{
    return(resGO)
  }
}

## TODO: modify this function to do an enrichment analysis separatly for the data1 and data2 using 2 different universe...
```

```{r fct_compareCluster2, echo=FALSE}
compare_GOenrich2 <- function(data1, data2, data1Names, data2Names, universe, ont, OrgDb = org.Hs.eg.db, 
                             keyType = "SYMBOL", minGSSize = 15, maxGSSize = 500, pvalueCutoff = 0.05, 
                             qvalueThreshold = 0.01, logFCThreshold = 0, simplify = TRUE, cutoffSimplify = 0.7){
  # prepare gene lists of up and down regulated genes 
  clusters <- list(data1$gene_name[which(data1$sexDE_subgroup == "Female1")],
                   data1$gene_name[which(data1$sexDE_subgroup == "Female2")],
                   data1$gene_name[which(data1$sexDE_subgroup == "Male")],
                   prepareGeneList(data2, direction = "UP", qvalueThreshold = qvalueThreshold, logFCThreshold = logFCThreshold),
                   prepareGeneList(data2, direction = "DOWN", qvalueThreshold = qvalueThreshold, logFCThreshold = logFCThreshold))
  names(clusters) <- c(data1Names, data2Names)
  # do enrichment analysis
  resGO <- compareCluster(clusters, fun = "enrichGO", universe = universe, ont = ont, OrgDb = OrgDb, keyType = keyType, 
                          minGSSize = minGSSize, maxGSSize = maxGSSize, pvalueCutoff = pvalueCutoff)
  if (simplify){
    # simplify the GO terms
    resGO_simplify <- simplify(resGO, cutoff=cutoffSimplify, by="pvalue", select_fun=min)
    return(resGO_simplify)
  } else{
    return(resGO)
  }
}
```

Function to count the number of GO terms containing a list of keywords.

```{r function_count_GO_keyword, echo=FALSE}
countKeyword <- function(resGO, cluster, keywords){
  # get only GO term enriched in the choosen cluster
  GOcluster <- resGO@compareClusterResult[which(resGO@compareClusterResult$Cluster == cluster), ]
  # Count number of GO terms containing at least one of the keywords
  containKeyword <- Reduce("|", lapply(keywords, grepl, x = GOcluster$Description))
  return(sum(containKeyword))
}
```



## Plot enrichment results

```{r fct_plot_enrichRes, echo=FALSE}
plot_resGO <- function(resGO){
  p <- dotplot(resGO, showCategory = 10, label_format = 100)
  p <- p + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
  p <- p + xlab("")
  p <- p + scale_color_gradient(low = "#e6e6e6", high = "#000000", trans = 'reverse')
  return(p)
}
```


# Load and prepare data

Load data.

```{r load_DEdata, echo=FALSE}
# load pseudotime data
load("data/2-pseudotime/forebrain_pseudotime_analysis.RData")

# load DE
de_results_sex <- read.table("data/3-DE/VoomDream_topTable_forebrain_pseudotime_sva_ind_interaction.txt",
                             header = TRUE, check.names=FALSE)
de_results_pseudotime <- read.table("data/3-DE/VoomDream_topTable_forebrain_pseudotime_sva_ind_interaction_pseudotime.txt",
                             header = TRUE, check.names=FALSE)

# load gene characteristics
gene_characteristics <- read.table("data/5-postprocessing/1-GeneProperties/prenatal/prenatal_forebrain_geneProperties.txt", header = TRUE)

# filter the genes to remove genes used to define the sex of samples
genes2remove <- c("XIST", as.character(de_results_sex$gene_name[which(de_results_sex$chr == "chrY" & de_results_sex$gene_type == "protein_coding")]))
de_results_sex <- de_results_sex[which(!de_results_sex$gene_name %in% genes2remove),]
de_results_pseudotime <- de_results_pseudotime[which(!de_results_pseudotime$gene_name %in% genes2remove),]

# for pseudotime-DE genes remove the genes used to compute pseudotime
genes2remove <- rownames(selectedCounts)
de_results_pseudotime <- de_results_pseudotime[which(!de_results_pseudotime$gene_name %in% genes2remove),]

# add gene characteristics to this data
de_results_sex <- merge(de_results_sex, gene_characteristics[, c(1,2,16)], all.x = TRUE)
de_results_pseudotime <- merge(de_results_pseudotime, gene_characteristics[, c(1,2,16)], all.x = TRUE)

# keep only autosomal genes
de_results_sex_auto <- de_results_sex[which(!de_results_sex$chr %in% c("chrX", "chrY", "chrM")),]
de_results_pseudotime_auto <- de_results_pseudotime[which(!de_results_pseudotime$gene_name %in% c("chrX", "chrY", "chrM")),]
```

From the pseudotime and interaction-DE genes remove the genes used to compute the pseudotime.

```{r remove_pseudotime_genes, echo=FALSE}
genes2remove <- rownames(selectedCounts)
de_results_pseudotime_filtered <- de_results_pseudotime[which(!de_results_pseudotime$gene_name %in% genes2remove),]
```

Create the universe gene list.

```{r universe_sexDE, echo=FALSE}
universe <- as.character(de_results_sex$gene_name)
universe_auto <- as.character(de_results_sex_auto$gene_name)
#universe_pseudotime <- as.character(de_results_pseudotime_filtered$gene_name)
```

Prepare the lists of keywords to search in the enriched GO terms.

```{r keyword_lists, echo=FALSE}
cellCycle <- list("cell cycle", "meisosis", "meiotic", "mitotic", "mitosis", "cytokinesis",
                  "DNA", "chromosome", " chromatid", "chromatin", "spindle", "centrosome", "kinetochore", "centriole",
                  "G1", "G2", "M phase", "S phase", "S transition", "M transition", "metaphase", "anaphase",
                  "segregation", "division", "replication", "fission")
development <- list("development", "morphogenesis", "embryonic", "maturation", "stem cell", "pattern formation", "neural tube")
sensoryPerception <- list("sensory", "perception", "stimulus", "stimuli")
mito <- list("mitochondria", "mitochondrial", "mitochondrion",
             "ATP", "respiration", "respiratory", "apoptotic", "apoptosis",
             "oxidative phosphorylation", "electron transport chain",
             "autophagy", "autophagic", "autophagosome")
neuro <- list("synaptic", "synapse", "presynapse", "postsynapse", "presynaptic", "postsynaptic",
              "neuron", "neurotransmitter", "neuronal", "nervous system",
              "axon", "axonogenesis", "axonal", "dendrite", "dendritic", "neural")
rna <- list("splicing", "spliceosome", "RNA")
```

# Analysis...

## ...on all genes

### Analysis for BP

```{r GOanalysis_BP, echo=FALSE}
resGO_BP <- compare_GOenrich(data1 = de_results_sex, data2 = de_results_pseudotime_filtered, universe = universe, ont = "BP",
                          data1Names = c("female-biased", "male-biased"), data2Names = c("up-regulated", "down-regulated"),
                          qvalueThreshold = qvalueThreshold, logFCThreshold = logFCThreshold, simplify = FALSE)
```

Number of significant GO terms for each gene list.

```{r nbr_signif_BP, echo=FALSE}
resGO_BP@compareClusterResult %>% group_by(Cluster) %>% count()
```

Count number of GO terms with specific keywords.

```{r count_keywords_BP, echo=FALSE}
## Sex-DE genes
# female-biased genes & terms related to cell cycle/division/proliferation & terms related to development
print("Female-biased genes - cell cycle keywords:")
countKeyword(resGO_BP, cluster = "female-biased", 
             keywords = cellCycle)
print("Female-biased genes - development keywords:")
countKeyword(resGO_BP, cluster = "female-biased", 
             keywords = development)
print("Female-biased genes - RNA keywords:")
countKeyword(resGO_BP, cluster = "female-biased", 
             keywords = rna)
print("Female-biased genes - Sensory percetion keywords:")
countKeyword(resGO_BP, cluster = "female-biased", 
             keywords = sensoryPerception)
# male-biased genes & terms related to mitochondria and autophagy & terms related to synapse
print("Male-biased genes - mitochodria keywords:")
countKeyword(resGO_BP, cluster = "male-biased", 
             keywords = mito)
print("Male-biased genes - neuron/synapse keywords:")
countKeyword(resGO_BP, cluster = "male-biased", 
             keywords = neuro)

## Pseudotime-DE genes
# upregulated genes & terms related to mitochondria and autophagy & terms related to synapse
print("Up-regulated genes - mitochodria keywords:")
countKeyword(resGO_BP, cluster = "up-regulated", 
             keywords = mito)
print("Up-regulated genes - neuron/synapse keywords:")
countKeyword(resGO_BP, cluster = "up-regulated", 
             keywords = neuro)
# downregulated genes & terms related to cell cycle/division/proliferation & terms related to development
print("Down-regulated genes - cell cycle keywords:")
countKeyword(resGO_BP, cluster = "down-regulated", 
             keywords = cellCycle)
print("Down-regulated genes - development keywords:")
countKeyword(resGO_BP, cluster = "down-regulated", 
             keywords = development)
print("Down-regulated genes - RNA keywords:")
countKeyword(resGO_BP, cluster = "down-regulated", 
             keywords = rna )
print("Down-regulated genes - Sensory percetion keywords:")
countKeyword(resGO_BP, cluster = "down-biased", 
             keywords = sensoryPerception)
```

```{r plot_GOcompare_BP, echo=FALSE}
p <- plot_resGO(resGO_BP)
p
ggsave("dotplot_BP_sexDE_pseudotimeDE.png", plot = p, 
       path = "images/5-postprocessing/3-GO/prenatal/",
       width = 25, height = 25, units = "cm")
```

Save the results in file.

```{r save_GOcompare_BP, echo=FALSE}
write.table(resGO_BP@compareClusterResult,
            "data/5-postprocessing/3-GO/prenatal/GOcompare_BP.txt", 
            row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")
```


### Analysis for CC

```{r GOanalysis_CC, echo=FALSE}
resGO_CC <- compare_GOenrich(data1 = de_results_sex, data2 = de_results_pseudotime, universe = universe, ont = "CC",
                          data1Names = c("female-biased", "male-biased"), data2Names = c("up-regulated", "down-regulated"),
                          qvalueThreshold = qvalueThreshold, logFCThreshold = logFCThreshold, simplify = FALSE)
```

Number of significant GO terms for each gene list.

```{r nbr_signif_CC, echo=FALSE}
resGO_CC@compareClusterResult %>% group_by(Cluster) %>% count()
```

Count number of GO terms with specific keywords.

```{r count_keywords_CC, echo=FALSE}
## Sex-DE genes
# female-biased genes & terms related to cell cycle/division/proliferation & terms related to development
print("Female-biased genes - cell cycle keywords:")
countKeyword(resGO_CC, cluster = "female-biased", 
             keywords = cellCycle)
print("Female-biased genes - development keywords:")
countKeyword(resGO_CC, cluster = "female-biased", 
             keywords = development)
print("Female-biased genes - rna keywords:")
countKeyword(resGO_CC, cluster = "female-biased", 
             keywords = rna)
print("Female-biased genes - Sensory percetion keywords:")
countKeyword(resGO_CC, cluster = "female-biased", 
             keywords = sensoryPerception)
# male-biased genes & terms related to mitochondria and autophagy & terms related to synapse
print("Male-biased genes - mitochodria keywords:")
countKeyword(resGO_CC, cluster = "male-biased", 
             keywords = mito)
print("Male-biased genes - neuron/synapse keywords:")
countKeyword(resGO_CC, cluster = "male-biased", 
             keywords = neuro)

## Pseudotime-DE genes
# upregulated genes & terms related to mitochondria and autophagy & terms related to synapse
print("Up-regulated genes - mitochodria keywords:")
countKeyword(resGO_CC, cluster = "up-regulated", 
             keywords = mito)
print("Up-regulated genes - neuron/synapse keywords:")
countKeyword(resGO_CC, cluster = "up-regulated", 
             keywords = neuro)
# downregulated genes & terms related to cell cycle/division/proliferation & terms related to development
print("Down-regulated genes - cell cycle keywords:")
countKeyword(resGO_CC, cluster = "down-regulated", 
             keywords = cellCycle)
print("Down-regulated genes - development keywords:")
countKeyword(resGO_CC, cluster = "down-regulated", 
             keywords = development)
print("Down-regulated genes - rna keywords:")
countKeyword(resGO_CC, cluster = "down-regulated", 
             keywords = rna)
print("Down-regulated genes - Sensory percetion keywords:")
countKeyword(resGO_CC, cluster = "down-biased", 
             keywords = sensoryPerception)
```

```{r plot_GOcompare_CC, echo=FALSE}
p <- plot_resGO(resGO_CC)
p
ggsave("dotplot_CC_sexDE_pseudotimeDE.png", plot = p, 
       path = "images/5-postprocessing/3-GO/prenatal/",
       width = 22, height = 20, units = "cm")
```

Save the results in file.

```{r save_GOcompare_CC, echo=FALSE}
write.table(resGO_CC@compareClusterResult,
            "data/5-postprocessing/3-GO/prenatal/GOcompare_CC.txt", 
            row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")
```


### Analysis for MF

```{r GOanalysis_MF, echo=FALSE}
resGO_MF <- compare_GOenrich(data1 = de_results_sex, data2 = de_results_pseudotime, universe = universe, ont = "MF",
                          data1Names = c("female-biased", "male-biased"), data2Names = c("up-regulated", "down-regulated"),
                          qvalueThreshold = qvalueThreshold, logFCThreshold = logFCThreshold, simplify = FALSE)
```

Number of significant GO terms for each gene list.

```{r nbr_signif_MF, echo=FALSE}
resGO_MF@compareClusterResult %>% group_by(Cluster) %>% count()
```

Count number of GO terms with specific keywords.

```{r count_keywords_MF, echo=FALSE}
## Sex-DE genes
# female-biased genes & terms related to cell cycle/division/proliferation & terms related to development
print("Female-biased genes - cell cycle keywords:")
countKeyword(resGO_MF, cluster = "female-biased", 
             keywords = cellCycle)
print("Female-biased genes - development keywords:")
countKeyword(resGO_MF, cluster = "female-biased", 
             keywords = development)
print("Female-biased genes - rna keywords:")
countKeyword(resGO_MF, cluster = "female-biased", 
             keywords = rna)
print("Female-biased genes - Sensory percetion keywords:")
countKeyword(resGO_MF, cluster = "female-biased", 
             keywords = sensoryPerception)
# male-biased genes & terms related to mitochondria and autophagy & terms related to synapse
print("Male-biased genes - mitochodria keywords:")
countKeyword(resGO_MF, cluster = "male-biased", 
             keywords = mito)
print("Male-biased genes - neuron/synapse keywords:")
countKeyword(resGO_MF, cluster = "male-biased", 
             keywords = neuro)

## Pseudotime-DE genes
# upregulated genes & terms related to mitochondria and autophagy & terms related to synapse
print("Up-regulated genes - mitochodria keywords:")
countKeyword(resGO_MF, cluster = "up-regulated", 
             keywords = mito)
print("Up-regulated genes - neuron/synapse keywords:")
countKeyword(resGO_MF, cluster = "up-regulated", 
             keywords = neuro)
# downregulated genes & terms related to cell cycle/division/proliferation & terms related to development
print("Down-regulated genes - cell cycle keywords:")
countKeyword(resGO_MF, cluster = "down-regulated", 
             keywords = cellCycle)
print("Down-regulated genes - development keywords:")
countKeyword(resGO_MF, cluster = "down-regulated", 
             keywords = development)
print("Down-regulated genes - rna keywords:")
countKeyword(resGO_MF, cluster = "down-regulated", 
             keywords = rna)
print("Down-regulated genes - Sensory percetion keywords:")
countKeyword(resGO_MF, cluster = "down-biased", 
             keywords = sensoryPerception)
```

```{r plot_GOcompare_MF, echo=FALSE}
p <- plot_resGO(resGO_MF)
p
ggsave("dotplot_MF_sexDE_pseudotimeDE.png", plot = p, 
       path = "images/5-postprocessing/3-GO/prenatal/",
       width = 22, height = 20, units = "cm")
```

Save the results in file.

```{r save_GOcompare_MF, echo=FALSE}
write.table(resGO_MF@compareClusterResult,
            "data/5-postprocessing/3-GO/prenatal//GOcompare_MF.txt", 
            row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")
```


## ...on autosomal genes

### Analysis for BP

```{r GOanalysis_BP_auto, echo=FALSE}
resGO_BP_auto <- compare_GOenrich(data1 = de_results_sex_auto, data2 = de_results_pseudotime_auto, universe = universe_auto, ont = "BP",
                                  data1Names = c("female-biased", "male-biased"), data2Names = c("up-regulated", "down-regulated"),
                                  qvalueThreshold = qvalueThreshold, logFCThreshold = logFCThreshold, simplify = FALSE)
```

Number of significant GO terms for each gene list.

```{r nbr_signif_BP_auto, echo=FALSE}
resGO_BP_auto@compareClusterResult %>% group_by(Cluster) %>% count()
```

Count number of GO terms with specific keywords.

```{r count_keywords_BP_auto, echo=FALSE}
## Sex-DE genes
# female-biased genes & terms related to cell cycle/division/proliferation & terms related to development
print("Female-biased genes - cell cycle keywords:")
countKeyword(resGO_BP_auto, cluster = "female-biased", 
             keywords = cellCycle)
print("Female-biased genes - development keywords:")
countKeyword(resGO_BP_auto, cluster = "female-biased", 
             keywords = development)
print("Female-biased genes - rna keywords:")
countKeyword(resGO_BP_auto, cluster = "female-biased", 
             keywords = rna)
print("Female-biased genes - Sensory percetion keywords:")
countKeyword(resGO_BP_auto, cluster = "female-biased", 
             keywords = sensoryPerception)
# male-biased genes & terms related to mitochondria and autophagy & terms related to synapse
print("Male-biased genes - mitochodria keywords:")
countKeyword(resGO_BP_auto, cluster = "male-biased", 
             keywords = mito)
print("Male-biased genes - neuron/synapse keywords:")
countKeyword(resGO_BP_auto, cluster = "male-biased", 
             keywords = neuro)

## Pseudotime-DE genes
# upregulated genes & terms related to mitochondria and autophagy & terms related to synapse
print("Up-regulated genes - mitochodria keywords:")
countKeyword(resGO_BP_auto, cluster = "up-regulated", 
             keywords = mito)
print("Up-regulated genes - neuron/synapse keywords:")
countKeyword(resGO_BP_auto, cluster = "up-regulated", 
             keywords = neuro)
# downregulated genes & terms related to cell cycle/division/proliferation & terms related to development
print("Down-regulated genes - cell cycle keywords:")
countKeyword(resGO_BP_auto, cluster = "down-regulated", 
             keywords = cellCycle)
print("Down-regulated genes - development keywords:")
countKeyword(resGO_BP_auto, cluster = "down-regulated", 
             keywords = development)
print("Down-regulated genes - rna keywords:")
countKeyword(resGO_BP_auto, cluster = "down-regulated", 
             keywords = rna)
print("Down-regulated genes - Sensory percetion keywords:")
countKeyword(resGO_BP_auto, cluster = "down-biased", 
             keywords = sensoryPerception)
```

```{r plot_GOcompare_BP_auto, echo=FALSE}
p <- plot_resGO(resGO_BP_auto)
p
ggsave("dotplot_BP_sexDE_pseudotimeDE_auto.png", plot = p, 
       path = "images/5-postprocessing/3-GO/prenatal/",
       width = 22, height = 20, units = "cm")
```

Save the results in file.

```{r save_GOcompare_BP_auto, echo=FALSE}
write.table(resGO_BP_auto@compareClusterResult,
            "data/5-postprocessing/3-GO/prenatal/GOcompare_BP_auto.txt", 
            row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")
```

### Analysis for CC

```{r GOanalysis_CC_auto, echo=FALSE}
resGO_CC_auto <- compare_GOenrich(data1 = de_results_sex_auto, data2 = de_results_pseudotime_auto, universe = universe, ont = "CC",
                                  data1Names = c("female-biased", "male-biased"), data2Names = c("up-regulated", "down-regulated"),
                                  qvalueThreshold = qvalueThreshold, logFCThreshold = logFCThreshold, simplify = FALSE)
```

Number of significant GO terms for each gene list.

```{r nbr_signif_CC_auto, echo=FALSE}
resGO_CC_auto@compareClusterResult %>% group_by(Cluster) %>% count()
```

Count number of GO terms with specific keywords.

```{r count_keywords_CC_auto, echo=FALSE}
## Sex-DE genes
# female-biased genes & terms related to cell cycle/division/proliferation & terms related to development
print("Female-biased genes - cell cycle keywords:")
countKeyword(resGO_CC_auto, cluster = "female-biased", 
             keywords = cellCycle)
print("Female-biased genes - development keywords:")
countKeyword(resGO_CC_auto, cluster = "female-biased", 
             keywords = development)
print("Female-biased genes - rna keywords:")
countKeyword(resGO_CC_auto, cluster = "female-biased", 
             keywords = rna)
print("Female-biased genes - Sensory percetion keywords:")
countKeyword(resGO_CC_auto, cluster = "female-biased", 
             keywords = sensoryPerception)
# male-biased genes & terms related to mitochondria and autophagy & terms related to synapse
print("Male-biased genes - mitochodria keywords:")
countKeyword(resGO_CC_auto, cluster = "male-biased", 
             keywords = mito)
print("Male-biased genes - neuron/synapse keywords:")
countKeyword(resGO_CC_auto, cluster = "male-biased", 
             keywords = neuro)

## Pseudotime-DE genes
# upregulated genes & terms related to mitochondria and autophagy & terms related to synapse
print("Up-regulated genes - mitochodria keywords:")
countKeyword(resGO_CC_auto, cluster = "up-regulated", 
             keywords = mito)
print("Up-regulated genes - neuron/synapse keywords:")
countKeyword(resGO_CC_auto, cluster = "up-regulated", 
             keywords = neuro)
# downregulated genes & terms related to cell cycle/division/proliferation & terms related to development
print("Down-regulated genes - cell cycle keywords:")
countKeyword(resGO_CC_auto, cluster = "down-regulated", 
             keywords = cellCycle)
print("Down-regulated genes - development keywords:")
countKeyword(resGO_CC_auto, cluster = "down-regulated", 
             keywords = development)
print("Down-regulated genes - rna keywords:")
countKeyword(resGO_CC_auto, cluster = "down-regulated", 
             keywords = rna)
print("Down-regulated genes - Sensory percetion keywords:")
countKeyword(resGO_CC_auto, cluster = "down-biased", 
             keywords = sensoryPerception)
```

```{r plot_GOcompare_CC_auto, echo=FALSE}
p <- plot_resGO(resGO_CC_auto)
p
ggsave("dotplot_CC_sexDE_pseudotimeDE_auto.png", plot = p, 
       path = "images/5-postprocessing/3-GO/prenatal/",
       width = 22, height = 20, units = "cm")
```

Save the results in file.

```{r save_GOcompare_CC_auto, echo=FALSE}
write.table(resGO_CC_auto@compareClusterResult,
            "data/5-postprocessing/3-GO/prenatal/GOcompare_CC_auto.txt", 
            row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")
```


### Analysis for MF

```{r GOanalysis_C´MF_auto, echo=FALSE}
resGO_MF_auto <- compare_GOenrich(data1 = de_results_sex_auto, data2 = de_results_pseudotime_auto, universe = universe, ont = "MF",
                                  data1Names = c("female-biased", "male-biased"), data2Names = c("up-regulated", "down-regulated"),
                                  qvalueThreshold = qvalueThreshold, logFCThreshold = logFCThreshold, simplify = FALSE)
```

Number of significant GO terms for each gene list.

```{r nbr_signif_MF_auto, echo=FALSE}
resGO_MF_auto@compareClusterResult %>% group_by(Cluster) %>% count()
```

Count number of GO terms with specific keywords.

```{r count_keywords_MF_auto, echo=FALSE}
## Sex-DE genes
# female-biased genes & terms related to cell cycle/division/proliferation & terms related to development
print("Female-biased genes - cell cycle keywords:")
countKeyword(resGO_MF_auto, cluster = "female-biased", 
             keywords = cellCycle)
print("Female-biased genes - development keywords:")
countKeyword(resGO_MF_auto, cluster = "female-biased", 
             keywords = development)
print("Female-biased genes - rna keywords:")
countKeyword(resGO_MF_auto, cluster = "female-biased", 
             keywords = rna)
print("Female-biased genes - Sensory percetion keywords:")
countKeyword(resGO_MF_auto, cluster = "female-biased", 
             keywords = sensoryPerception)
# male-biased genes & terms related to mitochondria and autophagy & terms related to synapse
print("Male-biased genes - mitochodria keywords:")
countKeyword(resGO_MF_auto, cluster = "male-biased", 
             keywords = mito)
print("Male-biased genes - neuron/synapse keywords:")
countKeyword(resGO_MF_auto, cluster = "male-biased", 
             keywords = neuro)

## Pseudotime-DE genes
# upregulated genes & terms related to mitochondria and autophagy & terms related to synapse
print("Up-regulated genes - mitochodria keywords:")
countKeyword(resGO_MF_auto, cluster = "up-regulated", 
             keywords = mito)
print("Up-regulated genes - neuron/synapse keywords:")
countKeyword(resGO_MF_auto, cluster = "up-regulated", 
             keywords = neuro)
# downregulated genes & terms related to cell cycle/division/proliferation & terms related to development
print("Down-regulated genes - cell cycle keywords:")
countKeyword(resGO_MF_auto, cluster = "down-regulated", 
             keywords = cellCycle)
print("Down-regulated genes - development keywords:")
countKeyword(resGO_MF_auto, cluster = "down-regulated", 
             keywords = development)
print("Down-regulated genes - rna keywords:")
countKeyword(resGO_MF_auto, cluster = "down-regulated", 
             keywords = rna)
print("Down-regulated genes - Sensory percetion keywords:")
countKeyword(resGO_MF_auto, cluster = "down-biased", 
             keywords = sensoryPerception)
```

```{r plot_GOcompare_MF_auto, echo=FALSE}
p <- plot_resGO(resGO_MF_auto)
p
ggsave("dotplot_MF_sexDE_pseudotimeDE_auto.png", plot = p, 
       path = "images/5-postprocessing/3-GO/prenatal/",
       width = 22, height = 20, units = "cm")
```

Save the results in file.

```{r save_GOcompare_MF_auto, echo=FALSE}
write.table(resGO_MF_auto@compareClusterResult,
            "data/5-postprocessing/3-GO/prenatal/GOcompare_MF_auto.txt", 
            row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")
```


# Session Info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
