---
title: "Sample repartition"
author: "Clara Benoit-Pilven"
date: "`r Sys.Date()`"
output: html_document
---

```{r load_library, echo=FALSE}
suppressPackageStartupMessages( library(tidyverse) )
```

```{r setup, echo=FALSE}
knitr::opts_knit$set(root.dir = '~/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/')
```

```{r load_utils, echo=FALSE}
source("scripts/analyses/1-preprocessing/utils.R")
```

# Load data and format it

```{r load_data, echo=FALSE}
phenotypes <- read.table("data/1-preprocessing/phenotypes/phenotypes_forebrain.txt", header = TRUE, sep = "\t")
phenotypes$newBrainRegion <- as.factor(phenotypes$newBrainRegion)
phenotypes$newDevStage <- as.factor(phenotypes$newDevStage)
```


# Repartition of the samples from HDBR

## Some numbers

In the selected samples, the number of samples is:

```{r number_samples_perSex_forebrain, echo=FALSE}
summary(as.factor(phenotypes$assigned_sex))
```

136 male samples, and 130 female samples.

In the selected samples, the number of individuals is:

```{r number_individuals_perSex_forebrain, echo=FALSE}
length(unique(phenotypes$ind[which(phenotypes$assigned_sex == 1)]))
length(unique(phenotypes$ind[which(phenotypes$assigned_sex == 2)]))
```

35 male individuals and 37 female individuals

Mean number of samples per individuals.

```{r mean_nbrSamplesPerInd, echo=FALSE}
nbrSamplePerInd <- phenotypes %>% group_by(ind) %>% summarise(nbrSample = n())
mean(nbrSamplePerInd$nbrSample)
```


## Repartition plots 

First, plot the repartition of the samples per sex for the developmental stage covariate.

```{r plot_sex_devStage_forebrain, echo=FALSE}
p <- PlotSexRepartition(phenotypes$assigned_sex, phenotypes$newDevStage, "Developmental stage")
p
# save plot
namePlot <- 'images/1-preprocessing/repartition_samples/samples_new_dev_stage_forebrain.png'
ggsave(namePlot, plot = p, width = 15, height = 10, units = "cm")
```

Then, plot the repartition of the samples per sex for the brain region covariate.

```{r plot_sex_brainRegion_forebrain, echo=FALSE}
p <- PlotSexRepartition(phenotypes$assigned_sex, phenotypes$newBrainRegion, "Brain region")
p
# save plot
namePlot <- 'images/1-preprocessing/repartition_samples/samples_brain_region_forebrain.png'
ggsave(namePlot, plot = p, width = 10, height = 8, units = "cm")
```

Finally, plot the repartition of the samples per sex for each association brain region - developmental stage.

```{r plot_sex_brainRegion_devStage_forebrain, echo=FALSE}
p <- PlotSexRepartition(phenotypes$assigned_sex, phenotypes$newDevStage, "Developmental stage", covariate2 = phenotypes$newBrainRegion, covariate2Name = "Brain region")
p
# save plot
namePlot <- 'images/1-preprocessing/repartition_samples/samples_newDevStage_and_brainRegion_forebrain.png'
ggsave(namePlot, plot = p, width = 10, height = 12, units = "cm")
```


# Session Info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
