---
title: "Prepare MAGMA data"
author: "Clara Benoit-Pilven"
date: "`r Sys.Date()`"
output: html_document
---

```{r load_library, echo=FALSE}
suppressPackageStartupMessages( library(readxl) )
suppressPackageStartupMessages( library(tidyverse) )
```

```{r, setup, include=FALSE, echo=FALSE}
knitr::opts_knit$set(root.dir = '~/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/')
```

Parameters:

```{r parameters, echo=FALSE}
qvalue_threshold <- 0.01
logFC_threshold <- 0
```


# Load all data needed

## HDBR data.

Load HDBR sex-DE data.

```{r load_data_HDBR_sexDE, echo=FALSE}
hdbr_sexDE_results <- read.table("data/3-DE/VoomDream_topTable_forebrain_pseudotime_sva_ind_interaction.txt",
                         header = TRUE, check.names=FALSE)
# remove genes used for sex assignment
genes2remove <- c("XIST", as.character(hdbr_sexDE_results$gene_name[which(hdbr_sexDE_results$chr == "chrY" & hdbr_sexDE_results$gene_type == "protein_coding")]))
hdbr_sexDE_results <- hdbr_sexDE_results[which(!hdbr_sexDE_results$gene_name %in% genes2remove),]
# format
hdbr_sexDE_results$ensemblID <- sapply(strsplit(as.character(hdbr_sexDE_results$gene_id), "\\."), function(x){x[1]})
```

Load HDBR pseudotime-DE data.

```{r load_data_HDBR_pseudotimeDE, echo=FALSE}
hdbr_pseudotimeDE_results <- read.table("data/3-DE/VoomDream_topTable_forebrain_pseudotime_sva_ind_interaction_pseudotime.txt",
                         header = TRUE, check.names=FALSE)
# remove genes used to infer pseudotime
load("data/2-pseudotime/forebrain_pseudotime_analysis.RData")
genes2remove <- rownames(selectedCounts)
hdbr_pseudotimeDE_results <- hdbr_pseudotimeDE_results[which(!hdbr_pseudotimeDE_results$gene_name %in% genes2remove),]
# format
hdbr_pseudotimeDE_results$ensemblID <- sapply(strsplit(as.character(hdbr_pseudotimeDE_results$gene_id), "\\."), function(x){x[1]})
```


## GTEx data.

Load GTEx sex-DE data.

```{r load_data_GTEx_sexDE, echo=FALSE}
gtex_sexDE_results <- read.table("data/4-GTEx/1-GTEx_analysis/brain_forebrain_v8_VoomDream_topTable_with_sva_ind_pseudotime_interaction.txt",
                         header = TRUE, check.names=FALSE)
# remove genes used for sex assignment
genes2remove <- c("XIST", as.character(gtex_sexDE_results$gene_name[which(gtex_sexDE_results$chr == "chrY" & gtex_sexDE_results$gene_type == "protein_coding")]))
gtex_sexDE_results <- gtex_sexDE_results[which(!gtex_sexDE_results$gene_name %in% genes2remove),]
# format
gtex_sexDE_results$ensemblID <- sapply(strsplit(as.character(gtex_sexDE_results$gene_id), "\\."), function(x){x[1]})
```

Load GTEx pseudotime-DE data.

```{r load_data_GTEx_pseudotimeDE, echo=FALSE}
gtex_pseudotimeDE_results <- read.table("data/4-GTEx/1-GTEx_analysis/brain_forebrain_v8_VoomDream_topTable_with_sva_ind_pseudotime_interaction_pseudotime.txt",
                                        header = TRUE, check.names=FALSE)
# remove genes used to infer pseudotime
load("data/4-GTEx/1-GTEx_analysis/GTEx_v8_forebrain_pseudotime_analysis.RData")
genes2remove <- rownames(selectedCounts)
gtex_pseudotimeDE_results <- gtex_pseudotimeDE_results[which(!gtex_pseudotimeDE_results$gene_name %in% genes2remove),]
# format
gtex_pseudotimeDE_results$ensemblID <- sapply(strsplit(as.character(gtex_pseudotimeDE_results$gene_id), "\\."), function(x){x[1]})
```


## Bayesian results.

```{r load_bayes_results, echo=FALSE}
bayesResults <- read.table("data/4-GTEx/2-comparison/lineModel/lineModel_prenatalVSadult_6models_withAnnotation.txt", 
                           header = TRUE, sep = "\t")
# Keep only autosomal genes
bayesResults <- bayesResults[which(bayesResults$annot == "Autosome"), ]
```



# Prepare the gene lists for the MAGMA analyses and save them

## Functions

```{r function_select_DEgenes, echo=FALSE}
selectDEgenes <- function(de_results, qvalue_threshold = 0.01, logFC_threshold = 0){
  # keep only autosomal genes
  de_results <- de_results[which(!de_results$chr %in% c("chrX", "chrY", "chrM")),]
  # Create the 3 genes lists
  DEgenes <- de_results$ensemblID[which(de_results$qvalue < qvalue_threshold & 
                                          abs(de_results$logFC) > logFC_threshold)]
  upGenes <- de_results$ensemblID[which(de_results$qvalue < qvalue_threshold & 
                                          de_results$logFC > logFC_threshold)]
  downGenes <- de_results$ensemblID[which(de_results$qvalue < qvalue_threshold & 
                                            de_results$logFC < logFC_threshold)]
  return(list("DEgenes" = DEgenes, "upGenes" = upGenes, "downGenes" = downGenes))
}
```

```{r function_select_bayesRes, echo=FALSE}
selectBayesRes <- function(results, data2select, logFC_threshold = 0){
  selectedSpecificity <- ""
  if (data2select == "shared"){
    selectedSpecificity <- "Shared"
  } else if (data2select == "prenatal_specific"){
    selectedSpecificity <- "Prenatal"
  }
  # Get gene in the selected category
  results <- results[which(results$specificity == selectedSpecificity),]
  # Create the 3 gene lists
  DEgenes <- results$ensemblID
  upGenes <- results$ensemblID[which(results$logFC_adult > logFC_threshold)]
  downGenes <- results$ensemblID[which(results$logFC_adult < logFC_threshold)]
  return(list("DEgenes" = DEgenes, "upGenes" = upGenes, "downGenes" = downGenes))
}
```

```{r function_save_results, echo=FALSE}
save_data_magma <- function(data, suffix, outputDir){
  fullFileName <- paste0(paste(paste0(outputDir, "MAGMA"), suffix, sep = "_"), ".txt")
  write.table(data, file = fullFileName, 
              col.names = FALSE, row.names = FALSE, quote = FALSE)
  return(1)
}
```


## Create gene lists to be used as MAGMA input

Save data from the DE analysis.

```{r prepareNsave_data, echo=FALSE}
listDEdata <- list(hdbr_sexDE_results, hdbr_pseudotimeDE_results,
                   gtex_sexDE_results, gtex_pseudotimeDE_results)
listTypeData <- c("sexDE", "pseudotimeDE", "sexDE", "pseudotimeDE")
listDataset <- c("hdbr", "hdbr", "gtex", "gtex")
i <- 1
for (DEdata in listDEdata){
  geneLists <- selectDEgenes(DEdata)
  j <- 1
  if (listTypeData[i] == "sexDE"){
    directionEffect <- c("", "_femaleBiased", "_maleBiased")
  } else if (listTypeData[i] == "pseudotimeDE"){
    directionEffect <- c("", "_upRegulated", "_downRegulated")
  }
  for (geneList in geneLists){
    suffix <- paste0(paste(listDataset[i], listTypeData[i], sep = "_"), directionEffect[j])
    print(suffix)
    save_data_magma(geneList, suffix, outputDir = "data/5-postprocessing/5-MAGMA/inputLists/")
    j <- j + 1
  }
  i <- i + 1
}
```

Save data from the bayesian analysis.

```{r prepareNsave_data_bayes, echo=FALSE}
analyses <- c("shared", "prenatal_specific")
directionEffect <- c("", "_femaleBiased", "_maleBiased")
for (currentAnalysis in analyses){
  bayesLists <- selectBayesRes(bayesResults, currentAnalysis)
  i <- 1
  for (geneList in bayesLists){
    suffix <- paste0(paste("sexDE_bayes", currentAnalysis, sep = "_"), directionEffect[i])
    print(suffix)
    save_data_magma(geneList, suffix, outputDir = "data/5-postprocessing/5-MAGMA/inputLists/")
    i <- i + 1
  }
}
```

Create the input file for the MAGMA gene level analysis.

```{r create_magma_input, echo=FALSE}
MAGMA_input <- list()
# sexDE and pseudotimeDE genes
listDEdata <- list(hdbr_sexDE_results, hdbr_pseudotimeDE_results,
                   gtex_sexDE_results, gtex_pseudotimeDE_results)
listTypeData <- c("sexDE", "pseudotimeDE", "sexDE", "pseudotimeDE")
listDataset <- c("hdbr", "hdbr", "gtex", "gtex")
i <- 1
for (DEdata in listDEdata){
  geneLists <- selectDEgenes(DEdata)
  j <- 1
  if (listTypeData[i] == "sexDE"){
    directionEffect <- c("", "_femaleBiased", "_maleBiased")
  } else if (listTypeData[i] == "pseudotimeDE"){
    directionEffect <- c("", "_upRegulated", "_downRegulated")
  }
  for (geneList in geneLists){
    suffix <- paste0(paste(listDataset[i], listTypeData[i], sep = "_"), directionEffect[j])
    print(suffix)
    MAGMA_input[[suffix]] <- c(suffix, geneList)
    j <- j + 1
  }
  i <- i + 1
}
# bayesian analysis
analyses <- c("shared", "prenatal_specific")
directionEffect <- c("", "_femaleBiased", "_maleBiased")
for (currentAnalysis in analyses){
  bayesLists <- selectBayesRes(bayesResults, currentAnalysis)
  i <- 1
  for (geneList in bayesLists){
    suffix <- paste0(currentAnalysis, directionEffect[i])
    print(suffix)
    MAGMA_input[[suffix]] <- c(suffix, geneList)
    i <- i + 1
  }
}
# write the file
file_conn <- file("data/5-postprocessing/5-MAGMA/inputLists/MAGMA_input.txt")
writeLines(unlist(lapply(MAGMA_input, paste, collapse=" ")), file_conn)
close(file_conn)
```


# Session info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

