---
title: "Compare GTEx sex-DE results with pseudotime or age as covariate"
author: "Clara Benoit-Pilven"
date: "`r Sys.Date()`"
output: html_document
---

```{r load_library, echo=FALSE}
suppressPackageStartupMessages( library(tidyverse) )
suppressPackageStartupMessages( library(edgeR) )
suppressPackageStartupMessages( library(limma) )
suppressPackageStartupMessages( library(qvalue) )
suppressPackageStartupMessages( library(ggalluvial) )
suppressPackageStartupMessages( library(gridExtra) )
suppressPackageStartupMessages( library(fmsb) )
suppressPackageStartupMessages( library(ComplexHeatmap))
suppressPackageStartupMessages( library(ggpubr) )
```

```{r, setup, include=FALSE, echo=FALSE}
knitr::opts_knit$set(root.dir = '~/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/')
set.seed(1234)
```

Thresholds:
```{r thresholds, echo=FALSE}
qvalue_threshold <- 0.01
```

# Load data and functions

## Load and prepare sex-DE genes with pseudotime covariate

```{r load_sexDE_pseudo, echo=FALSE}
results_pseudo <- read.table("data/4-GTEx/1-GTEx_analysis/brain_forebrain_v8_VoomDream_topTable_with_sva_ind_pseudotime_interaction.txt",
                           header = TRUE)
```

Prepare the data by removing the genes used to define the sex of the samples.
```{r prepare_sexDE_pseudo, echo=FALSE}
genes2remove <- c("XIST", as.character(results_pseudo$gene_name[which(results_pseudo$chr == "chrY" & results_pseudo$gene_type == "protein_coding")]))
results_pseudo <- results_pseudo[which(!results_pseudo$gene_name %in% genes2remove),]
results_pseudo_auto <- results_pseudo[which(!results_pseudo$chr %in% c("chrY", "chrX", "chrM")), ]
```


## Load and prepare sex-DE genes with age covariate

```{r load_sexDE_age, echo=FALSE}
results_age <- read.table("data/4-GTEx/1-GTEx_analysis/brain_forebrain_v8_VoomDream_topTable_with_sva_ind_age_interaction.txt",
                          header = TRUE)
```

Prepare the data by removing the genes used to define the sex of the samples.
```{r prepare_sexDE_age, echo=FALSE}
genes2remove <- c("XIST", as.character(results_age$gene_name[which(results_age$chr == "chrY" & results_age$gene_type == "protein_coding")]))
results_age <- results_age[which(!results_age$gene_name %in% genes2remove),]
results_age_auto <- results_age[which(!results_age$chr %in% c("chrY", "chrX", "chrM")), ]
```


# Comparison of the sex-DE genes

We compare the sex-DE genes obtained in the GTEx data with either the pseudotime as a covariate or the age instead.

## Summary

We get the number of sex-DE genes in each analysis.

```{r get_stats_pseudo, echo=FALSE}
nbr_signif_pseudo <- sum(results_pseudo$qvalue < qvalue_threshold)
nbr_signif_pseudo
```

```{r get_stats_age, echo=FALSE}
nbr_signif_age <- sum(results_age$qvalue < qvalue_threshold)
nbr_signif_age
```


## Overlap

Get the number of significant genes overlapping between the 2 datasets.

```{r permut_test_overlap_function, echo=FALSE}
overlap_signif <- function(data1, data2, columnMerge = "ensemblID", qvalueThreshold = 0.01, nbrRandom = 1000){
  # merge the 2 datasets
  de_results_merged <- merge(data1, data2, by = columnMerge, suffixes = c(".data1", ".data2"))
  # get the number of significant genes in each data
  nbr_signif_data1 <- length(de_results_merged[which(de_results_merged$qvalue.data1 < qvalueThreshold), columnMerge])
  nbr_signif_data2 <- length(de_results_merged[which(de_results_merged$qvalue.data2 < qvalueThreshold), columnMerge])
  print(sprintf("Number of genes significant in data1 = %s", nbr_signif_data1))
  print(sprintf("Number of genes significant in data2 = %s", nbr_signif_data2))
  # keep only the genes/transcripts that are significant in the 2 datasets
  de_results_merged_filtered <- de_results_merged[which(de_results_merged$qvalue.data1 < qvalueThreshold & de_results_merged$qvalue.data2 < qvalueThreshold), ]
  nbr_common_signif <- dim(de_results_merged_filtered)[1]
  print(sprintf("Number of genes significant in both dataset = %s", nbr_common_signif))
  print(sprintf("Percentage of significant genes from data1 also significant in data2 = %s", nbr_common_signif/nbr_signif_data1*100))
  # Compare this number to the number of overlaping genes if we take randomly 2 lists of the same size (nbr_sgnif_data1 & nbr_sgnif_data2)
  nbr_signif_values <- c()
  # to do nbrRandom times
  for (i in 1:nbrRandom){
    # randomly select "nbr_sgnif_data1" genes/transcripts in the data1
    selection_data1 <- sample(de_results_merged[, columnMerge], nbr_signif_data1)
    # randomly select "nbr_sgnif_data2" genes/transcripts in the data2
    selection_data2 <- sample(de_results_merged[, columnMerge], nbr_signif_data2)
    nbr_signif_th <- sum(selection_data1 %in% selection_data2)
    nbr_signif_values[i] <- nbr_signif_th
  }
  nbr_signif_values <- as.data.frame(nbr_signif_values)
  # plot the distribution of the number of overlaping genes obtained and the real value
  p <- ggplot(nbr_signif_values, aes(nbr_signif_values))
  p <- p + geom_histogram(binwidth=1, boundary = 0, fill="grey", color="black")
  p <- p + theme_bw() + xlab("Number of overlaping genes") + ylab("Frequency")
  p <- p + geom_vline(xintercept = nbr_common_signif, color = "red")
  return(p)
}
```

```{r permut_test_overlap_age, echo=FALSE}
p <- overlap_signif(results_age, results_pseudo, qvalueThreshold = qvalue_threshold, columnMerge = "gene_id")
p
```


## Replication rate

We compute the replication rate (pi1).

```{r function_pi1_score, echo=FALSE}
pi1_score <- function(data1, data2, columnMerge = "ensemblID"){
  # compute pi1 for second dataset compared to the first dataset
  selected_pvalue <- data2$P.Value[which(pull(data2, columnMerge) %in% pull(data1, columnMerge))]
  if (berryFunctions::is.error(qvalue::pi0est(p = selected_pvalue, lambda = 0.5, pi0.method = "smoother")) ) {
    pi1 <- qvalue::qvalue(p = selected_pvalue, pi0 = 1)$pi0
  }else{
    pi1 <- 1-qvalue::pi0est(p = selected_pvalue, lambda = 0.5, pi0.method = "smoother")$pi0
  }
  
  # continue only if the pi1 value was computed
  if (is.numeric(pi1)){
    return(pi1)
  } else{
    return("No pi1 analysis!")
  }
}
```

```{r function_pi1_normalized, echo=FALSE}
pi1_normalized <- function(data1, data2, columnMerge = "ensemblID", qvalue_threshold = 0.01, logFC_threshold = 0, verbose = TRUE){
  # get the significant genes in the first dataset
  data1_signif <- data1[which(data1$qvalue < qvalue_threshold & abs(data1$logFC) > logFC_threshold), ]
  
  # pi1 replication rate for the significant genes
  pi1_signif <- pi1_score(data1_signif, data2, columnMerge = columnMerge)
  if (verbose){
    print(sprintf("pi1 for signif genes = %s", pi1_signif))
  }
  # pi1 replication rate for all genes
  pi1_all <- pi1_score(data1, data2, columnMerge = columnMerge)
  if (verbose){
    print(sprintf("pi1 for all genes = %s", pi1_all))
  }
  
  # return the pi1 for significant genes normalized by the pi1 for all genes
  return(pi1_signif/pi1_all)
}
```

```{r pi1_score, echo=FALSE}
print("Replication of GTEx pseudotime sex-DE results in GTEx age sex-DE analysis:")
pi1_normalized(results_pseudo, results_age, qvalue_threshold = qvalue_threshold, columnMerge = "gene_id")
print("Replication of GTEx age sex-DE results in results_pseudo sex-DE analysis:")
pi1_normalized(results_age, results_pseudo, qvalue_threshold = qvalue_threshold, columnMerge = "gene_id")
```


## Effect size correlation 

Scatter plot of the logFC between the 2 analyses with information about the correlation (r and p-value of cor.test).

```{r function_scatter_plot_logFC, echo=FALSE}
scatter_logFC <- function(data1, data2, nameData1, nameData2, pointColor, columnMerge = "ensemblID"){
  # merge the 2 dataset
  de_results_merged <- merge(data1, data2, by = columnMerge, suffixes = c(".data1", ".data2"))
  # test correlation
  res <- cor.test(de_results_merged$logFC.data1, de_results_merged$logFC.data2)
  if (res$p.value == 0){
    annotation_text <- sprintf("r = %.3f\nP < 2.2e-16", res$estimate)
  } else{
    annotation_text <- sprintf("r = %.3f\nP = %1.2e", res$estimate, res$p.value)
  }
  yrng <- range(de_results_merged$logFC.data1)
  xrng <- range(de_results_merged$logFC.data2)
  # plot
  p <- ggplot(de_results_merged, aes(x = logFC.data1, y = logFC.data2))
  p <- p + geom_abline(intercept = 0, slope = 1, col = "black")
  p <- p + geom_hline(yintercept = 0, color = "grey")
  p <- p + geom_vline(xintercept = 0, color = "grey")
  p <- p + geom_point(color=pointColor, alpha = 0.3)
  p <- p + theme_bw() + labs(title = "")
  p <- p + theme(panel.grid = element_blank(), 
                 plot.title = element_text(hjust = 0.5),
                 plot.subtitle = element_text(hjust = 0.5))
  p <- p + xlab(paste0("logFC (", nameData1, ")")) + ylab(paste0("logFC (", nameData2, ")"))
  p <- p + annotate("text", xrng[1], yrng[2], hjust = 0, vjust = 1, 
                    label = annotation_text)
  return(p)
}
```

Correlation plot only of GTEx pseudo sex-DE genes.

```{r corr_pseudo_sexDE_age, echo=FALSE}
p <- scatter_logFC(results_pseudo[which(results_pseudo$qvalue < qvalue_threshold),], results_age, 
                   "GTEx pseudotime", "GTEx age", 
                   pointColor = "black", columnMerge = "gene_id")
p
```

Correlation plot only of autosomal GTEx pseudo sex-DE genes.

```{r corr_pseudo_sexDE_auto_age, echo=FALSE}
p <- scatter_logFC(results_pseudo_auto[which(results_pseudo_auto$qvalue < qvalue_threshold),], results_age, 
                   "GTEx pseudotime", "GTEx age", 
                   pointColor = "black", columnMerge = "gene_id")
p
```

Correlation plot only for GTEx age sex-DE genes.

```{r corr_age_sexDE_pseudo, echo=FALSE}
p <- scatter_logFC(results_age[which(results_age$qvalue < qvalue_threshold),], results_pseudo, 
                   "GTEx age", "GTEx pseudotime", 
                   pointColor = "black", columnMerge = "gene_id")
p
```

Correlation plot only for O'Brien autosomal sex-DE genes.

```{r corr_age_sexDE_pseudo_auto, echo=FALSE}
p <- scatter_logFC(results_age_auto[which(results_age_auto$qvalue < qvalue_threshold),], results_pseudo, "GTEx age", "GTEx pseudotime", 
                   pointColor = "black", columnMerge = "gene_id")
p
```



## Effect size consistency

Compute the consistency of log fold change.

```{r function_consistency, echo=FALSE}
consistency_test <- function(data1, data2, columnMerge = "ensemblID", qvalueThreshold = 0.01){
  # merge the 2 datasets
  de_results_merged <- merge(data1, data2, by = columnMerge, suffixes = c(".data1", ".data2"))
  # keep only the genes/transcripts that are significant in the first data
  de_results_merged_filtered <- de_results_merged[which(de_results_merged$qvalue.data1 < qvalueThreshold), ]
  print(sprintf("Number of genes compared = %s", dim(de_results_merged_filtered)[1]))
  # number of consistent genes/transcripts
  nbr_consistent <- sum(sign(de_results_merged_filtered$logFC.data1) == sign(de_results_merged_filtered$logFC.data2))
  consistency <- nbr_consistent/length(de_results_merged_filtered[, columnMerge])
  # compare this consistency score with the consistency of list of random genes
  cons_values <- rep(NA, 1000)
  # to do 1000 times
  for (i in 1:1000){
    # randomly select the same number of genes/transcripts in the data2 to test the overlap with the data1 signif genes
    selection <- sample(de_results_merged[, columnMerge], length(de_results_merged_filtered[, columnMerge]))
    nbr_consistent_th <- sum(sign(de_results_merged$logFC.data2[which(de_results_merged[, columnMerge] %in% selection)]) == sign(de_results_merged$logFC.data1[which(de_results_merged[, columnMerge] %in% selection)]))
    cons_values[i] <- nbr_consistent_th/length(de_results_merged_filtered[, columnMerge])
  }
  cons_values <- as.data.frame(cons_values)
  # plot the distribution of the consistency obtained and the real value
  p <- ggplot(cons_values, aes(cons_values))
  p <- p + geom_histogram(binwidth=0.02, boundary = 0, fill="grey", color="black")
  p <- p + theme_bw() + xlab("consistency") + ylab("Frequency")
  p <- p + scale_x_continuous(limits = c(0, 1))
  p <- p + geom_vline(xintercept = consistency, color = "red")
  print(p)
  # test this consistency with a permutation test
  nsmaller <- sum(cons_values >= consistency)
  permutation_test <- round((nsmaller+1)/(1000+1), 3)
  # test this consistency with a binomial test
  result_test <- binom.test(nbr_consistent, length(de_results_merged_filtered[, columnMerge]), p = 0.5)
  # print results
  print(sprintf("number of consistent gene = %s", nbr_consistent))
  print(sprintf("consistency = %s", consistency))
  print(sprintf("permutation test p-value = %s", permutation_test))
  print(sprintf("binomial test p-value = %s", result_test$p.value))
  return(consistency)
}
```

```{r consistency_score_pseudo_vs_age, echo=FALSE}
print("Consistency of GTEx pseudotime analysis in the results of GTEx age analysis:")
consistency_test(results_age, results_pseudo, qvalueThreshold = 0.01, columnMerge = "gene_id")
```

```{r consistency_score_age_vs_pseudo, echo=FALSE}
print("Consistency of GTEx age analysis in the results of GTEx pseudotime analysis:")
consistency_test(results_pseudo, results_age, qvalueThreshold = 0.01, columnMerge = "gene_id")
```



# Session info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```






