---
title: "Adult sex-DE genes properties"
author: "Clara Benoit-Pilven"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r load_library, echo=FALSE}
suppressPackageStartupMessages( library(tidyverse) )
suppressPackageStartupMessages( library(ggplot2) )
suppressPackageStartupMessages( library(ggpubr) )
suppressPackageStartupMessages( library(gghighlight) )
suppressPackageStartupMessages( library(ggrepel) )
suppressPackageStartupMessages( library(matrixStats) )
suppressPackageStartupMessages( library(readxl) )
suppressPackageStartupMessages( library(rtracklayer) )
```

```{r, setup, include=FALSE, echo=FALSE}
knitr::opts_knit$set(root.dir = '~/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/')
```

```{r param, echo=FALSE}
threshold_qvalue <- 0.01
```

# Load and prepare data

## Load and prepare adult sex-DE genes

```{r load_sexDE_data, echo=FALSE}
results <- read.table("data/4-GTEx/1-GTEx_analysis/brain_forebrain_v8_VoomDream_topTable_with_sva_ind_pseudotime_interaction.txt", header = TRUE)
results$ensemblID <- sapply(strsplit(as.character(results$gene_id), "\\."), function(x){x[1]})
genes2remove <- c("XIST", as.character(results$gene_name[which(results$chr == "chrY" & results$gene_type == "protein_coding")]))
results <- results[which(!results$gene_name %in% genes2remove),]
```

```{r prepare_sexDE_data, echo=FALSE}
results$sexDE <- ifelse(results$qvalue < 0.01, ifelse(results$logFC > 0, "Female", "Male"), "No")
results$sexDE <- factor(results$sexDE, levels = c("No", "Female", "Male"))
```

Select autosomal genes

```{r sexDE_auto, echo=FALSE}
results_auto <- results[which(!results$chr %in% c("chrX", "chrY", "chrM")), ]
```

## Prepare expression data

Create expression bins.

```{r create_expr_bin, echo=FALSE}
results$exprbin <- cut(results$AveExpr, 
                       quantile(results$AveExpr, seq(0,1,0.1)), 
                       labels=c(1:10), include.lowest=T)
```


## Load and prepare BrainSpan expression data

First, load the brainSpan data: normalized TPM values in the forebrain.

```{r load_brainspan_normalized_tpm, echo=FALSE}
load("data/5-postprocessing/1-GeneProperties/normalized_tpm_samples_forebrain.RData")
```

Formatting by variable (e.g. life stages or brain regions).
```{r format_byVariable, echo=FALSE}
format_byVariable <- function(data, phenotypes, variable){
  values <- levels(phenotypes[, variable])#$lifeStageFull
  result <- data.frame(Gene = data$Gene)
  for (curVal in values){
    # get the index of all the column corresponding to the current value
    idx <- which(phenotypes[, variable] == curVal)+1
    # compute the mean and sd of the columns
    if (length(idx) == 1){
      mean_expr <- data[, idx]
      sd_expr <- rep(0, length(data[, idx]))
    } else{
      mean_expr <- as.vector(rowMeans(data[, idx]))
      sd_expr <- rowSds(as.matrix(data[, idx]))
    }
    # write the results
    result[, paste(curVal, "mean", sep = "_")] <- mean_expr
    result[, paste(curVal, "sd", sep = "_")] <- sd_expr
  }
  return(result)
}
```
```{r format_brainspan_data, echo=FALSE}
normalized_tpm_byLifeStage <- format_byVariable(normalized_tpm, samples, "lifeStageFull2")
```

Merge this expression data with our data.

```{r merge_brainspan, echo=FALSE}
results_brainspan <- merge(results, normalized_tpm_byLifeStage, by.x = "gene_name", by.y = "Gene")
```


## Load and prepare constraint metric

```{r prepare_loeuf_score, echo=FALSE}
# load loeuf score
constraint_data <- read.table("data/5-postprocessing/1-GeneProperties/gnomad.v4.1.constraint_metrics.tsv", 
                              header = TRUE, sep = "\t")
constraint_data <- constraint_data[c("gene_id", "canonical", "mane_select", "lof.oe_ci.upper")]
# keep MANE transcript or canonical if no MANE transcript for a gene
constraint_data <- constraint_data[which((constraint_data$mane_select == "true" | constraint_data$canonical == "true") & startsWith(constraint_data$gene_id, "ENSG")), ]
constraint_data <- constraint_data[c("gene_id", "lof.oe_ci.upper")]
names(constraint_data) <- c("gene_id", "LOEUF")
# merge with results
results <- merge(results, constraint_data, by.x = "ensemblID", by.y = "gene_id", all.x = TRUE)
# create loeuf bins
results$LOEUF_bin <- cut(results$LOEUF, 
                     quantile(results$LOEUF, seq(0,1,0.1), na.rm = TRUE), 
                     labels=c(1:10), include.lowest=T)
```

## Load and prepare tissue-specificity metric (tau?)

```{r load_tau_data, echo=FALSE}
tau_data <- read.table("data/5-postprocessing/1-GeneProperties/Tau_gene_V8.csv",
                       header = TRUE, sep = ",")
tau_data$sumExpr <- rowSums(tau_data[,3:32])
tau_data$tau_modif <- ifelse(is.na(tau_data$tau) & tau_data$sumExpr == 0, -1, tau_data$tau)
# merge with results
results <- merge(results, tau_data, by.x = "ensemblID", by.y = "gene_id", all.x = TRUE)
```

## Load and prepare sexDE sharing across GTEx tissues

Load GTEx shared sex-DE genes.

```{r gtex_shared_sexDE, echo=FALSE}
load("data/5-postprocessing/1-GeneProperties/GTEx_tissues_sexDE_sharing_bayes_6models.RData")
```

Merge sex-DE results with sharing in GTEx tissues.

```{r gtexSharing_sexDE, echo=FALSE}
sharing_sexDE_annot$ensemblID <- sapply(strsplit(as.character(sharing_sexDE_annot$gene), "\\."),function(x){x[1]})
sharing_sexDE_annot$dir <- ifelse(sharing_sexDE_annot$sexDE_coherent == sharing_sexDE_annot$femaleBias, "female",
                                  ifelse(sharing_sexDE_annot$sexDE_coherent == sharing_sexDE_annot$maleBias, "male", "no"))
consistent_sexDE$ensemblID <- sapply(strsplit(as.character(consistent_sexDE$gene), "\\."),function(x){x[1]})
consistent_sexDE$dir <- ifelse(consistent_sexDE$sexDE_coherent == consistent_sexDE$femaleBias, "female",
                               ifelse(consistent_sexDE$sexDE_coherent == consistent_sexDE$maleBias, "male", "no"))
sharing_sexDE_annot_tmp <- sharing_sexDE_annot[, c(12,13,2,3,4,5,6,7,8)]
names(sharing_sexDE_annot_tmp) <- c("ensemblID", "dirGTEx", "nbrTissue_sexDE",
                                    "nbrTissue_femaleBias", "nbrTissue_maleBias",
                                    "nbrTissueAnalyzed", "sexDE_coherent",
                                    "percent_sexDE","percent_coherent")
sharing_sexDE_annot_tmp$dirGTEx <- ifelse(sharing_sexDE_annot$dir == "male", "Male",
                                          ifelse(sharing_sexDE_annot$dir == "female", "Female", NA))
overlap_gtexSharing_sexDE <- merge(results, sharing_sexDE_annot_tmp, by = "ensemblID")
```


# Plots

## SexDE vs expression

Plot the level of expression of sex-DE genes vs non sex-DE genes.

```{r violinplot_expr, echo=FALSE}
# function
data_summary <- function(x) {
   m <- mean(x)
   ymin <- m-sd(x)
   ymax <- m+sd(x)
   return(c(y=m,ymin=ymin,ymax=ymax))
}
p <- ggplot(results, aes(x = sexDE, y = AveExpr, fill = sexDE))
p <- p + geom_violin(alpha = 0.5, scale = "area", linewidth = 0.01, show.legend = FALSE) 
p <- p + stat_summary(fun.data=data_summary, size=0.1, show.legend = FALSE)
p <- p + theme_minimal() + xlab("sex-DE") + ylab("Mean expression")
p <- p + scale_fill_manual(values = c("Male" = "darkblue", "Female" = "darkred", "No" = "lightgrey"),
                           name = "sex-DE", )
p <- p + stat_compare_means(method = "wilcox.test",
                            label = "p.signif", hide.ns = TRUE, vjust = 0.7,
                            tip.length = 1, na.rm = TRUE, ref.group = "No")
p
# save plot
ggsave("adult_sexDE_vs_expr.png",
       path = "images/5-postprocessing/1-GeneProperties/adult/",
       width = 8, height = 8, units = "cm")
```


## SexDE per expression bin

Plot proportion of sex-DE gene for each expression bin coloring male and female-biased genes.

```{r plot_barplot_exprBin, echo=FALSE}
results %>% 
  dplyr::group_by(exprbin) %>%
  dplyr::count(sexDE) %>%
  dplyr::mutate(proportion = n/sum(n)) %>%
  ggplot(aes(exprbin, proportion, fill = sexDE)) + 
  geom_col(position = position_stack()) + 
  geom_text(aes(label = n), position = position_stack(vjust =0.5), colour = "white", size = 2) + 
  theme_minimal() + xlab("") + ylab("") +
  scale_fill_manual(values = c("Male" = "darkblue", "Female" = "darkred", "No" = "lightgrey"),
                           name = "sex-DE")
# save plot
ggsave("adult_sexDE_vs_exprBin.png",
       path = "images/5-postprocessing/1-GeneProperties/adult/",
       width = 12, height = 8, units = "cm")
```

## SexDE vs BrainSpan expression

```{r format_BrainSpan_expr, echo=FALSE}
results_brainspan2plot <-  results_brainspan[, c(1,22,24,26,28,30,32,34,36,38)] %>% pivot_longer(3:10, names_to = c("lifeStage", "mean"), names_sep = "_", values_to = "expr") 
results_brainspan2plot_grouped <- results_brainspan2plot[,-4] %>% group_by(sexDE, lifeStage) %>% summarise("meanExpr" = mean(expr), seExpr = sd(expr)/sqrt(length(expr)), sdExpr = sd(expr), pi2.5Expr = quantile(expr, 0.025), pi97.5Expr = quantile(expr, 0.975))
# order the lifeStage factor
results_brainspan2plot_grouped$lifeStage <- factor(results_brainspan2plot_grouped$lifeStage,
                                                   levels = c("Early prenatal", "Early mid-prenatal", "Late mid-prenatal", "Late prenatal", "Infancy", "Childhood", "Adolescence", "Adulthood"),
                                                   labels = c("Early\nprenatal", "Early\nmid-prenatal", "Late\nmid-prenatal", "Late\nprenatal", "Infancy", "Childhood", "Adolescence", "Adulthood"))
results_brainspan2plot$lifeStage <- factor(results_brainspan2plot$lifeStage,
                                           levels = c("Early prenatal", "Early mid-prenatal", "Late mid-prenatal", "Late prenatal", "Infancy", "Childhood", "Adolescence", "Adulthood"),
                                           labels = c("Early\nprenatal", "Early\nmid-prenatal", "Late\nmid-prenatal", "Late\nprenatal", "Infancy", "Childhood", "Adolescence", "Adulthood"))
```


```{r violinplot_BrainSpan_expr, echo=FALSE}
p <- ggplot(results_brainspan2plot, aes(x = sexDE, y = expr, fill = sexDE))
p <- p + geom_violin(alpha = 0.5, scale = "area", size = 0.01, show.legend = FALSE) 
p <- p + stat_summary(fun.data=data_summary, size=0.1, show.legend = FALSE)
p <- p + facet_grid(cols = vars(lifeStage))
p <- p + theme_minimal() + xlab("") + ylab("Mean expression")
p <- p + scale_fill_manual(values = c("Male" = "darkblue", "Female" = "darkred", "No" = "lightgrey"),
                           name = "sex-DE", )
p <- p + stat_compare_means(method = "wilcox.test",
                            label = "p.signif", hide.ns = TRUE, vjust = 0.7,
                            tip.length = 1, na.rm = TRUE, ref.group = "No")
p <- p + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
p
# save plot
ggsave("adult_sexDE_vs_BrainSpanExpr.png",
       path = "images/5-postprocessing/1-GeneProperties/adult/",
       width = 20, height = 8, units = "cm")
```

```{r violinplot_BrainSpan_expr_lifestage, echo=FALSE}
p <- ggplot(results_brainspan2plot_grouped, aes(x = lifeStage, y = meanExpr, ymin = pi2.5Expr, ymax = pi97.5Expr,
                                                fill = sexDE, color = sexDE, group = sexDE))
p <- p + geom_line(position = position_dodge(width = 0.2)) 
p <- p + geom_point(position = position_dodge(width = 0.2),  show.legend = FALSE) 
p <- p + geom_errorbar(position = position_dodge(width = 0.2), width = 0.2)
#p <- p + stat_summary(fun.data=data_summary, size=0.1, show.legend = FALSE, position = "nudge")
p <- p + theme_minimal() + xlab("Life stage") + ylab("Mean expression")
p <- p + scale_color_manual(values = c("Male" = "darkblue", "Female" = "darkred", "No" = "lightgrey"),
                           name = "sex-DE", )
p <- p + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
p
# save plot
ggsave("adult_sexDE_vs_BrainSpanExpr_lifeStage.png",
       path = "images/5-postprocessing/1-GeneProperties/adult/",
       width = 12, height = 8, units = "cm")
```

## SexDE vs constraint

Plot the LOEUF score of sex-DE genes vs non sex-DE genes.

```{r violinplot_constraint, echo=FALSE}
p <- ggplot(results[which(!is.na(results$LOEUF)),], aes(x = sexDE, y = LOEUF, fill = sexDE))
p <- p + geom_violin(alpha = 0.5, scale = "area", linewidth = 0.01, show.legend = FALSE) 
p <- p + stat_summary(fun.data=data_summary, size=0.1, show.legend = FALSE)
p <- p + theme_minimal() + xlab("sex-DE") + ylab("Constraint")
p <- p + scale_fill_manual(values = c("Male" = "darkblue", "Female" = "darkred", "No" = "lightgrey"),
                           name = "sex-DE", )
p <- p + stat_compare_means(method = "wilcox.test",
                            label = "p.signif", hide.ns = TRUE, vjust = 0.7,
                            tip.length = 1, na.rm = TRUE, ref.group = "No")
p
# save plot
ggsave("adult_sexDE_vs_constraint.png",
       path = "images/5-postprocessing/1-GeneProperties/adult/",
       width = 8, height = 8, units = "cm")
```


## SexDE per constraint bin

Plot proportion of sex-DE gene for each constraint bin coloring male and female-biased genes.

```{r plot_barplot_constraintBin, echo=FALSE}
results[which(!is.na(results$LOEUF)),] %>% 
  dplyr::group_by(LOEUF_bin) %>%
  dplyr::count(sexDE) %>%
  dplyr::mutate(proportion = n/sum(n)) %>%
  ggplot(aes(LOEUF_bin, proportion, fill = sexDE)) + 
  geom_col(position = position_stack()) + 
  geom_text(aes(label = n), position = position_stack(vjust =0.5), colour = "white", size = 2) + 
  theme_minimal() + xlab("") + ylab("") +
  scale_fill_manual(values = c("Male" = "darkblue", "Female" = "darkred", "No" = "lightgrey"),
                           name = "sex-DE")
# save plot
ggsave("adult_sexDE_vs_constraintBin.png",
       path = "images/5-postprocessing/1-GeneProperties/adult/",
       width = 12, height = 8, units = "cm")
```

## SexDE vs expression per constraint bin

As we know that the level of expression varies between different level of constraint, we look at both variable at the same time.

```{r violinplot_expr_constraintBin, echo=FALSE}
p <- ggplot(results[which(!is.na(results$LOEUF)),], aes(x = sexDE, y = AveExpr, fill = sexDE))
p <- p + geom_violin(alpha = 0.5, scale = "area", linewidth = 0.01) 
p <- p + stat_summary(fun.data=data_summary, size=0.1, show.legend = FALSE)
p <- p + facet_grid(cols = vars(LOEUF_bin), switch = "both")
p <- p + theme_minimal()
p <- p + theme(axis.text.x=element_blank(),
               axis.ticks.x=element_blank())
p <- p + xlab("Contraint bins") + ylab("Mean expression")
p <- p + scale_fill_manual(values = c("Male" = "darkblue", "Female" = "darkred", "No" = "lightgrey"),
                           name = "sex-DE")
p <- p + stat_compare_means(method = "wilcox.test", #p.adjust.method="BH",
                            label = "p.signif", hide.ns = TRUE, angle = 90, vjust = 0.7,
                            tip.length = 0, na.rm = TRUE, step.increase = 0,
                            ref.group = "No")
p
# save plot
ggsave("adult_sexDE_vs_expr_perConsBin.png",
       path = "images/5-postprocessing/1-GeneProperties/adult/",
       plot = p, width = 16, height = 8, units = "cm")
```


## SexDE vs constraint per expression bin

```{r violinplot_constraint_exprBin, echo=FALSE}
p <- ggplot(results[which(!is.na(results$LOEUF)),], aes(x = sexDE, y = LOEUF, fill = sexDE))
p <- p + geom_violin(alpha = 0.5, scale = "area", linewidth = 0.01) 
p <- p + stat_summary(fun.data=data_summary, size=0.1, show.legend = FALSE)
p <- p + facet_grid(cols = vars(exprbin), switch = "both")
p <- p + theme_minimal()
p <- p + theme(axis.text.x=element_blank(),
               axis.ticks.x=element_blank())
p <- p + xlab("Expression bins") + ylab("Constraint (LOEUF)")
p <- p + scale_fill_manual(values = c("Male" = "darkblue", "Female" = "darkred", "No" = "lightgrey"),
                           name = "sex-DE")
p <- p + stat_compare_means(method = "wilcox.test", #p.adjust.method="BH",
                            label = "p.signif", hide.ns = TRUE, angle = 90, vjust = 0.7,
                            tip.length = 0, na.rm = TRUE, step.increase = 0,
                            ref.group = "No")
p
# save plot
ggsave("adult_sexDE_vs_cons_perExprBin.png",
       path = "images/5-postprocessing/1-GeneProperties/adult/",
       plot = p, width = 16, height = 8, units = "cm")
```

## Sex-DE vs tissue-specificity

Plot the tau specificity score of sex-DE genes vs non sex-DE genes.

```{r violinplot_specificity, echo=FALSE}
p <- ggplot(results[which(!is.na(results$tau)),], aes(x = sexDE, y = tau, fill = sexDE))
p <- p + geom_violin(alpha = 0.5, scale = "area", linewidth = 0.01, show.legend = FALSE) 
p <- p + stat_summary(fun.data=data_summary, size=0.1, show.legend = FALSE)
p <- p + theme_minimal() + xlab("sex-DE") + ylab("Tissue-specificity (tau)")
p <- p + scale_fill_manual(values = c("Male" = "darkblue", "Female" = "darkred", "No" = "lightgrey"),
                           name = "sex-DE", )
p <- p + stat_compare_means(method = "wilcox.test",
                            label = "p.signif", hide.ns = TRUE, vjust = 0.7,
                            tip.length = 1, na.rm = TRUE, ref.group = "No")
p
# save plot
ggsave("adult_sexDE_vs_tissueSpecificity.png",
       path = "images/5-postprocessing/1-GeneProperties/adult/",
       width = 8, height = 8, units = "cm")
```

Plot the brain expression level of sex-DE genes vs non sex-DE genes.

```{r violinplot_brain_expression, echo=FALSE}
p <- ggplot(results[which(!is.na(results$tau)),], aes(x = sexDE, y = Brain, fill = sexDE))
p <- p + geom_violin(alpha = 0.5, scale = "area", linewidth = 0.01, show.legend = FALSE) 
p <- p + stat_summary(fun.data=data_summary, size=0.1, show.legend = FALSE)
p <- p + theme_minimal() + xlab("sex-DE") + ylab("Expression in brain")
p <- p + scale_fill_manual(values = c("Male" = "darkblue", "Female" = "darkred", "No" = "lightgrey"),
                           name = "sex-DE", )
p <- p + stat_compare_means(method = "wilcox.test",
                            label = "p.signif", hide.ns = TRUE, vjust = 0.7,
                            tip.length = 1, na.rm = TRUE, ref.group = "No")
p
# save plot
ggsave("adult_sexDE_vs_GTExBrainExpression.png",
       path = "images/5-postprocessing/1-GeneProperties/adult/",
       width = 8, height = 8, units = "cm")
```

Plot the level of expression in each GTEx tissue of sex-DE genes vs non sex-DE genes.

```{r violinplot_GTEx_expression, echo=FALSE}
plot_data <- results[,c(1,22,27:56)] %>% 
  pivot_longer(3:32, names_to = "Tissue", values_to = "Expression")
p <- ggplot(plot_data, aes(x = sexDE, y = Expression, fill = sexDE))
p <- p + geom_violin(alpha = 0.5, scale = "area", linewidth = 0.01, show.legend = FALSE, na.rm = TRUE) 
p <- p + stat_summary(fun.data=data_summary, size=0.1, show.legend = FALSE, na.rm = TRUE)
p <- p + facet_grid(cols = vars(Tissue), switch = "both")
p <- p + theme_minimal() + xlab("Tissue") + ylab("Expression")
p <- p + theme(axis.text.x=element_blank(),
               axis.ticks.x=element_blank(),
               strip.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
p <- p + scale_fill_manual(values = c("Male" = "darkblue", "Female" = "darkred", "No" = "lightgrey"),
                           name = "sex-DE", )
p <- p + stat_compare_means(method = "wilcox.test",
                            label = "p.signif", hide.ns = TRUE, angle = 90, vjust = 0.7,
                            tip.length = 0, na.rm = TRUE, step.increase = 0,
                            ref.group = "No")
p
# save plot
ggsave("adult_sexDE_vs_GTExExpression.png",
       path = "images/5-postprocessing/1-GeneProperties/adult/",
       width = 30, height = 8, units = "cm")
```

This plot is not very informative as in almost every tissue male-biased genes are more lowly expressed than non sex-DE genes while female-biased genes are more highly expressed...
There are 2 exceptions:
- in Pituitary, male-biased genes are more highly expressed then non sex-DE genes while there is no significant difference for female-biased genes.
- in Testis, all gene categories have the same mean level of expression.
As the male-biased genes are supposed to be more tissue-specific, how to determine in which tissue they are specific?

For each gene, if tau is > 0.8, get the tissue with the highest expression. Then, for the 3 genes categories, show a barplot of the number of genes with broad expression or the different tissues (for tissue specific expression).

```{r barplot_GTEx_expressionSpecificity_byCategory, echo=FALSE}
results$tissueSpecificity <- ifelse(is.na(results$tau_modif), "NA",
                                    ifelse(results$tau_modif>0.8, 
                                           unlist(apply(results[,27:56], 1, function(x){unlist(names(results[,27:56])[which.max(x)])})), 
                                           "No"))
results$tissueSpecificity <- factor(results$tissueSpecificity, levels = c("NA", "No", names(results[,27:56])))
# load color for GTEx tissue
gtex_tissues <- read.table("data/5-postprocessing/1-GeneProperties/gtex_tissue_colors_v8.csv", sep = ",", header = TRUE)
gtex_tissues <- gtex_tissues[, 4:5]
gtex_tissues$tissue_color_hex <- paste0("#", gtex_tissues$tissue_color_hex)
gtex_tissues <- gtex_tissues[!duplicated(gtex_tissues[,1:2]),]
colors <- gtex_tissues$tissue_color_hex
names(colors) <- gsub(" ", "_", gtex_tissues$tissue_site)
colors <- c("NA" = "#Ffffff", "No" = "#F7f7f7", colors)
# select some tissue for which we want to show the number of genes
selectedTissues <- c("Brain", "Pituitary", "Testis")
# plot
results %>% 
  dplyr::group_by(sexDE) %>%
  dplyr::count(tissueSpecificity) %>%
  dplyr::mutate(proportion = n/sum(n)) %>%
  ggplot(aes(sexDE, proportion, fill = tissueSpecificity)) + 
  geom_col(position = position_stack()) + 
  geom_text(aes(label=ifelse(tissueSpecificity %in% selectedTissues, n, "")), position = position_stack(vjust =0.5), colour = "white", size = 2) + 
  theme_minimal() + xlab("sexDE") + ylab("Gene proportion") +
  scale_fill_manual(values = colors,
                    name = "Tissue specificity")
# save plot
ggsave("adult_sexDE_vs_GTExTissueSpecificity_byCategory.png",
       path = "images/5-postprocessing/1-GeneProperties/adult/",
       width = 12, height = 12, units = "cm")
```

## SexDE vs sexDE sharing across GTEx tissues

Analyze the overlap between sex-DE genes in forebrain and shared sex-DE genes in GTEx tissues.

```{r analyze_overlap_gtex_sexDE, echo=FALSE}
# check if the shared effect has the same direction as in our analysis
overlap_gtexSharing_sexDE$sameDir <- ifelse(overlap_gtexSharing_sexDE$dirGTEx == overlap_gtexSharing_sexDE$sexDE, 1, 0)
# mean percent sex-DE
overlap_gtexSharing_sexDE %>% group_by(sexDE) %>% summarise(nb = n(), nb_sameDir = sum(sameDir), mean_percent_sexDE = mean(percent_sexDE))
# test difference in mean
compare_means(percent_sexDE~sexDE, overlap_gtexSharing_sexDE, method = "wilcox.test")
# format data for plot
overlap_gtexSharing_sexDE$sexDE <- factor(overlap_gtexSharing_sexDE$sexDE, 
                                         levels = c("No", "Male", "Female"))
```

Plots

```{r plot_overlap_gtex_sexDE, echo=FALSE}
# plot
p <- ggplot(overlap_gtexSharing_sexDE, aes(x = sexDE, y = percent_sexDE, fill = sexDE))
p <- p + geom_violin(alpha = 0.5, scale = "area", size = 0.01, show.legend = FALSE)
p <- p + stat_summary(fun.data=data_summary, size=0.1, show.legend = FALSE)
p <- p + scale_fill_manual(values = c("Male" = "darkblue", "Female" = "darkred", "No" = "lightgrey"),
                           name = "sex-DE", )
p <- p + theme_minimal() + xlab("") + ylab("% GTEx tissue pairs with\nconsistent sex-DE")
p <- p + stat_compare_means(method = "wilcox.test", label = "p.signif", 
                            hide.ns = TRUE, label.x.npc = 'middle',
                            vjust = 0.1, ref.group = "No")
p
# save plot
ggsave("adult_sexDE_sharing_byCat.png",
       path = "images/5-postprocessing/1-GeneProperties/adult/",
       plot = p, width = 8, height = 8, units = "cm")
```


# Write a file containing all data

```{r write_data, echo=FALSE}
results_final <- results[, c(1:26,59)]
write.table(results_final, file = "data/5-postprocessing/1-GeneProperties/adult/adult_forebrain_geneProperties.txt",
           quote = FALSE, sep = "\t", row.names = FALSE, col.names = TRUE)
```

