---
title: "Correlation  between pseudotime and cell type decomposition"
author: "Clara Benoit-Pilven"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
---

```{r load_libraries, echo=FALSE}
suppressPackageStartupMessages( library(tidyverse) )
suppressPackageStartupMessages( library(forecast) )
```

```{r, setup, include=FALSE, echo=FALSE}
knitr::opts_knit$set(root.dir = '~/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/')
```

# Load data

## Load pseudotime data

```{r load_pseudotime, echo=FALSE}
load("data/2-pseudotime/forebrain_metadata_pseudotime.RData")
```


## Load cell type decomposition analysis

Load CIBERSORTx result file

```{r load_cibersortx_result, echo=FALSE}
cell_prop <- read.table("data/5-postprocessing/2-CellType/CIBERSORTx/prenatal/results/CIBERSORTx_Adjusted.txt",
                        header = TRUE, sep = "\t")
```

Merge the cell type decomposition results and the pseudotime.

```{r merge_ct_pseudotime, echo=FALSE}
cell_prop <- merge(cell_prop, metadata, by.x = "Mixture", by.y = 0)
```

# Correlation of the cell types proportions with pseudotimes

For separate cell types:

```{r cor_pseudotime_NSC, echo=FALSE}
corTestRes <- cor.test(cell_prop$NPC, cell_prop$pseudotime)
p <- ggplot(cell_prop, aes(y = NPC, x = pseudotime))
p <- p + geom_point()
p <- p + ylab("Proportion of neuronal stem cells") + xlab("Pseudotime")
p <- p + theme_minimal()
p <- p + annotate("text", x = 0.9, y = 0.75, 
                  label = sprintf("r = %.2f\n p-value = %.2e", corTestRes$estimate, corTestRes$p.value))
p
# save plot
ggsave("NPC_VS_pseudotime.png",
       path = "images/5-postprocessing/2-CellType/prenatal/",
       plot = p, width = 15, height = 10, units = "cm")
```

```{r cor_pseudotime_neurons_ex, echo=FALSE}
corTestRes <- cor.test(cell_prop$ExN, cell_prop$pseudotime)
p <- ggplot(cell_prop, aes(y = ExN, x = pseudotime))
p <- p + geom_point()
p <- p + ylab("Proportion of excitatory neurons") + xlab("Pseudotime")
p <- p + theme_minimal()
p <- p + annotate("text", x = -0.75, y = 0.6, 
                  label = sprintf("r = %.2f\n p-value = %.2e", corTestRes$estimate, corTestRes$p.value))
p
# save plot
ggsave("ExNeurons_VS_pseudotime.png",
       path = "images/5-postprocessing/2-CellType/prenatal/",
       plot = p, width = 15, height = 10, units = "cm")
```

```{r cor_pseudotime_neurons_in, echo=FALSE}
corTestRes <- cor.test(cell_prop$InN, cell_prop$pseudotime)
p <- ggplot(cell_prop, aes(y = InN, x = pseudotime))
p <- p + geom_point()
p <- p + ylab("Proportion of interneurons") + xlab("Pseudotime")
p <- p + theme_minimal()
p <- p + annotate("text", x = -0.75, y = 0.5, 
                  label = sprintf("r = %.2f\n p-value = %.2e", corTestRes$estimate, corTestRes$p.value))
p
# save plot
ggsave("Interneurons_VS_pseudotime.png",
       path = "images/5-postprocessing/2-CellType/prenatal/",
       plot = p, width = 15, height = 10, units = "cm")
```

```{r cor_pseudotime_neurons, echo=FALSE}
corTestRes <- cor.test(cell_prop$ExN + cell_prop$InN, cell_prop$pseudotime)
p <- ggplot(cell_prop, aes(y = InN + ExN, x = pseudotime))
p <- p + geom_point()
p <- p + ylab("Proportion of neurons") + xlab("Pseudotime")
p <- p + theme_minimal()
p <- p + annotate("text", x = -0.75, y = 0.8, 
                  label = sprintf("r = %.2f\n p-value = %.2e", corTestRes$estimate, corTestRes$p.value))
p
# save plot
ggsave("AllNeurons_VS_pseudotime.png",
       path = "images/5-postprocessing/2-CellType/prenatal/",
       plot = p, width = 15, height = 10, units = "cm")
```

```{r cor_pseudotime_OPC, echo=FALSE}
corTestRes <- cor.test(cell_prop$OPC, cell_prop$pseudotime)
p <- ggplot(cell_prop, aes(y = OPC, x = pseudotime))
p <- p + geom_point()
p <- p + ylab("Proportion of OPC") + xlab("Pseudotime")
p <- p + theme_minimal()
p <- p + annotate("text", x = 0.75, y = 0.2, 
                  label = sprintf("r = %.2f\n p-value = %.2e", corTestRes$estimate, corTestRes$p.value))
p
# save plot
ggsave("OPC_VS_pseudotime.png",
       path = "images/5-postprocessing/2-CellType/prenatal/",
       plot = p, width = 15, height = 10, units = "cm")
```

Correlation of the different cell types with the pseudotime.

```{r cor_ct_VS_pseudotime, echo=FALSE}
print("NPC")
cor(cell_prop$NPC, cell_prop$pseudotime)
print("ExN")
cor(cell_prop$ExN, cell_prop$pseudotime)
print("InNC")
cor(cell_prop$InN, cell_prop$pseudotime)
print("Neurons")
cor(cell_prop$InN + cell_prop$ExN, cell_prop$pseudotime)
print("Micro")
cor(cell_prop$Micro, cell_prop$pseudotime)
print("OPC")
cor(cell_prop$OPC, cell_prop$pseudotime)
print("Astro")
cor(cell_prop$Astro, cell_prop$pseudotime)
print("Olig")
cor(cell_prop$Olig, cell_prop$pseudotime)
print("Endo")
cor(cell_prop$Endo, cell_prop$pseudotime)
print("Perc")
cor(cell_prop$Perc, cell_prop$pseudotime)
```


# Plot the cell type composition with a moving average

## Compute the moving average

Compute the moving average proportion for each cell type along pseudotime

```{r compute_ma, echo=FALSE}
cell_prop <- cell_prop[order(cell_prop$pseudotime),]
cell_prop_ma <- as.data.frame(lapply(cell_prop[2:10], ma, order = 20))
# remove the first and last lines
cell_prop_ma <- cell_prop_ma[complete.cases(cell_prop_ma), ]
```

Prepare the data for the plots.

```{r format_prop_ma, echo=FALSE}
# add an index for the plot
cell_prop_ma$idx <- rownames(cell_prop_ma)
cell_prop_ma_plot <- gather(cell_prop_ma, c(1:9), key = "cell_type", value = "proportion")
cell_prop_ma_plot$cell_type <- factor(cell_prop_ma_plot$cell_type, 
                                             levels = c("NPC", "ExN", "InN", "OPC", "Olig", "Astro", "Micro", "Endo", "Perc"))
```

Do the same moving average separating the samples by sex.

```{r compute_ma_sex, echo=FALSE}
## Males
cell_prop_M <- cell_prop[which(cell_prop$Sex == "1"), ]
cell_prop_M <- cell_prop_M[order(cell_prop_M$pseudotime),]
cell_prop_M_ma <- as.data.frame(lapply(cell_prop_M[2:10], ma, order = 20))
# remove the first and last lines
cell_prop_M_ma <- cell_prop_M_ma[complete.cases(cell_prop_M_ma), ]
rownames(cell_prop_M_ma) <- NULL
## Females
cell_prop_F <- cell_prop[which(cell_prop$Sex == "2"), ]
cell_prop_F <- cell_prop_F[order(cell_prop_F$pseudotime),]
cell_prop_F_ma <- as.data.frame(lapply(cell_prop_F[2:10], ma, order = 20))
# remove the first and last lines
cell_prop_F_ma <- cell_prop_F_ma[complete.cases(cell_prop_F_ma), ]
rownames(cell_prop_F_ma) <- NULL
```

```{r format_prop_ma_sex, echo=FALSE}
# add an index for the plot
cell_prop_M_ma$idx <- rownames(cell_prop_M_ma)
cell_prop_F_ma$idx <- rownames(cell_prop_F_ma)
# merge male and female data
cell_prop_sex_ma <- bind_rows(cell_prop_M_ma , cell_prop_F_ma , .id = 'Sex')
cell_prop_sex_ma_plot <- gather(cell_prop_sex_ma, c(2:10), 
                                       key = "cell_type", value = "proportion")
cell_prop_sex_ma_plot$cell_type <- factor(cell_prop_sex_ma_plot$cell_type, 
                                             levels = c("NPC", "ExN", "InN", "OPC", "Olig", "Astro", "Micro", "Endo", "Perc"))
cell_prop_sex_ma_plot$sex <- factor(cell_prop_sex_ma_plot$Sex, levels = c("1","2"), 
                                           labels = c("M", "F"))
```

## Plot the cell type composition with a moving average

### For both sexes:

```{r stacked_barplot_ma, echo=FALSE}
p <- ggplot(cell_prop_ma_plot, aes(x=as.integer(idx),y=proportion, fill=cell_type))
p <- p + geom_bar(position="stack", stat="identity", width = 1)
p <- p + scale_fill_manual(values = c("NPC" = "#f4ccd8", "ExN" = "#cce3c1", "InN" = "#f6c9a6",
                                      "Micro" = "#b2b2ae", "Astro" = "#c2b29d", "OPC" = "#c9c6d9",
                                      "Endo" = "#eed9c4", "Olig" = "#ccdff4", "Perc" = "#e4d5b7"),
                           name = "Cell types")
p <- p + theme_bw() + ylab("Cell type proportions") + xlab("pseudotime")
p <- p + theme(panel.border = element_blank(),
               panel.grid.major = element_blank(), 
               panel.grid.minor = element_blank(),
               axis.text.x = element_blank(),
               axis.ticks = element_blank()) 
p
# save plot
ggsave("Cibersortx_Song_forebrain_pseudotime_ma_stackedBarplot_9CellTypes.png", 
       path = "images/5-postprocessing/2-CellType/prenatal/",
       plot = p, width = 30, height = 12, units = "cm")
```

## Separating by sexes:

```{r stacked_barplot_sex_ma, echo=FALSE}
p <- ggplot(cell_prop_sex_ma_plot, aes(x=as.integer(idx),y=proportion, fill=cell_type))
p <- p + geom_bar(position="stack", stat="identity", width = 1)
p <- p + scale_fill_manual(values = c("NPC" = "#f4ccd8", "ExN" = "#cce3c1", "InN" = "#f6c9a6",
                                      "Micro" = "#b2b2ae", "Astro" = "#c2b29d", "OPC" = "#c9c6d9",
                                      "Endo" = "#eed9c4", "Olig" = "#ccdff4", "Perc" = "#e4d5b7"),
                           name = "Cell types")
p <- p + facet_grid(cell_prop_sex_ma_plot$Sex~., labeller = labeller(Sex = c("1" = "Males", "2" = "Female")))
p <- p + theme_bw() + ylab("Cell type proportions") + xlab("pseudotime")
p <- p + theme(strip.background = element_blank(),
               panel.border = element_blank(),
               panel.grid.major = element_blank(), 
               panel.grid.minor = element_blank(),
               axis.text.x = element_blank(),
               axis.ticks = element_blank()) 
p
# save plot
ggsave("Cibersortx_Song_forebrain_pseudotime_ma_sex_stackedBarplot_9CellTypes.png", 
       path = "images/5-postprocessing/2-CellType/prenatal/",
       plot = p, width = 30, height = 12, units = "cm")
```


# Session Info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

