---
title: "Enrichment with hypergeometric test of genes differentially expressed between case and controls for different diseases"
author: "Clara Benoit-Pilven"
date: "`r Sys.Date()`"
output: html_document
---

```{r load_library, echo=FALSE}
suppressPackageStartupMessages( library(edgeR) )
suppressPackageStartupMessages( library(limma) )
suppressPackageStartupMessages( library(qvalue) )
suppressPackageStartupMessages( library(tidyverse) )
```

```{r, setup, include=FALSE, echo=FALSE}
knitr::opts_knit$set(root.dir = '~/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/')
```

# Load data and functions

## Load functions

```{r load_function, echo=FALSE}
source("scripts/analyses/5-postprocessing/4-DiseaseEnrichment/enrichment_hyper.R")
```


## Load DE results

### HDBR

Load HDBR sex-DE data.

```{r load_data_HDBR_sexDE, echo=FALSE}
hdbr_sexDE_results <- read.table("data/3-DE/VoomDream_topTable_forebrain_pseudotime_sva_ind_interaction.txt",
                         header = TRUE, check.names=FALSE)
# remove genes used for sex assignment
genes2remove <- c("XIST", as.character(hdbr_sexDE_results$gene_name[which(hdbr_sexDE_results$chr == "chrY" & hdbr_sexDE_results$gene_type == "protein_coding")]))
hdbr_sexDE_results <- hdbr_sexDE_results[which(!hdbr_sexDE_results$gene_name %in% genes2remove),]
# format
hdbr_sexDE_results$ensemblID <- sapply(strsplit(as.character(hdbr_sexDE_results$gene_id), "\\."), function(x){x[1]})
hdbr_sexDE_results$exprbin <- cut(hdbr_sexDE_results$AveExpr, quantile(hdbr_sexDE_results$AveExpr, seq(0,1,0.1)), 
                                  labels=c(1:10), include.lowest=T) # add expression bins
```

Load HDBR pseudotime-DE data.

```{r load_data_HDBR_pseudotimeDE, echo=FALSE}
hdbr_pseudotimeDE_results <- read.table("data/3-DE/VoomDream_topTable_forebrain_pseudotime_sva_ind_interaction_pseudotime.txt",
                         header = TRUE, check.names=FALSE)
# remove genes used to infer pseudotime
load("data/2-pseudotime/forebrain_pseudotime_analysis.RData")
genes2remove <- rownames(selectedCounts)
hdbr_pseudotimeDE_results <- hdbr_pseudotimeDE_results[which(!hdbr_pseudotimeDE_results$gene_name %in% genes2remove),]
# format
hdbr_pseudotimeDE_results$ensemblID <- sapply(strsplit(as.character(hdbr_pseudotimeDE_results$gene_id), "\\."), function(x){x[1]})
hdbr_pseudotimeDE_results$exprbin <- cut(hdbr_pseudotimeDE_results$AveExpr, quantile(hdbr_pseudotimeDE_results$AveExpr, seq(0,1,0.1)), 
                                         labels=c(1:10), include.lowest=T) # add expression bins
```


### GTEx

Load GTEx sex-DE data.

```{r load_data_GTEx_sexDE, echo=FALSE}
gtex_sexDE_results <- read.table("data/4-GTEx/1-GTEx_analysis/brain_forebrain_v8_VoomDream_topTable_with_sva_ind_pseudotime_interaction.txt",
                         header = TRUE, check.names=FALSE)
# remove genes used for sex assignment
genes2remove <- c("XIST", as.character(gtex_sexDE_results$gene_name[which(gtex_sexDE_results$chr == "chrY" & gtex_sexDE_results$gene_type == "protein_coding")]))
gtex_sexDE_results <- gtex_sexDE_results[which(!gtex_sexDE_results$gene_name %in% genes2remove),]
# format
gtex_sexDE_results$ensemblID <- sapply(strsplit(as.character(gtex_sexDE_results$gene_id), "\\."), function(x){x[1]})
gtex_sexDE_results$exprbin <- cut(gtex_sexDE_results$AveExpr, quantile(gtex_sexDE_results$AveExpr, seq(0,1,0.1)), 
                                  labels=c(1:10), include.lowest=T) # add expression bins
```

Load GTEx pseudotime-DE data.

```{r load_data_GTEx_pseudotimeDE, echo=FALSE}
gtex_pseudotimeDE_results <- read.table("data/4-GTEx/1-GTEx_analysis/brain_forebrain_v8_VoomDream_topTable_with_sva_ind_pseudotime_interaction_pseudotime.txt",
                                        header = TRUE, check.names=FALSE)
# remove genes used to infer pseudotime
load("data/4-GTEx/1-GTEx_analysis/GTEx_v8_forebrain_pseudotime_analysis.RData")
genes2remove <- rownames(selectedCounts)
gtex_pseudotimeDE_results <- gtex_pseudotimeDE_results[which(!gtex_pseudotimeDE_results$gene_name %in% genes2remove),]
# format
gtex_pseudotimeDE_results$ensemblID <- sapply(strsplit(as.character(gtex_pseudotimeDE_results$gene_id), "\\."), function(x){x[1]})
gtex_pseudotimeDE_results$exprbin <- cut(gtex_pseudotimeDE_results$AveExpr, quantile(gtex_pseudotimeDE_results$AveExpr, seq(0,1,0.1)), 
                                         labels=c(1:10), include.lowest=T) # add expression bins
```


### Bayesian score

```{r load_bayes_results, echo=FALSE}
bayesResults <- read.table("data/4-GTEx/2-comparison/lineModel/lineModel_prenatalVSadult_6models.txt", header = TRUE)
# merge bayes results with prenatal sex-DE results
bayesResults_merged <- merge(bayesResults, hdbr_sexDE_results, by = "ensemblID", suffixes = c("", "_DE"))
# format specificity variable
bayesResults_merged$specificity <- factor(bayesResults_merged$specificity,
                                          levels = c("ADULT", "OPPOSITE", "PRENATAL", "SHARED_UNSPECIFIED", 
                                                     "SHARED0.5", "SHARED1", "SHARED2", "UNCLASSIFIED"),
                                          labels = c("adult_specific", "opposite", "prenatal_specific", "shared", 
                                                     "shared", "shared", "shared", "unclassified"))
```


## Load genesets

```{r load_geneList, echo=FALSE}
load("data/5-postprocessing/8-CaseControlEnrichment/geneLists/caseCtrlGeneLists.RData")
```


## Load constraint data

LOEUF score

```{r load_LOEUF, echo=FALSE}
# load loeuf score
constraint_data <- read.table("data/5-postprocessing/1-GeneProperties/gnomad.v4.1.constraint_metrics.tsv", 
                              header = TRUE, sep = "\t")
constraint_data <- constraint_data[c("gene_id", "canonical", "mane_select", "lof.oe_ci.upper")]
# keep MANE transcript or canonical if no MANE transcript for a gene
constraint_data <- constraint_data[which((constraint_data$mane_select == "true" | constraint_data$canonical == "true") & startsWith(constraint_data$gene_id, "ENSG")), ]
constraint_data <- constraint_data[c("gene_id", "lof.oe_ci.upper")]
names(constraint_data) <- c("gene_id", "LOEUF")
```


# Enrichment analysis

The enrichment analysis is done for the following condition:
- 6 analyses lists: prenatal sex-DE and pseudotime-DE, adult sex-DE and pseudotime-DE, shared sex-DE and prenatal-specific sex-DE genes.
- 3 directions: both, female-biased and male-biased
- with or without controlling for gene expression level

```{r enrichment_analyses, echo=FALSE}
enrichRes <- data.frame("analysis" = NA, "direction" = NA, "geneList" = NA, "nbr_DE" = NA, "nbr_geneList" = NA, 
                        "overlap" = NA, "N" = NA, "pval_enrich" = NA, "pval_deple" = NA, "relative_enrich" = NA,
                        "pval_enrich_random" = NA, "pval_deple_random" = NA, "relative_enrich_random" = NA, 
                        "ctrlExpr" = NA)
enrichRes <- enrichRes[0,]

analysesList <- list(hdbr_sexDE_results, hdbr_pseudotimeDE_results, gtex_sexDE_results, gtex_pseudotimeDE_results, bayesResults_merged, bayesResults_merged)
analysesNameList <- c("prenatal_sexDE", "prenatal_pseudotimeDE", "adult_sexDE", "adult_pseudotimeDE", "prenatal_specific", "shared")
analysesTypeList <- c("sexDE", "pseudotimeDE", "sexDE", "pseudotimeDE", "sexDE", "sexDE")
directions <- c("both", "up", "down")
ctrlExprList <- c(FALSE, TRUE)
for (i in 1:length(analysesList)){
  analyseData <- analysesList[[i]]
  analyseName <- analysesNameList[i]
  analyseType <- analysesTypeList[i]
  print(paste0(">>>>> ", analyseName))
  for (ctrlExpr in ctrlExprList){
    print(paste0(">>> ctrlExpr=", as.character(ctrlExpr)))
    for (myDirection in directions){
      print(paste0("> direction=", myDirection))
      bayes <- FALSE
      bayesDataType <- NULL
      if (analyseName %in% c("prenatal_specific", "shared")){
        bayes <- TRUE
        bayesDataType <- analyseName
      }
      resTmp <- enrichment_analysis_permut(DEresults = analyseData, geneList = genesLists, 
                                           analysisName = analyseName, geneListNames = names(genesLists),
                                           nbr_permut = 1000, pval_threshold = 0.05, direction = myDirection, 
                                           bayes = bayes, bayesDataType = bayesDataType, ctrlExpr = ctrlExpr, 
                                           geneName = FALSE)
      if (analyseType == "sexDE"){
        if (myDirection == "up"){
          resTmp$direction <- "male"
        } else if (myDirection == "down"){
          resTmp$direction <- "female"
        }
      }
      # add results to the global enrichment abject
      enrichRes <- rbind(enrichRes, resTmp)
    }
  }
}
```

# Write results

Add percentage of the DE genes are in gene list.

```{r add_percent, echo=FALSE}
enrichRes$percent <- enrichRes$overlap/enrichRes$nbr_DE*100
```

Write the complete enrichment results into a file.

```{r write_res, echo=FALSE}
write.table(enrichRes, file = "data/5-postprocessing/8-CaseControlEnrichment/results_hyper/enrichment_results.txt", 
            quote = FALSE, row.names = FALSE, col.names = TRUE, sep = "\t")
```


# Plot the results

What is the best layout to plot all these information?
We do 2 different plots: one for the bayes results and another for the other results
For both plots, we plot the phenotypes on the y axis and separate the adult and prenatal results on the x axis as well as the sex-DE and pseudotime-DE for each dataset. For each combination, we plot both the relative enrichment value for the geneset of interest as well as a violinplot of the 1000 relative enrichment value for the random genesets.

## Plot for sex-DE and pseudotime-DE results without controlling for gene level of expression

We plot the results for all phenotypes with both direction (up/down or male/female) in the pseudotime-DE and sex-DE analyses for the adult and the prenatal dataset.

First, we format a bit more the data. The chosen threshold to consider a result as significant is 0.05/6 datasets * 2 directions * 10 phenotypes (~120 analyses) = 0.0004.

```{r format4plot, echo=FALSE}
enrichRes_formated <- enrichRes[which(enrichRes$analysis %in% c("prenatal_sexDE", "prenatal_pseudotimeDE", "adult_sexDE", "adult_pseudotimeDE") & enrichRes$direction != "both" & !enrichRes$ctrlExpr), ]
# create groups for facet
enrichRes_formated$dataset <- ifelse(enrichRes_formated$analysis %in% c("prenatal_sexDE", "prenatal_pseudotimeDE"), "prenatal", "adult")
# add * for signif enrichment or depletion (with a threshold of 0.05/120=0.0004)
enrichRes_formated$signif <- ifelse(enrichRes_formated$relative_enrich > 1, 
                                    ifelse(enrichRes_formated$pval_enrich<0.0004 & enrichRes_formated$permut_pval < 0.0004, "*", ""),
                                    ifelse(enrichRes_formated$pval_deple<0.0004 & enrichRes_formated$permut_pval < 0.0004, "*", ""))
# order the different direction of effect
enrichRes_formated$direction <- factor(enrichRes_formated$direction,
                                       levels = c("female", "male", "down", "up"))
# order the different phenotype
enrichRes_formated$geneList <- factor(enrichRes_formated$geneList,
                                      levels = c("sczDown", "sczUp", "asdDown", "asdUp", "asdM12", "asdM16", 
                                                 "mddDownFemale", "mddUpFemale", "mddDownMale", "mddUpMale"),
                                      labels = c("Schizophrenia \ndown", "Schizophrenia \nup", "Autism \ndown", "Autism \nup",
                                                 "Autism \ndown M12", "Autism \nup M16",
                                                 "Major Depressive Disorder \ndown female", "Major Depressive Disorder \nup female",
                                                 "Major Depressive Disorder \ndown male", "Major Depressive Disorder \nup male"))
# rename the 4 analyses
enrichRes_formated$analysis <- factor(enrichRes_formated$analysis,
                                      levels = c("adult_pseudotimeDE", "adult_sexDE", "prenatal_pseudotimeDE", "prenatal_sexDE"),
                                      labels = c("Adult\nPseudotime-DE", "Adult\nSex-DE", "Prenatal\nPseudotime-DE", "Prenatal\nSex-DE"))

# extract and transform the results for the random set of genes
enrichRes_random_formated <- enrichRes_formated[, c("geneList", "analysis", "direction", "dataset", "pval_enrich_random", "pval_deple_random", "relative_enrich_random")] %>% separate_longer_delim(cols = c(pval_enrich_random, pval_deple_random, relative_enrich_random), delim = ";")
enrichRes_random_formated$pval_enrich_random <- as.numeric(enrichRes_random_formated$pval_enrich_random)
enrichRes_random_formated$pval_deple_random <- as.numeric(enrichRes_random_formated$pval_deple_random)
enrichRes_random_formated$relative_enrich_random <- as.numeric(enrichRes_random_formated$relative_enrich_random)
```


```{r plot, echo=FALSE}
p <- ggplot(enrichRes_random_formated, aes(y = geneList, x = relative_enrich_random, fill = direction, alpha = 0.4))
p <- p + geom_violin(linewidth = 0.1)
p <- p + geom_vline(xintercept = 1)
p <- p + scale_fill_manual(values = c("female" = "darkred", "male" = "darkblue", "down" = "chocolate", "up" = "aquamarine4"))
p <- p + geom_point(data = enrichRes_formated, aes(y = geneList, x = relative_enrich, color = direction), 
                    position = position_dodge(0.9), size = 2, alpha = 1)
p <- p + geom_text(data = enrichRes_formated, aes(y = geneList, x = ifelse(relative_enrich>1, relative_enrich + 0.3, relative_enrich - 0.3), label = signif),
                   position = position_dodge(0.9), size = 4, alpha = 1)
p <- p + scale_color_manual(values = c("female" = "darkred", "male" = "darkblue", "down" = "chocolate", "up" = "aquamarine4"))
p <- p + facet_grid(cols = vars(analysis), scales = "free", space = "free")
p <- p + theme_minimal()
p <- p + ylab("") + xlab("Relative enrichment")
p <- p + guides(alpha="none")
p <- p + coord_cartesian(xlim = c(0,5))
p <- p + scale_y_discrete(limits = rev(levels(enrichRes_random_formated$geneList))) # reverse order of phenotype
p
# save plot
ggsave("enrichment_plot_sexDE_pseudotimeDE_adult_prenatal.png",
      path = "images/5-postprocessing/8-CaseControlEnrichment/hyper/",
      units = "cm", width = 20, height = 12)
```

## Plot for bayes results without controlling for gene level of expression

We plot the results for all phenotypes with both direction (male/female) in the prenatal-specific and shared sex-DE genes.

First, we format a bit more the data. The chosen threshold to consider a result as significant is 0.05/6 datasets * 2 directions * 10 phenotypes (~120 analyses) = 0.0004.

```{r format4plot_bayes, echo=FALSE}
enrichRes_formated_bayes <- enrichRes[which(enrichRes$analysis %in% c("prenatal_specific", "shared") & enrichRes$direction != "both" & !enrichRes$ctrlExpr), ]
# add * for signif enrichment or depletion (with a threshold of 0.05/120=0.0004)
enrichRes_formated_bayes$signif <- ifelse(enrichRes_formated_bayes$relative_enrich > 1, 
                                    ifelse(enrichRes_formated_bayes$pval_enrich<0.0004 & enrichRes_formated_bayes$permut_pval < 0.0004, "*", ""),
                                    ifelse(enrichRes_formated_bayes$pval_deple<0.0004 & enrichRes_formated_bayes$permut_pval < 0.0004, "*", ""))
# order the different direction of effect
enrichRes_formated_bayes$direction <- factor(enrichRes_formated_bayes$direction,
                                       levels = c("female", "male"))
# order the different phenotype
enrichRes_formated_bayes$geneList <- factor(enrichRes_formated_bayes$geneList,
                                            levels = c("sczDown", "sczUp", "asdDown", "asdUp", "asdM12", "asdM16", 
                                                       "mddDownFemale", "mddUpFemale", "mddDownMale", "mddUpMale"),
                                            labels = c("Schizophrenia \ndown", "Schizophrenia \nup", "Autism \ndown", "Autism \nup",
                                                       "Autism \ndown M12", "Autism \nup M16",
                                                       "Major Depressive Disorder \ndown female", "Major Depressive Disorder \nup female",
                                                       "Major Depressive Disorder \ndown male", "Major Depressive Disorder \nup male"))
# rename the 4 analyses
enrichRes_formated_bayes$analysis <- factor(enrichRes_formated_bayes$analysis,
                                      levels = c("prenatal_specific", "shared"),
                                      labels = c("Prenatal\nspecific", "Shared"))

# extract and transform the results for the random set of genes
enrichRes_random_formated_bayes <- enrichRes_formated_bayes[, c("geneList", "analysis", "direction", "pval_enrich_random", "pval_deple_random", "relative_enrich_random")] %>% separate_longer_delim(cols = c(pval_enrich_random, pval_deple_random, relative_enrich_random), delim = ";")
enrichRes_random_formated_bayes$pval_enrich_random <- as.numeric(enrichRes_random_formated_bayes$pval_enrich_random)
enrichRes_random_formated_bayes$pval_deple_random <- as.numeric(enrichRes_random_formated_bayes$pval_deple_random)
enrichRes_random_formated_bayes$relative_enrich_random <- as.numeric(enrichRes_random_formated_bayes$relative_enrich_random)
```


```{r plot_bayes, echo=FALSE}
p <- ggplot(enrichRes_random_formated_bayes, aes(y = geneList, x = relative_enrich_random, fill = direction, alpha = 0.4))
p <- p + geom_violin(linewidth = 0.1)
p <- p + geom_vline(xintercept = 1)
p <- p + scale_fill_manual(values = c("female" = "darkred", "male" = "darkblue"))
p <- p + geom_point(data = enrichRes_formated_bayes, aes(y = geneList, x = relative_enrich, color = direction), 
                    position = position_dodge(0.9), size = 2, alpha = 1)
p <- p + geom_text(data = enrichRes_formated_bayes, aes(y = geneList, x = ifelse(relative_enrich>1, relative_enrich + 0.3, relative_enrich - 0.3), label = signif),
                   position = position_dodge(0.9), size = 4, alpha = 1)
p <- p + scale_color_manual(values = c("female" = "darkred", "male" = "darkblue"))
p <- p + facet_grid(cols = vars(analysis), scales = "free", space = "free")
p <- p + theme_minimal()
p <- p + ylab("") + xlab("Relative enrichment")
p <- p + guides(alpha="none")
p <- p + coord_cartesian(xlim = c(0,7))
p <- p + scale_y_discrete(limits = rev(levels(enrichRes_random_formated_bayes$geneList))) # reverse order of phenotype
p
# save plot
ggsave("enrichment_plot_sexDE_bayes.png",
      path = "images/5-postprocessing/8-CaseControlEnrichment/hyper/",
      units = "cm", width = 15, height = 12)
```

## Plot for sex-DE and pseudotime-DE results controlling for gene level of expression

We plot the results for all phenotypes with both direction (up/down or male/female) in the pseudotime-DE and sex-DE analyses for the adult and the prenatal dataset.

First, we format a bit more the data. The chosen threshold to consider a result as significant is 0.05/6 datasets * 2 directions * 10 phenotypes (~120 analyses) = 0.0004.

```{r format4plot_ctrlExpr, echo=FALSE}
enrichRes_formated_ctrlExpr <- enrichRes[which(enrichRes$analysis %in% c("prenatal_sexDE", "prenatal_pseudotimeDE", "adult_sexDE", "adult_pseudotimeDE") & enrichRes$direction != "both" & enrichRes$ctrlExpr), ]
# create groups for facet
enrichRes_formated_ctrlExpr$dataset <- ifelse(enrichRes_formated_ctrlExpr$analysis %in% c("prenatal_sexDE", "prenatal_pseudotimeDE"), "prenatal", "adult")
# add * for signif enrichment or depletion (with a threshold of 0.05/120=0.0004)
enrichRes_formated_ctrlExpr$signif <- ifelse(enrichRes_formated_ctrlExpr$relative_enrich > 1, 
                                             ifelse(enrichRes_formated_ctrlExpr$pval_enrich<0.0004 & enrichRes_formated_ctrlExpr$permut_pval < 0.0004, "*", ""),
                                             ifelse(enrichRes_formated_ctrlExpr$pval_deple<0.0004 & enrichRes_formated_ctrlExpr$permut_pval < 0.0004, "*", ""))
# order the different direction of effect
enrichRes_formated_ctrlExpr$direction <- factor(enrichRes_formated_ctrlExpr$direction,
                                                levels = c("female", "male", "down", "up"))
# order the different phenotype
enrichRes_formated_ctrlExpr$geneList <- factor(enrichRes_formated_ctrlExpr$geneList,
                                               levels = c("sczDown", "sczUp", "asdDown", "asdUp", "asdM12", "asdM16", 
                                                          "mddDownFemale", "mddUpFemale", "mddDownMale", "mddUpMale"),
                                               labels = c("Schizophrenia \ndown", "Schizophrenia \nup", "Autism \ndown", "Autism \nup",
                                                          "Autism \ndown M12", "Autism \nup M16",
                                                          "Major Depressive Disorder \ndown female", "Major Depressive Disorder \nup female",
                                                          "Major Depressive Disorder \ndown male", "Major Depressive Disorder \nup male"))
# rename the 4 analyses
enrichRes_formated_ctrlExpr$analysis <- factor(enrichRes_formated_ctrlExpr$analysis,
                                               levels = c("adult_pseudotimeDE", "adult_sexDE", "prenatal_pseudotimeDE", "prenatal_sexDE"),
                                               labels = c("Adult\nPseudotime-DE", "Adult\nSex-DE", "Prenatal\nPseudotime-DE", "Prenatal\nSex-DE"))

# extract and transform the results for the random set of genes
enrichRes_random_formated_ctrlExpr <- enrichRes_formated_ctrlExpr[, c("geneList", "analysis", "direction", "dataset", "pval_enrich_random", "pval_deple_random", "relative_enrich_random")] %>% separate_longer_delim(cols = c(pval_enrich_random, pval_deple_random, relative_enrich_random), delim = ";")
enrichRes_random_formated_ctrlExpr$pval_enrich_random <- as.numeric(enrichRes_random_formated_ctrlExpr$pval_enrich_random)
enrichRes_random_formated_ctrlExpr$pval_deple_random <- as.numeric(enrichRes_random_formated_ctrlExpr$pval_deple_random)
enrichRes_random_formated_ctrlExpr$relative_enrich_random <- as.numeric(enrichRes_random_formated_ctrlExpr$relative_enrich_random)
```


```{r plot_ctrlExpr, echo=FALSE}
p <- ggplot(enrichRes_random_formated_ctrlExpr, aes(y = geneList, x = relative_enrich_random, fill = direction, alpha = 0.4))
p <- p + geom_violin(linewidth = 0.1)
p <- p + geom_vline(xintercept = 1)
p <- p + scale_fill_manual(values = c("female" = "darkred", "male" = "darkblue", "down" = "chocolate", "up" = "aquamarine4"))
p <- p + geom_point(data = enrichRes_formated_ctrlExpr, aes(y = geneList, x = relative_enrich, color = direction), 
                    position = position_dodge(0.9), size = 2, alpha = 1)
p <- p + geom_text(data = enrichRes_formated_ctrlExpr, aes(y = geneList, x = ifelse(relative_enrich>1, relative_enrich + 0.3, relative_enrich - 0.3), label = signif),
                   position = position_dodge(0.9), size = 4, alpha = 1)
p <- p + scale_color_manual(values = c("female" = "darkred", "male" = "darkblue", "down" = "chocolate", "up" = "aquamarine4"))
p <- p + facet_grid(cols = vars(analysis), scales = "free", space = "free")
p <- p + theme_minimal()
p <- p + ylab("") + xlab("Relative enrichment")
p <- p + guides(alpha="none")
p <- p + coord_cartesian(xlim = c(0,5))
p <- p + scale_y_discrete(limits = rev(levels(enrichRes_random_formated_ctrlExpr$geneList))) # reverse order of phenotype
p
# save plot
ggsave("enrichment_plot_sexDE_pseudotimeDE_adult_prenatal_ctrlExpr.png",
      path = "images/5-postprocessing/8-CaseControlEnrichment/hyper/",
      units = "cm", width = 20, height = 12)
```


## Plot for bayes results controlling for gene level of expression

We plot the results for all phenotypes with both direction (male/female) in the prenatal-specific and shared sex-DE genes.

First, we format a bit more the data. The chosen threshold to consider a result as significant is 0.05/6 datasets * 2 directions * 10 phenotypes (~120 analyses) = 0.0004.

```{r format4plot_bayes_ctrlExpr, echo=FALSE}
enrichRes_formated_bayes_ctrlExpr <- enrichRes[which(enrichRes$analysis %in% c("prenatal_specific", "shared") & enrichRes$direction != "both" & enrichRes$ctrlExpr), ]
# add * for signif enrichment or depletion (with a threshold of 0.05/120=0.0004)
enrichRes_formated_bayes_ctrlExpr$signif <- ifelse(enrichRes_formated_bayes_ctrlExpr$relative_enrich > 1, 
                                                   ifelse(enrichRes_formated_bayes_ctrlExpr$pval_enrich<0.0004 & enrichRes_formated_bayes_ctrlExpr$permut_pval < 0.0004, "*", ""),
                                                   ifelse(enrichRes_formated_bayes_ctrlExpr$pval_deple<0.0004 & enrichRes_formated_bayes_ctrlExpr$permut_pval < 0.0004, "*", ""))
# order the different direction of effect
enrichRes_formated_bayes_ctrlExpr$direction <- factor(enrichRes_formated_bayes_ctrlExpr$direction,
                                                      levels = c("female", "male"))
# order the different phenotype
enrichRes_formated_bayes_ctrlExpr$geneList <- factor(enrichRes_formated_bayes_ctrlExpr$geneList,
                                                     levels = c("sczDown", "sczUp", "asdDown", "asdUp", "asdM12", "asdM16", 
                                                                "mddDownFemale", "mddUpFemale", "mddDownMale", "mddUpMale"),
                                                     labels = c("Schizophrenia \ndown", "Schizophrenia \nup", "Autism \ndown", "Autism \nup",
                                                                "Autism \ndown M12", "Autism \nup M16",
                                                                "Major Depressive Disorder \ndown female", "Major Depressive Disorder \nup female",
                                                                "Major Depressive Disorder \ndown male", "Major Depressive Disorder \nup male"))
# rename the 4 analyses
enrichRes_formated_bayes_ctrlExpr$analysis <- factor(enrichRes_formated_bayes_ctrlExpr$analysis,
                                                     levels = c("prenatal_specific", "shared"),
                                                     labels = c("Prenatal\nspecific", "Shared"))

# extract and transform the results for the random set of genes
enrichRes_random_formated_bayes_ctrlExpr <- enrichRes_formated_bayes_ctrlExpr[, c("geneList", "analysis", "direction", "pval_enrich_random", "pval_deple_random", "relative_enrich_random")] %>% separate_longer_delim(cols = c(pval_enrich_random, pval_deple_random, relative_enrich_random), delim = ";")
enrichRes_random_formated_bayes_ctrlExpr$pval_enrich_random <- as.numeric(enrichRes_random_formated_bayes_ctrlExpr$pval_enrich_random)
enrichRes_random_formated_bayes_ctrlExpr$pval_deple_random <- as.numeric(enrichRes_random_formated_bayes_ctrlExpr$pval_deple_random)
enrichRes_random_formated_bayes_ctrlExpr$relative_enrich_random <- as.numeric(enrichRes_random_formated_bayes_ctrlExpr$relative_enrich_random)
```


```{r plot_bayes_ctrlExpr, echo=FALSE}
p <- ggplot(enrichRes_random_formated_bayes_ctrlExpr, aes(y = geneList, x = relative_enrich_random, fill = direction, alpha = 0.4))
p <- p + geom_violin(linewidth = 0.1)
p <- p + geom_vline(xintercept = 1)
p <- p + scale_fill_manual(values = c("female" = "darkred", "male" = "darkblue"))
p <- p + geom_point(data = enrichRes_formated_bayes_ctrlExpr, aes(y = geneList, x = relative_enrich, color = direction), 
                    position = position_dodge(0.9), size = 2, alpha = 1)
p <- p + geom_text(data = enrichRes_formated_bayes_ctrlExpr, aes(y = geneList, x = ifelse(relative_enrich>1, relative_enrich + 0.3, relative_enrich - 0.3), label = signif),
                   position = position_dodge(0.9), size = 4, alpha = 1)
p <- p + scale_color_manual(values = c("female" = "darkred", "male" = "darkblue"))
p <- p + facet_grid(cols = vars(analysis), scales = "free", space = "free")
p <- p + theme_minimal()
p <- p + ylab("") + xlab("Relative enrichent")
p <- p + guides(alpha="none")
p <- p + coord_cartesian(xlim = c(0,7))
p <- p + scale_y_discrete(limits = rev(levels(enrichRes_random_formated_bayes_ctrlExpr$geneList))) # reverse order of phenotype
p
# save plot
ggsave("enrichment_plot_sexDE_bayes_ctrlExpr.png",
      path = "images/5-postprocessing/8-CaseControlEnrichment/hyper/",
      units = "cm", width = 15, height = 16)
```


# Session info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

