---
title: "Create disease gene lists"
author: "Clara Benoit-Pilven"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
---


```{r load_library, echo=FALSE}
suppressPackageStartupMessages( library(readxl) )
suppressPackageStartupMessages( library(tibble) )
suppressPackageStartupMessages( library(dplyr) )
suppressPackageStartupMessages( library(genefilter) )
suppressPackageStartupMessages( library(biomaRt) )
suppressPackageStartupMessages( library(readxl) )
```

```{r, setup, include=FALSE, echo=FALSE}
knitr::opts_knit$set(root.dir = '~/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/')
```

# Load data

Load DE results.

```{r load_DE_results_hdbr, echo=FALSE}
de_results_hdbr <- read.table("data/3-DE/VoomDream_topTable_forebrain_pseudotime_sva_ind_interaction.txt",
                         header = TRUE, check.names=FALSE)
# keep only autosomal genes
de_results_hdbr <- de_results_hdbr[which(!de_results_hdbr$chr %in% c("chrX", "chrY", "chrM")),]
# remove the version number from the ensembl id
de_results_hdbr$gene_id <-sapply(strsplit(as.character(de_results_hdbr$gene_id), "\\."), function(x){x[1]})
```

```{r load_DE_results_gtex, echo=FALSE}
de_results_gtex <- read.table( "data/4-GTEx/1-GTEx_analysis/brain_forebrain_v8_VoomDream_topTable_with_sva_ind_pseudotime_interaction.txt",
                         header = TRUE, check.names=FALSE)
# keep only autosomal genes
de_results_gtex <- de_results_gtex[which(!de_results_gtex$chr %in% c("chrX", "chrY", "chrM")),]
# remove the version number from the ensembl id
de_results_gtex$gene_id <-sapply(strsplit(as.character(de_results_gtex$gene_id), "\\."), function(x){x[1]})
```

Function to check if the selected genes are present in the list of analyzed genes (and not in the sex chromosomes).

```{r check_genes_present, echo=FALSE}
getOnlyGenesAnalyzed <- function(x, genesAnalyzed){
  return(x[which(x %in% genesAnalyzed)])
}
```

Initisalize biomart connection.

```{r init_biomart, echo=FALSE}
ensembl <- useEnsembl(biomart="ensembl", dataset="hsapiens_gene_ensembl", version = "98")
```


# Prepare genesets

Initialize gene lists

```{r init, echo=FALSE}
genesLists <- list()
```

## Schizophrenia

Load data from Clifton et al., 2023.

```{r load_scz, echo=FALSE}
scz_gene_list <- read_excel("data/5-postprocessing/8-CaseControlEnrichment/geneLists/Clifton2023_SCZ_ajmgb32931-sup-0001-tables.xlsx",
                            sheet = 2)
```

Select only the schizophrenia gene with a p-value less than 0.05 and split the up and down regulated genes.

```{r select_scz, echo=FALSE}
scz_gene_selected_up <- scz_gene_list$ensembl_ID[which(scz_gene_list$adj.P.Val < 0.05 & scz_gene_list$logFC > 0)] # genes upregulated in SCZ cases
scz_gene_selected_down <- scz_gene_list$ensembl_ID[which(scz_gene_list$adj.P.Val < 0.05 & scz_gene_list$logFC < 0)] # genes downregulated in SCZ cases
```

Keep only genes that were also analyzed in our sex-DE analysis?

```{r filter_scz, echo=FALSE, eval=FALSE}
scz_gene_selected_hdbr <- scz_gene_selected[which(scz_gene_selected %in% de_results_hdbr$gene_id)]
scz_gene_selected_gtex <- scz_gene_selected[which(scz_gene_selected %in% de_results_gtex$gene_id)]
```

Add to gene list.

```{r add_scz, echo=FALSE}
genesLists$sczUp <- scz_gene_selected_up
genesLists$sczDown <- scz_gene_selected_down
```


## ASD

Load data from Werling et al., 2016.

```{r load_asd, echo=FALSE}
asdUp_gene_list <- read_excel("data/5-postprocessing/8-CaseControlEnrichment/geneLists/Werling2016_ASD_41467_2016_BFncomms10717_MOESM1122_ESM.xlsx",
                            sheet = 5, skip = 1)
asdDown_gene_list <- read_excel("data/5-postprocessing/8-CaseControlEnrichment/geneLists/Werling2016_ASD_41467_2016_BFncomms10717_MOESM1122_ESM.xlsx",
                            sheet = 6, skip = 1)
asdM12_gene_list <- read_excel("data/5-postprocessing/8-CaseControlEnrichment/geneLists/Werling2016_ASD_41467_2016_BFncomms10717_MOESM1122_ESM.xlsx",
                            sheet = 7, skip = 1)
asdM16_gene_list <- read_excel("data/5-postprocessing/8-CaseControlEnrichment/geneLists/Werling2016_ASD_41467_2016_BFncomms10717_MOESM1122_ESM.xlsx",
                            sheet = 8, skip = 1)
```

Add to gene list.

```{r add_asd, echo=FALSE}
genesLists$asdUp <- asdUp_gene_list$ensembl_gene_id
genesLists$asdDown <- asdDown_gene_list$ensembl_gene_id
genesLists$asdM12 <- asdM12_gene_list$ensembl_gene_id
genesLists$asdM16 <- asdM16_gene_list$ensembl_gene_id
```


## MDD

Load data from Maitra et al., 2023.

```{r load_mdd, echo=FALSE}
mdd_males_gene_list <- read_excel("data/5-postprocessing/8-CaseControlEnrichment/geneLists/Maitra2023_MDD_41467_2023_38530_MOESM3_ESM.xlsx",
                            sheet = 5, skip = 2)
mdd_females_gene_list <- read_excel("data/5-postprocessing/8-CaseControlEnrichment/geneLists/Maitra2023_MDD_41467_2023_38530_MOESM3_ESM.xlsx",
                            sheet = 6, skip = 2)
```

For each sex, get the up and down regulated genes.

```{r filter_mdd, echo=FALSE}
mdd_males_gene_list_up <- mdd_males_gene_list$gene[which(mdd_males_gene_list$p_adj.glb < 0.05 & mdd_males_gene_list$logFC > 0)]
mdd_males_gene_list_down <- mdd_males_gene_list$gene[which(mdd_males_gene_list$p_adj.glb < 0.05 & mdd_males_gene_list$logFC < 0)]
mdd_females_gene_list_up <- mdd_females_gene_list$gene[which(mdd_females_gene_list$p_adj.glb < 0.05 & mdd_females_gene_list$logFC > 0)]
mdd_females_gene_list_down <- mdd_females_gene_list$gene[which(mdd_females_gene_list$p_adj.glb < 0.05 & mdd_females_gene_list$logFC < 0)]
```

Convert gene name to ensemblID.

```{r convert_geneName_mdd, echo=FALSE}
mdd_males_gene_list_up_ensemblID <- getBM(filters = "hgnc_symbol", attributes = c("ensembl_gene_id"), 
                                          values = mdd_males_gene_list_up, mart = ensembl)[,1]
mdd_males_gene_list_down_ensemblID <- getBM(filters = "hgnc_symbol", attributes = c("ensembl_gene_id"), 
                                            values = mdd_males_gene_list_down, mart = ensembl)[,1]
mdd_females_gene_list_up_ensemblID <- getBM(filters = "hgnc_symbol", attributes = c("ensembl_gene_id"), 
                                            values = mdd_females_gene_list_up, mart = ensembl)[,1]
mdd_females_gene_list_down_ensemblID <- getBM(filters = "hgnc_symbol", attributes = c("ensembl_gene_id"), 
                                              values = mdd_females_gene_list_down, mart = ensembl)[,1]
```

Add to gene list.

```{r add_mdd, echo=FALSE}
genesLists$mddUpFemale <- mdd_females_gene_list_up_ensemblID
genesLists$mddDownFemale <- mdd_females_gene_list_down_ensemblID
genesLists$mddUpMale <- mdd_males_gene_list_up_ensemblID
genesLists$mddDownMale <- mdd_males_gene_list_down_ensemblID
```


# Save gene list

```{r save_geneList, echo=FALSE}
save(genesLists, file = "data/5-postprocessing/8-CaseControlEnrichment/geneLists/caseCtrlGeneLists.RData")
```


# Some statistics of the gene sets

```{r stat_geneList, echo=FALSE}
unlist(lapply(genesLists, length))
```


# Session info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
