---
title: "sex-DE dream analysis on HDBR forebrain samples using pseudotime bins as covariate"
author: "Clara Benoit-Pilven"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
---

```{r load_libraries, echo=FALSE}
suppressPackageStartupMessages( library(edgeR) )
suppressPackageStartupMessages( library(limma) )
suppressPackageStartupMessages( library(qvalue) )
suppressPackageStartupMessages( library(sva) )
suppressPackageStartupMessages( library(doParallel) )
suppressPackageStartupMessages( library(variancePartition) )
```

```{r parallel_processing_param, echo=FALSE}
# Specify parallel processing parameters
# this is used implicitly by dream() to run in parallel
param <- SnowParam(4, "SOCK", progressbar=TRUE)
register(param)
```

```{r setup, echo=FALSE}
knitr::opts_knit$set(root.dir = '~/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/')
```

# Sex-DE analysis on genes with pseudotime bins covariate

## Load the counts and the phenotype data

```{r load_data, echo=FALSE}
load("data/1-preprocessing/counts_forebrain_phenotypes.RData")
load("data/2-pseudotime/forebrain_metadata_pseudotime.RData")
metadata <- metadata[,c(1,6,3,2)]
```

## Create bins of pseudotime

We create the bins of pseudotime. The number of bins is the same as the number of developmental stages and the number of samples per bins is equal.

```{r pseudotime_bins, echo=FALSE}
nbrBins <- length(levels(metadata$DevStage))
metadata$pseudotimeBins <- cut(metadata$pseudotime, 
                               quantile(metadata$pseudotime, seq(0,1,1/nbrBins)), 
                               labels=c(1:nbrBins), include.lowest=T)
metadata$pseudotimeBins <- as.factor(metadata$pseudotimeBins)
```

Check the correlation between the pseudotime bins and the developmental stages.

```{r plot_pseudotimeBins_vs_devStages, echo=FALSE}
# x = bin pseudotime
p <- ggplot(metadata, aes(x = pseudotimeBins, fill = DevStage))
p <- p + geom_bar(position = "fill")
p <- p + labs(y = "Proportion")
p <- p + theme_minimal()
p
# x = dev stages
p <- ggplot(metadata, aes(x = DevStage, fill = pseudotimeBins))
p <- p + geom_bar(position = "fill")
p <- p + labs(y = "Proportion")
p <- p + theme_minimal()
p
```



## Prepare design

```{r prepare_metadata, echo=FALSE}
# design for voomDream (the variable to be tested must be a fixed effect)
form <- ~ Sex + pseudotimeBins + Sex:pseudotimeBins + (1|Individual)
# design for sva computation
design <- model.matrix(~metadata$Sex + metadata$pseudotimeBins + metadata$Sex:metadata$pseudotimeBins)
```

## Prepare counts for DE analysis

Estimate weights using linear mixed model of dream

```{r voom_norm, echo=FALSE, warning=FALSE}
png('images/3-DE/analysis_pseudotimeBins/trendplot/trendplot_VoomDream_forebrain_pseudotimeBins_ind_interaction.png')
suppressWarnings( vobjDream <- voomWithDreamWeights(x, form, metadata, plot = TRUE) )
dev.off()
```

Estimate surrogates variables (svas) and add them to the model.

```{r estimate_sva, echo=FALSE}
n.sv <- num.sv(vobjDream$E, design, method = "be")
sva.obj <- sva(vobjDream$E, design, method = "irw", n.sv = n.sv)
# add svas to metadata
metadata <- cbind(metadata, sva.obj$sv)
names(metadata) <- c("Sex", "pseudotime", "Individual", "DevStage", "pseudotimeBins", paste0("X", 1:n.sv))
# include svas in formula
form <- reformulate(c("Sex", "(1|pseudotimeBins)", "Sex:pseudotimeBins", "(1|Individual)", paste0("X", 1:n.sv)))
```

Redo voom normalization with new design matrix (containing svas).

```{r voom_norm_svas, echo=FALSE, warning=FALSE}
suppressWarnings( y <- voomWithDreamWeights(x, form, metadata, plot = TRUE) )
```

Save the normalized counts and the phenotypes vectors in a RData file.

```{r save_normCounts, echo=FALSE}
save(y, phenotypes, form, sva.obj, metadata,
     file = "data/3-DE/analysis_pseudotimeBins/normCounts_dream_pseudotimeBins_sva_ind_interaction_phenotypes.RData")
```


## Sex-Differential Expression (sex-DE) analysis

Fit the dream model on each gene.

```{r sexDE_analysis, echo=FALSE}
# By default, uses the Satterthwaite approximation for the hypothesis test
suppressWarnings( fitmm <- dream(y, form, metadata) )
fitmm2 <- topTable(fitmm, coef = 2, number = 30000, sort.by = "none") 
fitmm2$qvalue <- qvalue(fitmm2$P.Value)$qvalues
```

Save the fitmm object.
```{r save_fitmm, echo=FALSE}
save(fitmm, file = "data/3-DE/analysis_pseudotimeBins/fitmm_dream_pseudotimeBins_sva_ind_interaction_phenotypes.RData")
```

Compute the mean expression of each genes for male and female.

```{r compute_mean_expr, echo=FALSE}
fitmm2$Male <- rowMeans(y$E[,which(metadata$Sex==1)]) # for male samples
fitmm2$VarMale <- rowVars(y$E[,which(metadata$Sex==1)])
fitmm2$Female <- rowMeans(y$E[,which(metadata$Sex==2)]) # for female samples
fitmm2$VarFemale <- rowVars(y$E[,which(metadata$Sex==2)])
```

Save results.

```{r write_output, echo=FALSE}
write.table(fitmm2, 
            "data/3-DE/analysis_pseudotimeBins/VoomDream_topTable_forebrain_pseudotimeBins_sva_ind_interaction.txt", 
            col.names = TRUE, row.names = FALSE, quote = FALSE)
```


# Session Info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

