---
title: "Replication of results with other datasets"
author: "Clara Benoit-Pilven"
date: "`r Sys.Date()`"
output: html_document
---

```{r load_library, echo=FALSE}
suppressPackageStartupMessages( library(edgeR) )
suppressPackageStartupMessages( library(limma) )
suppressPackageStartupMessages( library(qvalue) )
suppressPackageStartupMessages( library(tidyverse) )
suppressPackageStartupMessages( library(readxl))
```

```{r, setup, include=FALSE, echo=FALSE}
knitr::opts_knit$set(root.dir = '~/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/')
```

```{r parameters, echo=FALSE}
qvalue_threshold <- 0.01
logFC_threshold <- 0
```

# Load data

## Our own data

Load HDBR sex-DE data.

```{r load_data_HDBR_sexDE, echo=FALSE}
hdbr_results <- read.table("data/3-DE/VoomDream_topTable_forebrain_pseudotime_sva_ind_interaction.txt",
                           header = TRUE, check.names=FALSE)
# format
hdbr_results$ensemblID <- sapply(strsplit(as.character(hdbr_results$gene_id), "\\."), function(x){x[1]})
# autosomal genes
sexChr <- c("chrY", "chrX")
hdbr_results_auto <- hdbr_results[which(!hdbr_results$chr %in% sexChr), ]
```

## O'Brien 2018 data

Load the data.

```{r load_Obrien_data, echo=FALSE}
OBrien_results <- read.table("data/5-postprocessing/9-Replication/datasets/OBrien2018_SEXdiffs.txt",
                          header = TRUE)
# rename some columns to match our naming system
OBrien_results$ensemblID <- OBrien_results$Id
OBrien_results$qvalue <- OBrien_results$padj
OBrien_results$P.Value <- OBrien_results$pvalue
OBrien_results$logFC <- OBrien_results$log2FoldDiff
# the logFC in the O'Brien data is in the opposite direction than in our data
# modify it to make it match our data 
OBrien_results$logFC <- -OBrien_results$logFC
# autosomal genes
chrToRemove <- c("chrY", "chrX", "chrMT")
OBrien_results_auto <- OBrien_results[which(!OBrien_results$Chr %in% chrToRemove), ]
```

## Kissel 2024 data

Load the data.

```{r load_Kissel_data, echo=FALSE}
Kissel_results <- read_excel("data/5-postprocessing/9-Replication/datasets/Kissel2024_SuppTable4_MainModel_DifferentialExpression_andMetaAnalysis.xlsx",
                              sheet = 2)
# separate the UCLA and BV data
Kissel_UCLA_results <- Kissel_results[, c(1:20)]
Kissel_BV_results <- Kissel_results[, c(1:5, 21:35)]
# rename some columns to match our naming system
Kissel_UCLA_results$logFC <- Kissel_UCLA_results$logFC_UCLA
Kissel_UCLA_results$qvalue <- Kissel_UCLA_results$adj.P.Val_UCLA
Kissel_UCLA_results$P.Value <- Kissel_UCLA_results$P.Value_UCLA
Kissel_UCLA_results$ensemblID <- Kissel_UCLA_results$EnsemblID
Kissel_BV_results$logFC <- Kissel_BV_results$logFC_BV
Kissel_BV_results$qvalue <- Kissel_BV_results$adj.P.Val_BV
Kissel_BV_results$P.Value <- Kissel_BV_results$P.Value_BV
Kissel_BV_results$ensemblID <- Kissel_BV_results$EnsemblID
# the logFC in the Kissel data is in the opposite direction than in our data
# modify it to make it match our data 
Kissel_UCLA_results$logFC <- -Kissel_UCLA_results$logFC
Kissel_BV_results$logFC <- -Kissel_BV_results$logFC
# autosomal genes
chrToRemove <- c("chrY", "chrX", "chrM")
Kissel_UCLA_results_auto <- Kissel_UCLA_results[which(!Kissel_UCLA_results$Chr %in% chrToRemove), ]
Kissel_BV_results_auto <- Kissel_BV_results[which(!Kissel_BV_results$Chr %in% chrToRemove), ]
```



# Compare results

For each dataset, we compare our data using several different metrics:
- percentage of significant genes from the dataset that overlap our significant genes
- replication rate of our significant sex-DE genes in the dataset (pi1 signif)
- effect size (logFC) correlation for our autosomal significant sex-DE genes
- effect size (logFC) consistency of our significant sex-DE genes in the dataset


## HDBR vs O'Brien data

The O'Brien et al., 2018 preprint analyze 120 samples (from 50 female individuals and 70 male individuals) from the second trimester of pregnancy (12-19 post-conception weeks).
From the manuscript, the authors found 1468 male-biased and 1288 female-biased genes (with FDR < 0.1).

### Data summary

Get the number of sex-DE genes with adjusted p-value < 0.01 (same threshold we used in our analysis).

```{r get_stats_OBrien, echo=FALSE}
nbr_signif <- sum(OBrien_results$qvalue < qvalue_threshold)
nbr_signif
```

### Overlap

Get the number of significant genes overlapping between the 2 datasets.

```{r permut_test_overlap_function, echo=FALSE}
overlap_signif <- function(data1, data2, columnMerge = "ensemblID", qvalueThreshold = 0.01, nbrRandom = 1000){
  # merge the 2 datasets
  de_results_merged <- merge(data1, data2, by = columnMerge, suffixes = c(".data1", ".data2"))
  # get the number of significant genes in each data
  nbr_signif_data1 <- length(de_results_merged[which(de_results_merged$qvalue.data1 < qvalueThreshold), columnMerge])
  nbr_signif_data2 <- length(de_results_merged[which(de_results_merged$qvalue.data2 < qvalueThreshold), columnMerge])
  print(sprintf("Number of genes significant in data1 = %s", nbr_signif_data1))
  print(sprintf("Number of genes significant in data2 = %s", nbr_signif_data2))
  # keep only the genes/transcripts that are significant in the 2 datasets
  de_results_merged_filtered <- de_results_merged[which(de_results_merged$qvalue.data1 < qvalueThreshold & de_results_merged$qvalue.data2 < qvalueThreshold), ]
  nbr_common_signif <- dim(de_results_merged_filtered)[1]
  print(sprintf("Number of genes significant in both dataset = %s", nbr_common_signif))
  print(sprintf("Percentage of significant genes from data1 also significant in data2 = %s", nbr_common_signif/nbr_signif_data1*100))
  # Compare this number to the number of overlaping genes if we take randomly 2 lists of the same size (nbr_sgnif_data1 & nbr_sgnif_data2)
  nbr_signif_values <- c()
  # to do nbrRandom times
  for (i in 1:nbrRandom){
    # randomly select "nbr_sgnif_data1" genes/transcripts in the data1
    selection_data1 <- sample(de_results_merged[, columnMerge], nbr_signif_data1)
    # randomly select "nbr_sgnif_data2" genes/transcripts in the data2
    selection_data2 <- sample(de_results_merged[, columnMerge], nbr_signif_data2)
    nbr_signif_th <- sum(selection_data1 %in% selection_data2)
    nbr_signif_values[i] <- nbr_signif_th
  }
  nbr_signif_values <- as.data.frame(nbr_signif_values)
  # plot the distribution of the number of overlaping genes obtained and the real value
  p <- ggplot(nbr_signif_values, aes(nbr_signif_values))
  p <- p + geom_histogram(binwidth=1, boundary = 0, fill="grey", color="black")
  p <- p + theme_bw() + xlab("Number of overlaping genes") + ylab("Frequency")
  p <- p + geom_vline(xintercept = nbr_common_signif, color = "red")
  return(p)
}
```

```{r permut_test_overlap_OBrien, echo=FALSE}
p <- overlap_signif(OBrien_results, hdbr_results, qvalueThreshold = qvalue_threshold)
p
```

### Replication rate

We compute the replication rate (pi1).

```{r function_pi1_score, echo=FALSE}
pi1_score <- function(data1, data2, columnMerge = "ensemblID"){
  # compute pi1 for second dataset compared to the first dataset
  selected_pvalue <- data2$P.Value[which(pull(data2, columnMerge) %in% pull(data1, columnMerge))]
  if (berryFunctions::is.error(qvalue::pi0est(p = selected_pvalue, lambda = 0.5, pi0.method = "smoother")) ) {
    pi1 <- qvalue::qvalue(p = selected_pvalue, pi0 = 1)$pi0
  }else{
    pi1 <- 1-qvalue::pi0est(p = selected_pvalue, lambda = 0.5, pi0.method = "smoother")$pi0
  }
  
  # continue only if the pi1 value was computed
  if (is.numeric(pi1)){
    return(pi1)
  } else{
    return("No pi1 analysis!")
  }
}
```

```{r function_pi1_normalized, echo=FALSE}
pi1_normalized <- function(data1, data2, columnMerge = "ensemblID", qvalue_threshold = 0.01, logFC_threshold = 0, verbose = TRUE){
  # get the significant genes in the first dataset
  data1_signif <- data1[which(data1$qvalue < qvalue_threshold & abs(data1$logFC) > logFC_threshold), ]
  
  # pi1 replication rate for the significant genes
  pi1_signif <- pi1_score(data1_signif, data2, columnMerge = "ensemblID")
  if (verbose){
    print(sprintf("pi1 for signif genes = %s", pi1_signif))
  }
  # pi1 replication rate for all genes
  pi1_all <- pi1_score(data1, data2, columnMerge = "ensemblID")
  if (verbose){
    print(sprintf("pi1 for all genes = %s", pi1_all))
  }
  
  # return the pi1 for significant genes normalized by the pi1 for all genes
  return(pi1_signif/pi1_all)
}
```

```{r pi1_score_OBrien, echo=FALSE}
print("Replication of HDBR sex-DE results in O'Brien sex-DE analysis:")
pi1_normalized(hdbr_results, OBrien_results, qvalue_threshold = qvalue_threshold)
print("Replication of O'Brien sex-DE results in HDBR sex-DE analysis:")
pi1_normalized(OBrien_results, hdbr_results, qvalue_threshold = qvalue_threshold)
```

### Effect size correlation

Scatter plot of the logFC between the 2 analyses with information about the correlation (r and p-value of cor.test).

```{r function_scatter_plot_logFC, echo=FALSE}
scatter_logFC <- function(data1, data2, nameData1, nameData2, pointColor, columnMerge = "ensemblID"){
  # merge the 2 dataset
  de_results_merged <- merge(data1, data2, by = columnMerge, suffixes = c(".data1", ".data2"))
  # test correlation
  res <- cor.test(de_results_merged$logFC.data1, de_results_merged$logFC.data2)
  if (res$p.value == 0){
    annotation_text <- sprintf("r = %.3f\nP < 2.2e-16", res$estimate)
  } else{
    annotation_text <- sprintf("r = %.3f\nP = %1.2e", res$estimate, res$p.value)
  }
  yrng <- range(de_results_merged$logFC.data1)
  xrng <- range(de_results_merged$logFC.data2)
  # plot
  p <- ggplot(de_results_merged, aes(x = logFC.data1, y = logFC.data2))
  p <- p + geom_abline(intercept = 0, slope = 1, col = "black")
  p <- p + geom_hline(yintercept = 0, color = "grey")
  p <- p + geom_vline(xintercept = 0, color = "grey")
  p <- p + geom_point(color=pointColor, alpha = 0.3)
  p <- p + theme_bw() + labs(title = "")
  p <- p + theme(panel.grid = element_blank(), 
                 plot.title = element_text(hjust = 0.5),
                 plot.subtitle = element_text(hjust = 0.5))
  p <- p + xlab(paste0("logFC (", nameData1, ")")) + ylab(paste0("logFC (", nameData2, ")"))
  p <- p + annotate("text", xrng[1], yrng[2], hjust = 0, vjust = 1, 
                    label = annotation_text)
  return(p)
}
```

Correlation plot only of HDBR sex-DE genes.

```{r corr_pseudo_sexDE_OBrien, echo=FALSE}
p <- scatter_logFC(hdbr_results[which(hdbr_results$qvalue < qvalue_threshold),], OBrien_results, "HDBR", "OBrien", 
                   pointColor = "black")
p
```

Correlation plot only of autosomal HDBR sex-DE genes.

```{r corr_pseudo_sexDE_auto_OBrien, echo=FALSE}
p <- scatter_logFC(hdbr_results_auto[which(hdbr_results_auto$qvalue < qvalue_threshold),], OBrien_results, "HDBR", "OBrien", 
                   pointColor = "black")
p
```

Correlation plot only for O'Brien sex-DE genes.

```{r corr_OBrien_sexDE_pseudo, echo=FALSE}
p <- scatter_logFC(OBrien_results[which(OBrien_results$qvalue < qvalue_threshold),], hdbr_results, "OBrien", "HDBR", 
                   pointColor = "black")
p
```

Correlation plot only for O'Brien autosomal sex-DE genes.

```{r corr_OBrien_sexDE_pseudo_auto, echo=FALSE}
p <- scatter_logFC(OBrien_results_auto[which(OBrien_results_auto$qvalue < qvalue_threshold),], hdbr_results, "OBrien", "HDBR", 
                   pointColor = "black")
p
```



### Effect size consistency

Compute the consistency of log fold change.

```{r function_consistency, echo=FALSE}
consistency_test <- function(data1, data2, columnMerge = "ensemblID", qvalueThreshold = 0.01){
  # merge the 2 datasets
  de_results_merged <- merge(data1, data2, by = columnMerge, suffixes = c(".data1", ".data2"))
  # keep only the genes/transcripts that are significant in the first data
  de_results_merged_filtered <- de_results_merged[which(de_results_merged$qvalue.data1 < qvalueThreshold), ]
  print(sprintf("Number of genes compared = %s", dim(de_results_merged_filtered)[1]))
  # number of consistent genes/transcripts
  nbr_consistent <- sum(sign(de_results_merged_filtered$logFC.data1) == sign(de_results_merged_filtered$logFC.data2))
  consistency <- nbr_consistent/length(de_results_merged_filtered[, columnMerge])
  # compare this consistency score with the consistency of list of random genes
  cons_values <- rep(NA, 1000)
  # to do 1000 times
  for (i in 1:1000){
    # randomly select the same number of genes/transcripts in the data2 to test the overlap with the data1 signif genes
    selection <- sample(de_results_merged[, columnMerge], length(de_results_merged_filtered[, columnMerge]))
    nbr_consistent_th <- sum(sign(de_results_merged$logFC.data2[which(de_results_merged[, columnMerge] %in% selection)]) == sign(de_results_merged$logFC.data1[which(de_results_merged[, columnMerge] %in% selection)]))
    cons_values[i] <- nbr_consistent_th/length(de_results_merged_filtered[, columnMerge])
  }
  cons_values <- as.data.frame(cons_values)
  # plot the distribution of the consistency obtained and the real value
  p <- ggplot(cons_values, aes(cons_values))
  p <- p + geom_histogram(binwidth=0.02, boundary = 0, fill="grey", color="black")
  p <- p + theme_bw() + xlab("consistency") + ylab("Frequency")
  p <- p + scale_x_continuous(limits = c(0, 1))
  p <- p + geom_vline(xintercept = consistency, color = "red")
  print(p)
  # test this consistency with a permutation test
  nsmaller <- sum(cons_values >= consistency)
  permutation_test <- round((nsmaller+1)/(1000+1), 3)
  # test this consistency with a binomial test
  result_test <- binom.test(nbr_consistent, length(de_results_merged_filtered[, columnMerge]), p = 0.5)
  # print results
  print(sprintf("number of consistent gene = %s", nbr_consistent))
  print(sprintf("consistency = %s", consistency))
  print(sprintf("permutation test p-value = %s", permutation_test))
  print(sprintf("binomial test p-value = %s", result_test$p.value))
  return(consistency)
}
```

```{r consistency_score_OBrien_vs_HDBR, echo=FALSE}
print("Consistency of OBrien analysis compared to the results of HDBR analysis:")
consistency_test(hdbr_results, OBrien_results, qvalueThreshold = 0.01)
```




## HDBR vs Kissel BrainVar data

The Kissel et al., 2014 BrainVar data contains 87 samples (41 males, 46 females) from 14 to 21 PCW from DLPFC.
For the sex-DE analysis, the age was used as a covariate.

### Data summary

Get the number of sex-DE genes with adjusted p-value < 0.01 (same threshold we used in our analysis).

```{r get_stats_BV, echo=FALSE}
nbr_signif_BV <- sum(Kissel_BV_results$qvalue < qvalue_threshold)
nbr_signif_BV
```


### Overlap

Get the number of significant genes overlapping between the 2 datasets.

```{r permut_test_overlap_BV, echo=FALSE}
p <- overlap_signif(Kissel_BV_results, hdbr_results, qvalueThreshold = qvalue_threshold)
p
```

### Replication rate

We compute the replication rate (pi1).

```{r pi1_score_BV, echo=FALSE}
print("Replication of HDBR sex-DE results in Kissel BrainVar sex-DE analysis:")
pi1_normalized(hdbr_results, Kissel_BV_results, qvalue_threshold = qvalue_threshold)
print("Replication of Kissel BrainVar sex-DE results in HDBR sex-DE analysis:")
pi1_normalized(Kissel_BV_results, hdbr_results, qvalue_threshold = qvalue_threshold)
```


### Effect size correlation

Correlation plot only of HDBR sex-DE genes.

```{r corr_pseudo_sexDE_BV, echo=FALSE}
p <- scatter_logFC(hdbr_results[which(hdbr_results$qvalue < qvalue_threshold),], Kissel_BV_results, "HDBR", "Kissel_BV", 
                   pointColor = "black")
p
```

Correlation plot only of autosomal HDBR sex-DE genes.

```{r corr_pseudo_sexDE_auto_BV, echo=FALSE}
p <- scatter_logFC(hdbr_results_auto[which(hdbr_results_auto$qvalue < qvalue_threshold),], Kissel_BV_results, "HDBR", "Kissel_BV", 
                   pointColor = "black")
p
```


### Effect size consistency

Compute the consistency of log fold change.

```{r consistency_score_Kissel_BV_vs_HDBR, echo=FALSE}
print("Consistency of Kissel BrainVar analysis compared to the results of HDBR analysis:")
consistency_test(hdbr_results, Kissel_BV_results, qvalueThreshold = 0.01)
```




## HDBR vs Kissel UCLA data

The Kissel et al., 2014 UCLA data contains 186 samples (101 males, 85 females) from 14 to 21 PCW from the whole cortex.
For the sex-DE analysis, the age was used as a covariate.

### Data summary

Get the number of sex-DE genes with adjusted p-value < 0.01 (same threshold we used in our analysis).

```{r get_stats_UCLA, echo=FALSE}
nbr_signif_UCLA <- sum(Kissel_UCLA_results$qvalue < qvalue_threshold)
nbr_signif_UCLA
```

### Overlap

Get the number of significant genes overlapping between the 2 datasets.

```{r permut_test_overlap_UCLA, echo=FALSE}
p <- overlap_signif(Kissel_UCLA_results, hdbr_results, qvalueThreshold = qvalue_threshold)
p
```

### Replication rate

We compute the replication rate (pi1).

```{r pi1_score_UCLA, echo=FALSE}
print("Replication of HDBR sex-DE results in Kissel UCLA sex-DE analysis:")
pi1_normalized(hdbr_results, Kissel_UCLA_results, qvalue_threshold = qvalue_threshold)
print("Replication of Kissel UCLA sex-DE results in HDBR sex-DE analysis:")
pi1_normalized(Kissel_UCLA_results, hdbr_results, qvalue_threshold = qvalue_threshold)
```

### Effect size correlation

Correlation plot only of HDBR sex-DE genes.

```{r corr_pseudo_sexDE_UCLA, echo=FALSE}
p <- scatter_logFC(hdbr_results[which(hdbr_results$qvalue < qvalue_threshold),], Kissel_UCLA_results, "HDBR", "Kissel_UCLA", 
                   pointColor = "black")
p
```

Correlation plot only of autosomal HDBR sex-DE genes.

```{r corr_pseudo_sexDE_auto_UCLA, echo=FALSE}
p <- scatter_logFC(hdbr_results_auto[which(hdbr_results_auto$qvalue < qvalue_threshold),], Kissel_UCLA_results, "HDBR", "Kissel_UCLA", 
                   pointColor = "black")
p
```

### Effect size consistency

Compute the consistency of log fold change.

```{r consistency_score_Kissel_UCLA_vs_HDBR, echo=FALSE}
print("Consistency of Kissel UCLA analysis compared to the results of HDBR analysis:")
consistency_test(hdbr_results, Kissel_UCLA_results, qvalueThreshold = 0.01)
```




## O'Brien vs Kissel BV data

The O'Brien and Kissel BV data both contain samples from the second trimester of pregnancy and use the age as a covariate.


### Overlap

Get the number of significant genes overlapping between the 2 datasets.

```{r permut_test_overlap_OBrien_BV, echo=FALSE}
p <- overlap_signif(Kissel_BV_results, OBrien_results, qvalueThreshold = qvalue_threshold)
p
p <- overlap_signif(OBrien_results, Kissel_BV_results, qvalueThreshold = qvalue_threshold)
p
```


### Replication rate

We compute the replication rate (pi1).

```{r pi1_score_OBrien_BV, echo=FALSE}
print("Replication of O'Brien sex-DE results in Kissel BV sex-DE analysis:")
pi1_normalized(OBrien_results, Kissel_BV_results, qvalue_threshold = qvalue_threshold)
print("Replication of Kissel BV sex-DE results in O'Brien sex-DE analysis:")
pi1_normalized(Kissel_BV_results, OBrien_results, qvalue_threshold = qvalue_threshold)
```


### Effect size correlation

Correlation plot only of O'Brien sex-DE genes.

```{r corr_OBrien_sexDE_BV, echo=FALSE}
p <- scatter_logFC(OBrien_results[which(OBrien_results$qvalue < qvalue_threshold),], Kissel_BV_results, "OBrien", "Kissel_BV", 
                   pointColor = "black")
p
```

Correlation plot only of autosomal O'Brien sex-DE genes.

```{r corr_OBrien_sexDE_auto_BV, echo=FALSE}
p <- scatter_logFC(OBrien_results_auto[which(OBrien_results_auto$qvalue < qvalue_threshold),], Kissel_BV_results, "OBrien", "Kissel_BV", 
                   pointColor = "black")
p
```


### Effect size consistency

Compute the consistency of log fold change.

```{r consistency_score_BV_vs_OBrien, echo=FALSE}
print("Consistency of Kissel BV analysis compared to the results of O'Brien analysis:")
consistency_test(OBrien_results, Kissel_BV_results, qvalueThreshold = 0.01)
```




## O'Brien vs Kissel UCLA data

The O'Brien and Kissel UCLA data both contain samples from the second trimester of pregnancy and use the age as a covariate.

### Overlap

Get the number of significant genes overlapping between the 2 datasets.

```{r permut_test_overlap_OBrien_UCLA, echo=FALSE}
p <- overlap_signif(Kissel_UCLA_results, OBrien_results, qvalueThreshold = qvalue_threshold)
p
p <- overlap_signif(OBrien_results, Kissel_UCLA_results, qvalueThreshold = qvalue_threshold)
p
```


### Replication rate

We compute the replication rate (pi1).

```{r pi1_score_OBrien_UCLA, echo=FALSE}
print("Replication of O'Brien sex-DE results in Kissel UCLA sex-DE analysis:")
pi1_normalized(OBrien_results, Kissel_UCLA_results, qvalue_threshold = qvalue_threshold)
print("Replication of Kissel UCLA sex-DE results in O'Brien sex-DE analysis:")
pi1_normalized(Kissel_UCLA_results, OBrien_results, qvalue_threshold = qvalue_threshold)
```


### Effect size correlation

Correlation plot only of O'Brien sex-DE genes.

```{r corr_OBrien_sexDE_UCLA, echo=FALSE}
p <- scatter_logFC(OBrien_results[which(OBrien_results$qvalue < qvalue_threshold),], Kissel_UCLA_results, "OBrien", "Kissel_UCLA", 
                   pointColor = "black")
p
```

Correlation plot only of autosomal O'Brien sex-DE genes.

```{r corr_OBrien_sexDE_auto_UCLA, echo=FALSE}
p <- scatter_logFC(OBrien_results_auto[which(OBrien_results_auto$qvalue < qvalue_threshold),], Kissel_UCLA_results, "OBrien", "Kissel_UCLA", 
                   pointColor = "black")
p
```


### Effect size consistency

Compute the consistency of log fold change.

```{r consistency_score_UCLA_vs_OBrien, echo=FALSE}
print("Consistency of Kissel UCLA analysis compared to the results of O'Brien analysis:")
consistency_test(OBrien_results, Kissel_UCLA_results, qvalueThreshold = 0.01)
```




## Kissel BV vs Kissel UCLA data

The Kissel BV and Kissel UCLA data both contain samples from the second trimester of pregnancy and use the age as a covariate. Also they come from the same paper, so the methods used are exactly the same.
They should be the most similar results...

### Overlap

Get the number of significant genes overlapping between the 2 datasets.

```{r permut_test_overlap_BV_UCLA, echo=FALSE}
p <- overlap_signif(Kissel_UCLA_results, Kissel_BV_results, qvalueThreshold = qvalue_threshold)
p
p <- overlap_signif(Kissel_BV_results, Kissel_UCLA_results, qvalueThreshold = qvalue_threshold)
p
```


### Replication rate

We compute the replication rate (pi1).

```{r pi1_score_BV_UCLA, echo=FALSE}
print("Replication of Kissel BV sex-DE results in Kissel UCLA sex-DE analysis:")
pi1_normalized(Kissel_BV_results, Kissel_UCLA_results, qvalue_threshold = qvalue_threshold)
print("Replication of Kissel UCLA sex-DE results in Kissel BV sex-DE analysis:")
pi1_normalized(Kissel_UCLA_results, Kissel_BV_results, qvalue_threshold = qvalue_threshold)
```


### Effect size correlation

Correlation plot only of Kissel UCLA sex-DE genes.

```{r corr_UCLA_sexDE_BV, echo=FALSE}
p <- scatter_logFC(Kissel_UCLA_results[which(Kissel_UCLA_results$qvalue < qvalue_threshold),], Kissel_BV_results, "Kissel_UCLA", "Kissel_BV", 
                   pointColor = "black")
p
```

Correlation plot only of autosomal Kissel UCLA sex-DE genes.

```{r corr_UCLA_sexDE_auto_BV, echo=FALSE}
p <- scatter_logFC(Kissel_UCLA_results_auto[which(Kissel_UCLA_results_auto$qvalue < qvalue_threshold),], Kissel_BV_results, "Kissel_UCLA", "Kissel_BV", 
                   pointColor = "black")
p
```


### Effect size consistency

Compute the consistency of log fold change.

```{r consistency_score_UCLA_vs_BV, echo=FALSE}
print("Consistency of Kissel UCLA analysis compared to the results of Kissel BV analysis:")
consistency_test(Kissel_BV_results, Kissel_UCLA_results, qvalueThreshold = 0.01)
```


# Session info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
