---
title: "Visualize TFBS analysis results in adult brain"
author: "Clara Benoit-Pilven"
date: "`r Sys.Date()`"
output: html_document
---


```{r library, echo=FALSE}
suppressPackageStartupMessages( library(data.table) )
suppressPackageStartupMessages( library(tidyverse) )
suppressPackageStartupMessages( library(UpSetR) )
suppressPackageStartupMessages( library(gridExtra) )
```

```{r, setup, include=FALSE, echo=FALSE}
knitr::opts_knit$set(root.dir = '~/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/')
```

Parameters
```{r thresholds, echo=FALSE}
threshold_qvalue <- 0.01
threshold_logFC <- 0
```

# Load data

## Load DE genes

Load pseudotime analysis data.
```{r load_peudotime, echo=FALSE}
load("data/4-GTEx/1-GTEx_analysis/GTEx_v8_forebrain_pseudotime_analysis.RData")
```

Sex-DE genes
```{r load_sexDE, echo=FALSE}
de_results <- read.table("data/4-GTEx/1-GTEx_analysis/brain_forebrain_v8_VoomDream_topTable_with_sva_ind_pseudotime_interaction.txt",
                         header = TRUE)
# remove genes used for sex assignment
genes2remove <- c("XIST", as.character(de_results$gene_name[which(de_results$chr == "chrY" & de_results$gene_type == "protein_coding")]))
de_results_filtered <- de_results[which(!de_results$gene_name %in% genes2remove),]
```

Pseudotime-DE genes
```{r load_pseudotimeDE, echo=FALSE}
de_results_pseudotime <- read.table("data/4-GTEx/1-GTEx_analysis/brain_forebrain_v8_VoomDream_topTable_with_sva_ind_pseudotime_interaction_pseudotime.txt",
                                    header = TRUE)
# remove genes used for pseudotime inference
genes2remove <- rownames(selectedCounts)
de_results_pseudotime_filtered <- de_results_pseudotime[which(!de_results_pseudotime$gene_name %in% genes2remove),]
```

## Load TFBS results

```{r load_TF_results, echo=FALSE}
# sex-DE results
femaleBiased_TF <- readRDS("data/5-postprocessing/7-TFBS/adult/Unibind_femaleBiased/lolaLocResults.RDS")
maleBiased_TF <- readRDS("data/5-postprocessing/7-TFBS/adult/Unibind_maleBiased/lolaLocResults.RDS")
# pseudotime-DE results
downReg_TF <- readRDS("data/5-postprocessing/7-TFBS/adult/Unibind_downReg/lolaLocResults.RDS")
upReg_TF <- readRDS("data/5-postprocessing/7-TFBS/adult/Unibind_upReg/lolaLocResults.RDS")
```


# Explore results

## Compute the confidence interval 

For each enrichment, compute the confidence interval for the Fisher exact's test.

```{r function_compute_confInt, echo=FALSE}
confint_OR <- function(res, confLevel = 0.95){
  contTable <- matrix(as.numeric(res), 2, 2)
  return(fisher.test(contTable, conf.level = confLevel)$conf.int[1:2])
}
confint_OR_all <- function(data){
  res <- as.data.frame(data)
  #browser()
  CI_res <- apply(res[c("support", "b", "c", "d")], 1, confint_OR)
  res$CI_low <- CI_res[1,]
  res$CI_high <- CI_res[2,]
  res <- as.data.table(res)
  return(res)
}
```

```{r compute_confInt_allEnrichment, echo=FALSE}
# female-biased genes
femaleBiased_TF_withCI <- confint_OR_all(femaleBiased_TF)
# male-biased genes
maleBiased_TF_withCI <- confint_OR_all(maleBiased_TF)
# upregulated genes
upReg_TF_withCI <- confint_OR_all(upReg_TF)
# downregulated genes
downReg_TF_withCI <- confint_OR_all(downReg_TF)
```


## Extract significant TF

For each analysis, look at each TF and get the smallest q-value. Then, output the list of TF with q-value < 0.05.

```{r function_get_signif, echo=FALSE}
get_signif <- function(data_TF, qvalue_threshold = 0.05){
  data_TF_signif <- data_TF[which(data_TF$qValue < qvalue_threshold), ]
  TF_OR <- data_TF_signif %>% group_by(collection) %>% summarise(maxOddsRatio = suppressWarnings(max(oddsRatio)))
  names(TF_OR) <- c("TF", "max_oddsRatio")
  return(TF_OR)
}
```

```{r get_signif_TF, echo=FALSE}
print("Female-biased TF")
femaleBiased_TF_signif <- get_signif(femaleBiased_TF)
dim(femaleBiased_TF_signif)[1]
femaleBiased_TF_signif %>% print(n = nrow(femaleBiased_TF_signif))
print("Male-biased TF")
maleBiased_TF_signif <- get_signif(maleBiased_TF)
dim(maleBiased_TF_signif)[1]
maleBiased_TF_signif %>% print(n = nrow(maleBiased_TF_signif))

print("Pseudotime upregulated TF")
upReg_TF_signif <- get_signif(upReg_TF)
dim(upReg_TF_signif)[1]
upReg_TF_signif %>% print(n = nrow(upReg_TF_signif))
print("Pseudotime downregulated TF")
downReg_TF_signif <- get_signif(downReg_TF)
dim(downReg_TF_signif)[1]
downReg_TF_signif %>% print(n = nrow(downReg_TF_signif))
```

How many of these significant TF are differentially expressed between males and females.

```{r sexDE_TF, echo=FALSE}
# female TF
femaleBiased_TF_signif_sexDE <- de_results_filtered[which(de_results_filtered$gene_name %in% femaleBiased_TF_signif$TF & de_results_filtered$qvalue < threshold_qvalue), ]
sprintf("Number of female-enriched TF also sex-DE: %s", dim(femaleBiased_TF_signif_sexDE)[1])
print(femaleBiased_TF_signif_sexDE)
# male TF
maleBiased_TF_signif_sexDE <- de_results_filtered[which(de_results_filtered$gene_name %in% maleBiased_TF_signif$TF & de_results_filtered$qvalue < threshold_qvalue), ]
sprintf("Number of male-enriched TF also sex-DE: %s", dim(maleBiased_TF_signif_sexDE)[1])
print(maleBiased_TF_signif_sexDE)
```


## Overlap of significant TF between the 4 lists analyzed

```{r overlap_TF, echo=FALSE}
TF_lists <- list(female = femaleBiased_TF_signif$TF, male = maleBiased_TF_signif$TF, 
                 up = upReg_TF_signif$TF, down = downReg_TF_signif$TF)
upset(fromList(TF_lists), order.by = "freq")
```


## Format and save result table

Format adult sex-DE results.

```{r format_adult_sexDE_res, echo=FALSE}
# female biased
femaleBiased_TF_formatted <- femaleBiased_TF
femaleBiased_TF_formatted$userSet <- as.character(femaleBiased_TF_formatted$userSet)
femaleBiased_TF_formatted$userSet <- "femaleBiased"
names(femaleBiased_TF_formatted)[1] <- "data"
# male biased
maleBiased_TF_formatted <- maleBiased_TF
maleBiased_TF_formatted$userSet <- as.character(maleBiased_TF_formatted$userSet)
maleBiased_TF_formatted$userSet <- "maleBiased"
names(maleBiased_TF_formatted)[1] <- "data"
# merge
sexDE_TF_formatted <- rbind(femaleBiased_TF_formatted, maleBiased_TF_formatted)
```

Save adult sex-DE results.

```{r save_adult_sexDE_res, echo=FALSE}
write.table(sexDE_TF_formatted, file = "data/5-postprocessing/7-TFBS/adult/sexDE_TF_results.txt",
            row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)
```

Format adult pseudotime-DE results.

```{r format_adult_pseudotimeDE_res, echo=FALSE}
# female biased
upReg_TF_formatted <- upReg_TF
upReg_TF_formatted$userSet <- as.character(upReg_TF_formatted$userSet)
upReg_TF_formatted$userSet <- "upRegulated"
names(upReg_TF_formatted)[1] <- "data"
# male biased
downReg_TF_formatted <- downReg_TF
downReg_TF_formatted$userSet <- as.character(downReg_TF_formatted$userSet)
downReg_TF_formatted$userSet <- "downRegulated"
names(downReg_TF_formatted)[1] <- "data"
# merge
pseudotimeDE_TF_formatted <- rbind(upReg_TF_formatted, downReg_TF_formatted)
```

Save adult pseudotime-DE results.

```{r save_adult_pseudotimeDE_res, echo=FALSE}
write.table(pseudotimeDE_TF_formatted, file = "data/5-postprocessing/7-TFBS/adult/pseudotimeDE_TF_results.txt",
            row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)
```


## Forest plot

```{r function_format_forestPlot, echo=FALSE}
formatData_forestPlot <- function(data_TF, qvalue_threshold = 0.05){
  tfs <- unique(data_TF$collection)
  selected_TF <- data.frame()
  # for each TF
  for (myTF in tfs){
    # get all data for this TF
    currentData <- data_TF[which(data_TF$collection == myTF), ]
    # get all data with qvalue < threshold
    currentData_signif <- currentData[which(currentData$qValue < qvalue_threshold), ]
    if (dim(currentData_signif)[1] >= 1){
      # and keep only the highest OR
      selected_TF <- rbind(selected_TF, currentData_signif[which.max(currentData_signif$oddsRatio), ])
    } else{
      # if no signif qvalue keep lower qvalue
      selected_TF <- rbind(selected_TF, currentData[which.min(currentData$qValue), ])
    }
  }
  # select only columns of interest
  selected_TF_filtered <- selected_TF[, c("collection", "qValue", "oddsRatio", "CI_low", "CI_high")]
  return(selected_TF_filtered)
}
```

Format data.

```{r format_data_for_forestPlot, echo=FALSE}
femaleBiased_TF_withCI_forestPlot <- formatData_forestPlot(femaleBiased_TF_withCI)
femaleBiased_TF_withCI_forestPlot$data <- "sex_femaleBiased"
maleBiased_TF_withCI_forestPlot <- formatData_forestPlot(maleBiased_TF_withCI)
maleBiased_TF_withCI_forestPlot$data <- "sex_maleBiased"
upReg_TF_withCI_forestPlot <- formatData_forestPlot(upReg_TF_withCI)
upReg_TF_withCI_forestPlot$data <- "pseudotime_upregulated"
downReg_TF_withCI_forestPlot <- formatData_forestPlot(downReg_TF_withCI)
downReg_TF_withCI_forestPlot$data <- "pseudotime_downregulated"
# merge the 4 data
TF_withCI_forestPlot <- rbind(femaleBiased_TF_withCI_forestPlot, maleBiased_TF_withCI_forestPlot, 
                              upReg_TF_withCI_forestPlot, downReg_TF_withCI_forestPlot)
```

Add significance annotation.

```{r add_signif_annot, echo=FALSE}
add_signif_annot <- function(data_TF, qvalue_threshold = 0.05){
  data_TF$signif_annot <- ifelse(data_TF$qValue < qvalue_threshold, "*", "")
  return(data_TF)
}
femaleBiased_TF_withCI_forestPlot <- add_signif_annot(femaleBiased_TF_withCI_forestPlot)
maleBiased_TF_withCI_forestPlot <- add_signif_annot(maleBiased_TF_withCI_forestPlot)
upReg_TF_withCI_forestPlot <- add_signif_annot(upReg_TF_withCI_forestPlot)
downReg_TF_withCI_forestPlot <- add_signif_annot(downReg_TF_withCI_forestPlot)
TF_withCI_forestPlot <- add_signif_annot(TF_withCI_forestPlot)
```

Choose TF to show: get the TF significant in at least 3 datasets + some choosen ones.

```{r choose_TF, echo=FALSE}
nSignif_data <- TF_withCI_forestPlot %>% group_by(collection) %>% summarise(nSignif = sum(signif_annot == "*"))
choosenTF <- nSignif_data[which(nSignif_data$nSignif >= 3), "collection"]
choosenTF_withCI_forestPlot <- TF_withCI_forestPlot[which(TF_withCI_forestPlot$collection %in% choosenTF$collection | TF_withCI_forestPlot$collection %in% c("EGR1", "ELK1", "KLF5") ), ]
# separate the TFBS in 3 lists.
choosenTF_withCI_forestPlot$listPlot <- ifelse(choosenTF_withCI_forestPlot$collection %in% c("REST", "ESR1", "MYC", "E2F1", "E2F4"), 
                                           "list1",
                                           ifelse(choosenTF_withCI_forestPlot$collection %in% c("NFE2L2", "E2F6", "NFKB1", "CREB1", "ELK1"),
                                                  "list2",
                                           ifelse(choosenTF_withCI_forestPlot$collection %in% c("MAX", "ELF1", "TFAP2A", "EGR1", "KLF5"),
                                                  "list3",
                                                  NA)))
```

Forest plot.

```{r function_forest_plot, echo=FALSE}
forestPlot <- function(dataPlot, legend = TRUE, annot_dist = 0.1){
  p <- ggplot(dataPlot, aes(x = oddsRatio, y = collection, xmin = CI_low, xmax = CI_high,
                            color = data, fill = data))
  p <- p + geom_vline(xintercept = 1, color = "darkgrey")
  p <- p + geom_pointrange(position = position_dodge(width = 0.75))
  p <- p + scale_color_manual(values = c("sex_maleBiased" = 'darkblue', "sex_femaleBiased" = 'darkred',
                                         "pseudotime_upregulated" = 'aquamarine4', "pseudotime_downregulated" = "chocolate"), 
                              labels = c("sex_maleBiased" = 'Male biased', "sex_femaleBiased" = 'Female biased',
                                         "pseudotime_upregulated" = 'Pseudotime upreg', "pseudotime_downregulated" = "Pseudotime downreg"),
                              name = "")
  p <- p + scale_fill_manual(values = c("sex_maleBiased" = 'darkblue', "sex_femaleBiased" = 'darkred',
                                        "pseudotime_upregulated" = 'aquamarine4', "pseudotime_downregulated" = "chocolate"), 
                             labels = c("sex_maleBiased" = 'Male biased', "sex_femaleBiased" = 'Female biased',
                                        "pseudotime_upregulated" = 'Pseudotime upreg', "pseudotime_downregulated" = "Pseudotime downreg"),
                             name = "")
  p <- p + geom_text(aes(x = CI_high+annot_dist, y = collection, label = signif_annot), size = 8, position = position_dodge(width = 0.75))
  p <- p + theme_minimal()
  p <- p + labs(x = "", y = "")
  p <- p + guides(fill = guide_legend(override.aes = list(label = "")))
  if (!legend){
    p <- p + guides(color = "none", fill = "none")
  }
  return(p)
}
```

```{r forest_plot, echo=FALSE}
p1 <- forestPlot(choosenTF_withCI_forestPlot[which(choosenTF_withCI_forestPlot$listPlot == "list1")], 
                 legend = FALSE, annot_dist = 1.5)
p1 <- p1 + coord_cartesian(xlim = c(0,11))
p2 <- forestPlot(choosenTF_withCI_forestPlot[which(choosenTF_withCI_forestPlot$listPlot == "list2")], legend = FALSE)
p2 <- p2 + coord_cartesian(xlim = c(0.5,2.5))
p3 <- forestPlot(choosenTF_withCI_forestPlot[which(choosenTF_withCI_forestPlot$listPlot == "list3")])
p3 <- p3 + coord_cartesian(xlim = c(0.5,2))
p <- grid.arrange(p1, p2, p3, ncol = 3, widths = c(1, 1, 2.2),
                  left = "Transcription factor binding site", bottom = "odds ratio")
# ggsave("images/4-postprocessing/7-TFBS/TFBS_enrichment_forest_plot.png", plot = p, 
#        width = 25, height = 15, units = "cm")
```


# Session Info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
