---
title: "Enrichment for transcription factor binding sites in promoter regions of adult sex-biased genes"
author: "Clara Benoit-Pilven"
date: "`r Sys.Date()`"
output: html_document
---

```{r library, echo=FALSE}
suppressPackageStartupMessages( library(rtracklayer) )
suppressPackageStartupMessages( library(data.table) )
```

```{r, setup, include=FALSE, echo=FALSE}
knitr::opts_knit$set(root.dir = '~/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/')
```

Parameters
```{r thresholds, echo=FALSE}
threshold_qvalue <- 0.01
threshold_logFC <- 0
```

Functions
```{r function_save_bed_file, echo=FALSE}
save_bedFile <- function(data, fileName){
  bed <- data[, c("chrom", "promoter_start", "promoter_end", "gene_id", "gene_name", "strand")]
  colnames(bed) <- c('chrom', 'chromStart', 'chromEnd', 'name', 'score', 'strand')
  bed$score <- 0
  write.table(bed, file = fileName, quote = FALSE, sep = "\t", col.names = FALSE, row.names = FALSE)
}
```


# Generate bed files for Unibind analysis

## Load DE genes

Load pseudotime analysis data.
```{r load_peudotime, echo=FALSE}
load("data/4-GTEx/1-GTEx_analysis/GTEx_v8_forebrain_pseudotime_analysis.RData")
```

Sex-DE genes
```{r load_sexDE, echo=FALSE}
de_results <- read.table("data/4-GTEx/1-GTEx_analysis/brain_forebrain_v8_VoomDream_topTable_with_sva_ind_pseudotime_interaction.txt",
                         header = TRUE)
# remove genes used for sex assignment
genes2remove <- c("XIST", as.character(de_results$gene_name[which(de_results$chr == "chrY" & de_results$gene_type == "protein_coding")]))
de_results_filtered <- de_results[which(!de_results$gene_name %in% genes2remove),]
```

Pseudotime-DE genes
```{r load_pseudotimeDE, echo=FALSE}
de_results_pseudotime <- read.table("data/4-GTEx/1-GTEx_analysis/brain_forebrain_v8_VoomDream_topTable_with_sva_ind_pseudotime_interaction_pseudotime.txt",
                                    header = TRUE)
# remove genes used for pseudotime inference
genes2remove <- rownames(selectedCounts)
de_results_pseudotime_filtered <- de_results_pseudotime[which(!de_results_pseudotime$gene_name %in% genes2remove),]
```


## Load annotation and prepare promoters

```{r load_annotation, echo=FALSE}
gencode26_annot <- rtracklayer::import("data/0-input/gencode.v26.annotation.gtf.gz")
gencode26 <- gencode26_annot[which(elementMetadata(gencode26_annot)$type == "gene"),]
gencode26 <- data.frame(gene_id = elementMetadata(gencode26)$gene_id,
                    gene_name = elementMetadata(gencode26)$gene_name,
                    gene_type = elementMetadata(gencode26)$gene_type,
                    chr = seqnames(gencode26),
                    start = start(gencode26),
                    end = end(gencode26),
                    strand = strand(gencode26))
```

Prepare promoters.

```{r prepare_coord_prom, echo=FALSE}
promoters <- data.frame(gene_id = gencode26$gene_id, gene_name = gencode26$gene_name, chrom = gencode26$chr, strand = gencode26$strand, 
                        promoter_start = 0, promoter_end = 0)
for (i in 1:dim(gencode26)[1]){
  if (gencode26$strand[i] == "+"){
    promoters$promoter_start[i] <- as.integer(gencode26$start[i]-2000-1) # -1 because bed files are 0 based
    promoters$promoter_end[i] <- as.integer(gencode26$start[i])
  } else if (gencode26$strand[i] == "-"){
    promoters$promoter_start[i] <- as.integer(gencode26$end[i]-1) # -1 because bed files are 0 based
    promoters$promoter_end[i] <- as.integer(gencode26$end[i]+2000)
  } else{
    print("Unknown strand...")
  }
}
```

## Get promoters for all genes analyzed

```{r save_bed_file_universe, echo=FALSE}
promoters_universe <- promoters[which(promoters$gene_id %in% de_results_filtered$gene_id), ]
save_bedFile(promoters_universe, "data/5-postprocessing/7-TFBS/adult/universe_promoters.bed")
```

```{r save_bed_file_universe_pseudotime, echo=FALSE}
promoters_universe_pseudotime <- promoters[which(promoters$gene_id %in% de_results_pseudotime_filtered$gene_id), ]
save_bedFile(promoters_universe_pseudotime, "data/5-postprocessing/7-TFBS/adult/universe_pseudotime_promoters.bed")
```

## Extract promoters of DE genes

### Sex-DE genes

Separate male and female-biased genes.
```{r separate_male_female_bias, echo=FALSE}
maleBiased <- de_results_filtered[which(de_results_filtered$qvalue < threshold_qvalue & de_results_filtered$logFC <= threshold_logFC), ]
femaleBiased <- de_results_filtered[which(de_results_filtered$qvalue < threshold_qvalue & de_results_filtered$logFC >= threshold_logFC), ]
```

Select the promoters for the sex-biased genes.
```{r promoter_sexBiased, echo=FALSE}
promoters_maleBiased <- promoters[which(promoters$gene_id %in% maleBiased$gene_id), ]
promoters_femaleBiased <- promoters[which(promoters$gene_id %in% femaleBiased$gene_id), ]
```

Save these promoters as bed files.
```{r save_bed_files, echo=FALSE}
save_bedFile(promoters_maleBiased, "data/5-postprocessing/7-TFBS/adult/maleBiased_promoters.bed")
save_bedFile(promoters_femaleBiased, "data/5-postprocessing/7-TFBS/adult/femaleBiased_promoters.bed")
```

### Pseudotime-DE genes

Separate up and down-regulated genes.
```{r separate_up_down_reg, echo=FALSE}
downReg <- de_results_pseudotime_filtered[which(de_results_pseudotime_filtered$qvalue < threshold_qvalue & de_results_pseudotime_filtered$logFC <= threshold_logFC), ]
upReg <- de_results_pseudotime_filtered[which(de_results_pseudotime_filtered$qvalue < threshold_qvalue & de_results_pseudotime_filtered$logFC >= threshold_logFC), ]
```

Select the promoters for the sex-biased genes.
```{r promoter_pseudotimeBiased, echo=FALSE}
promoters_downReg <- promoters[which(promoters$gene_id %in% downReg$gene_id), ]
promoters_upReg <- promoters[which(promoters$gene_id %in% upReg$gene_id), ]
```

Save these promoters as bed files.
```{r save_bed_files_pseudotime, echo=FALSE}
save_bedFile(promoters_downReg, "data/5-postprocessing/7-TFBS/adult/downReg_promoters.bed")
save_bedFile(promoters_upReg, "data/5-postprocessing/7-TFBS/adult/upReg_promoters.bed")
```


# Run analysis in bash

```{bash unibind_analysis, echo=FALSE, eval= FALSE}
# from /Users/benoicla/Desktop/Tools/unibind_enrichment/ directory
## analysis for male-biased genes
bin/UniBind_enrich.sh oneSetBg /Users/benoicla/Desktop/Tools/unibind_enrichment/data/hg38_robust_UniBind_LOLA.RDS /Users/benoicla/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/data/5-postprocessing/7-TFBS/adult/maleBiased_promoters.bed /Users/benoicla/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/data/5-postprocessing/7-TFBS/adult/universe_promoters.bed /Users/benoicla/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/data/5-postprocessing/7-TFBS/adult/Unibind_maleBiased
## analysis for female-biased genes
bin/UniBind_enrich.sh oneSetBg /Users/benoicla/Desktop/Tools/unibind_enrichment/data/hg38_robust_UniBind_LOLA.RDS /Users/benoicla/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/data/5-postprocessing/7-TFBS/adult/femaleBiased_promoters.bed /Users/benoicla/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/data/5-postprocessing/7-TFBS/adult/universe_promoters.bed /Users/benoicla/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/data/5-postprocessing/7-TFBS/adult/Unibind_femaleBiased

## analysis for downregulated genes
bin/UniBind_enrich.sh oneSetBg /Users/benoicla/Desktop/Tools/unibind_enrichment/data/hg38_robust_UniBind_LOLA.RDS /Users/benoicla/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/data/5-postprocessing/7-TFBS/adult/downReg_promoters.bed /Users/benoicla/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/data/5-postprocessing/7-TFBS/adult/universe_pseudotime_promoters.bed /Users/benoicla/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/data/5-postprocessing/7-TFBS/adult/Unibind_downReg
## analysis for upregulated genes
bin/UniBind_enrich.sh oneSetBg /Users/benoicla/Desktop/Tools/unibind_enrichment/data/hg38_robust_UniBind_LOLA.RDS /Users/benoicla/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/data/5-postprocessing/7-TFBS/adult/upReg_promoters.bed /Users/benoicla/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/data/5-postprocessing/7-TFBS/adult/universe_pseudotime_promoters.bed /Users/benoicla/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/data/5-postprocessing/7-TFBS/adult/Unibind_upReg
```


# Session Info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```