---
title: "Visualize TFBS analysis results in shared and prenatal-specific sex-DE brain"
author: "Clara Benoit-Pilven"
date: "`r Sys.Date()`"
output: html_document
---

```{r library, echo=FALSE}
suppressPackageStartupMessages( library(data.table) )
suppressPackageStartupMessages( library(tidyverse) )
suppressPackageStartupMessages( library(UpSetR) )
suppressPackageStartupMessages( library(gridExtra) )
```

```{r, setup, include=FALSE, echo=FALSE}
knitr::opts_knit$set(root.dir = '~/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/')
```

Parameters
```{r thresholds, echo=FALSE}
threshold_qvalue <- 0.01
threshold_logFC <- 0
```

# Load data

## Load bayes genes

```{r load_bayes_results, echo=FALSE}
bayesResults <- read.table("data/4-GTEx/2-comparison/lineModel/lineModel_prenatalVSadult_6models_withAnnotation.txt", 
                           header = TRUE, sep = "\t")
```

## Load DE genes

Sex-DE genes
```{r load_sexDE, echo=FALSE}
de_results <- read.table("data/3-DE/VoomDream_topTable_forebrain_pseudotime_sva_ind_interaction.txt",
                         header = TRUE)
# remove genes used for sex assignment
genes2remove <- c("XIST", as.character(de_results$gene_name[which(de_results$chr == "chrY" & de_results$gene_type == "protein_coding")]))
de_results_filtered <- de_results[which(!de_results$gene_name %in% genes2remove),]
```


## Load TFBS results

```{r load_TF_results, echo=FALSE}
# shared sex-DE results
shared_femaleBiased_TF <- readRDS("data/5-postprocessing/7-TFBS/shared/Unibind_shared_femaleBiased/lolaLocResults.RDS")
shared_maleBiased_TF <- readRDS("data/5-postprocessing/7-TFBS/shared/Unibind_shared_maleBiased/lolaLocResults.RDS")
# prenatal-specific sex-DE results
prenatal_femaleBiased_TF <- readRDS("data/5-postprocessing/7-TFBS/shared/Unibind_prenatal_femaleBiased/lolaLocResults.RDS")
prenatal_maleBiased_TF <- readRDS("data/5-postprocessing/7-TFBS/shared/Unibind_prenatal_maleBiased/lolaLocResults.RDS")
```


# Explore results

## Compute the confidence interval 

For each enrichment, compute the confidence interval for the Fisher exact's test.

```{r function_compute_confInt, echo=FALSE}
confint_OR <- function(res, confLevel = 0.95){
  contTable <- matrix(as.numeric(res), 2, 2)
  return(fisher.test(contTable, conf.level = confLevel)$conf.int[1:2])
}
confint_OR_all <- function(data){
  res <- as.data.frame(data)
  #browser()
  CI_res <- apply(res[c("support", "b", "c", "d")], 1, confint_OR)
  res$CI_low <- CI_res[1,]
  res$CI_high <- CI_res[2,]
  res <- as.data.table(res)
  return(res)
}
```

```{r compute_confInt_allEnrichment, echo=FALSE}
# shared female-biased genes
shared_femaleBiased_TF_withCI <- confint_OR_all(shared_femaleBiased_TF)
# shared male-biased genes
shared_maleBiased_TF_withCI <- confint_OR_all(shared_maleBiased_TF)
# prenatal female-biased genes
prenatal_femaleBiased_TF_withCI <- confint_OR_all(prenatal_femaleBiased_TF)
# prenatal male-biased genes
prenatal_maleBiased_TF_withCI <- confint_OR_all(prenatal_maleBiased_TF)
```


## Extract significant TF

For each analysis, look at each TF and get the smallest q-value. Then, output the list of TF with q-value < 0.05.

```{r function_get_signif, echo=FALSE}
get_signif <- function(data_TF, qvalue_threshold = 0.05){
  data_TF_signif <- data_TF[which(data_TF$qValue < qvalue_threshold), ]
  if (dim(data_TF_signif)[1] == 0){
    return(data.frame("TF" = NULL, "max_oddsRatio" = NULL))
  }
  TF_OR <- data_TF_signif %>% group_by(collection) %>% summarise(maxOddsRatio = suppressWarnings(max(oddsRatio, na.rm = TRUE)))
  names(TF_OR) <- c("TF", "max_oddsRatio")
  return(TF_OR)
}
```

```{r get_signif_TF, echo=FALSE}
print("Shared female-biased TF")
shared_femaleBiased_TF_signif <- get_signif(shared_femaleBiased_TF)
dim(shared_femaleBiased_TF_signif)[1]
shared_femaleBiased_TF_signif %>% print(n = nrow(shared_femaleBiased_TF_signif))
print("Shared male-biased TF")
shared_maleBiased_TF_signif <- get_signif(shared_maleBiased_TF)
dim(shared_maleBiased_TF_signif)[1]
shared_maleBiased_TF_signif %>% print(n = nrow(shared_maleBiased_TF_signif))

print("Prenatal female-biased TF")
prenatal_femaleBiased_TF_signif <- get_signif(prenatal_femaleBiased_TF)
dim(prenatal_femaleBiased_TF_signif)[1]
prenatal_femaleBiased_TF_signif %>% print(n = nrow(prenatal_femaleBiased_TF_signif))
print("Prenatal male-biased TF")
prenatal_maleBiased_TF_signif <- get_signif(prenatal_maleBiased_TF)
dim(prenatal_maleBiased_TF_signif)[1]
prenatal_maleBiased_TF_signif %>% print(n = nrow(prenatal_maleBiased_TF_signif))
```

How many of these significant TF are differentially expressed between males and females in prenatal and/or adult dataset.

```{r sexDE_TF, echo=FALSE}
# prenatal female TF
prenatal_femaleBiased_TF_signif_sexDE <- de_results_filtered[which(de_results_filtered$gene_name %in% prenatal_femaleBiased_TF_signif$TF & de_results_filtered$qvalue < threshold_qvalue), ]
sprintf("Number of prenatal female-enriched TF also sex-DE: %s", dim(prenatal_femaleBiased_TF_signif_sexDE)[1])
print(prenatal_femaleBiased_TF_signif_sexDE)
#  prenatal male TF
prenatal_maleBiased_TF_signif_sexDE <- de_results_filtered[which(de_results_filtered$gene_name %in% prenatal_maleBiased_TF_signif$TF & de_results_filtered$qvalue < threshold_qvalue), ]
sprintf("Number of prenatal male-enriched TF also sex-DE: %s", dim(prenatal_maleBiased_TF_signif_sexDE)[1])
print(prenatal_maleBiased_TF_signif_sexDE)
```


## Overlap of significant TF between the 4 lists analyzed

```{r overlap_TF, echo=FALSE}
TF_lists <- list(prenatal_female = prenatal_femaleBiased_TF_signif$TF, prenatal_male = prenatal_maleBiased_TF_signif$TF, 
                 shared_female = shared_femaleBiased_TF_signif$TF, shared_male = shared_maleBiased_TF_signif$TF)
upset(fromList(TF_lists), order.by = "freq")
```

As there was no TF enriched in the shared gene list, only the prenatal data is shown.


## Format and save result table

Format prenatal-specific sex-DE results.

```{r format_prenatalSpe_sexDE_res, echo=FALSE}
# female biased
prenatal_femaleBiased_TF_formatted <- prenatal_femaleBiased_TF
prenatal_femaleBiased_TF_formatted$userSet <- as.character(prenatal_femaleBiased_TF_formatted$userSet)
prenatal_femaleBiased_TF_formatted$userSet <- "femaleBiased"
names(prenatal_femaleBiased_TF_formatted)[1] <- "data"
# male biased
prenatal_maleBiased_TF_formatted <- prenatal_maleBiased_TF
prenatal_maleBiased_TF_formatted$userSet <- as.character(prenatal_maleBiased_TF_formatted$userSet)
prenatal_maleBiased_TF_formatted$userSet <- "maleBiased"
names(prenatal_maleBiased_TF_formatted)[1] <- "data"
# merge
prenatal_sexDE_TF_formatted <- rbind(prenatal_femaleBiased_TF_formatted, prenatal_maleBiased_TF_formatted)
```

Save adult sex-DE results.

```{r save_prenatalSpe_sexDE_res, echo=FALSE}
write.table(prenatal_sexDE_TF_formatted, file = "data/5-postprocessing/7-TFBS/shared/prenatalSpecific_sexDE_TF_results.txt",
            row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)
```

Format shared sex-DE results.

```{r format_shared_sexDE_res, echo=FALSE}
# female biased
shared_femaleBiased_TF_formatted <- shared_femaleBiased_TF
shared_femaleBiased_TF_formatted$userSet <- as.character(shared_femaleBiased_TF_formatted$userSet)
shared_femaleBiased_TF_formatted$userSet <- "femaleBiased"
names(shared_femaleBiased_TF_formatted)[1] <- "data"
# male biased
shared_maleBiased_TF_formatted <- shared_maleBiased_TF
shared_maleBiased_TF_formatted$userSet <- as.character(shared_maleBiased_TF_formatted$userSet)
shared_maleBiased_TF_formatted$userSet <- "maleBiased"
names(shared_maleBiased_TF_formatted)[1] <- "data"
# merge
shared_sexDE_TF_formatted <- rbind(shared_femaleBiased_TF_formatted, shared_maleBiased_TF_formatted)
```

Save adult sex-DE results.

```{r save_shared_sexDE_res, echo=FALSE}
write.table(shared_sexDE_TF_formatted, file = "data/5-postprocessing/7-TFBS/shared/shared_sexDE_TF_results.txt",
            row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)
```


## Extract the genes linked to TF or interest

To do this, I need to manually run the analysis between the TF of interest and the sex-DE genes.

```{r extract_genesTFBS_prenatalMale_AR, echo=FALSE, eval=FALSE}
# create GRange object for gene promoters
prenatal_maleBiased_promoters <- readBed("~/Desktop/Workspace/HDBR/sexDE_prenatal_brain/data/4-postprocessing/7-TFBS/Unibind/prenatal_maleBiased_promoters.bed")
# load Unibind regionDB
regionDB <- readRDS("/Users/benoicla/Desktop/Workspace/HDBR/sexDE_prenatal_brain/data/4-postprocessing/7-TFBS/Unibind/hg38_robust_UniBind_LOLA.RDS")
# subset the regionDB to contain only AR data
regionDB_AR <- regionDB
#regionDB_AR$regionGRL <- regionDB_AR$regionGRL[[which(regionDB_AR$regionAnno$collection == "AR")]] # doesn't work...
#regionDB_AR$regionGRL <- subset(regionDB_AR$regionGRL, regionDB_AR$regionAnno$collection == "AR") # doesn't work either...
listGRL <- list()
for (i in which(regionDB_AR$regionAnno$collection == "AR")){
  listGRL[[length(listGRL)+1]] <- regionDB_AR$regionGRL[[i]]
}
regionDB_AR$regionGRL <- GRangesList(listGRL)
regionDB_AR$regionAnno <- regionDB_AR$regionAnno[which(regionDB_AR$regionAnno$collection == "AR"), ]
regionDB_AR$collectionAnno <- regionDB_AR$collectionAnno[which(regionDB_AR$collectionAnno == "AR"), ]
# enrichment analysis
AR_prenatal_maleBiased <- runLOLA(prenatal_maleBiased_promoters, universe, regionDB_AR)
AR_prenatal_maleBiased_signif <- AR_prenatal_maleBiased[1,]
# extract the region responsible for the enrichment
AR_prenatal_maleBiased_regions <- extractEnrichmentOverlaps(AR_prenatal_maleBiased_signif, prenatal_maleBiased_promoters, regionDB_AR)
# get the genes corresponding to these region responsible for the enrichment
AR_prenatal_maleBiased_genes <- rownames(as.data.frame(subsetByOverlaps(prenatal_maleBiased_promoters, AR_prenatal_maleBiased_regions, ignore.strand = TRUE, type = "any")))
AR_prenatal_maleBiased_genesDE <- bayesResults[which(bayesResults$ensemblID %in% AR_prenatal_maleBiased_genes), ]

# Load prenatal sex-DE results
de_results <- read.table("data/3-DE/VoomDream_topTable_forebrain_pseudotime_sva_ind_interaction.txt", header = TRUE)
AR_prenatal_maleBiased_genesDE2 <- merge(AR_prenatal_maleBiased_genesDE, de_results)
```

```{r extract_genesTFBS_prenatalFemale_AR, echo=FALSE, eval=FALSE}
# create GRange object for gene promoters
prenatal_femaleBiased_promoters <- readBed("~/Desktop/Workspace/HDBR/sexDE_prenatal_brain/data/4-postprocessing/7-TFBS/Unibind/prenatal_femaleBiased_promoters.bed")
# load Unibind regionDB
regionDB <- readRDS("/Users/benoicla/Desktop/Workspace/HDBR/sexDE_prenatal_brain/data/4-postprocessing/7-TFBS/Unibind/hg38_robust_UniBind_LOLA.RDS")
# subset the regionDB to contain only AR data
regionDB_AR <- regionDB
#regionDB_AR$regionGRL <- regionDB_AR$regionGRL[[which(regionDB_AR$regionAnno$collection == "AR")]] # doesn't work...
#regionDB_AR$regionGRL <- subset(regionDB_AR$regionGRL, regionDB_AR$regionAnno$collection == "AR") # doesn't work either...
listGRL <- list()
for (i in which(regionDB_AR$regionAnno$collection == "AR")){
  listGRL[[length(listGRL)+1]] <- regionDB_AR$regionGRL[[i]]
}
regionDB_AR$regionGRL <- GRangesList(listGRL)
regionDB_AR$regionAnno <- regionDB_AR$regionAnno[which(regionDB_AR$regionAnno$collection == "AR"), ]
regionDB_AR$collectionAnno <- regionDB_AR$collectionAnno[which(regionDB_AR$collectionAnno == "AR"), ]
# enrichment analysis
AR_prenatal_femaleBiased <- runLOLA(prenatal_femaleBiased_promoters, universe, regionDB_AR)
AR_prenatal_femaleBiased_signif <- AR_prenatal_femaleBiased[1,]
# extract the region responsible for the enrichment
AR_prenatal_femaleBiased_regions <- extractEnrichmentOverlaps(AR_prenatal_femaleBiased_signif, prenatal_femaleBiased_promoters, regionDB_AR)
# get the genes corresponding to these region responsible for the enrichment
AR_prenatal_femaleBiased_genes <- rownames(as.data.frame(subsetByOverlaps(prenatal_femaleBiased_promoters, AR_prenatal_femaleBiased_regions, ignore.strand = TRUE, type = "any")))
AR_prenatal_femaleBiased_genesDE <- bayesResults[which(bayesResults$ensemblID %in% AR_prenatal_femaleBiased_genes), ]

# Load prenatal sex-DE results
de_results <- read.table("~/Desktop/Workspace/HDBR/sexDE_prenatal_brain/data/3-DE/VoomDream_topTable_forebrain_pseudotime_sva_ind_interaction.txt", header = TRUE)
AR_prenatal_femaleBiased_genesDE2 <- merge(AR_prenatal_femaleBiased_genesDE, de_results)
```


```{r extract_genesTFBS_prenatalFemale_ESR1, echo=FALSE, eval=FALSE}
# create GRange object for gene promoters
prenatal_femaleBiased_promoters <- readBed("~/Desktop/Workspace/HDBR/sexDE_prenatal_brain/data/4-postprocessing/7-TFBS/Unibind/prenatal_femaleBiased_promoters.bed")
# load Unibind regionDB
#regionDB <- readRDS("/Users/benoicla/Desktop/Workspace/HDBR/sexDE_prenatal_brain/data/4-postprocessing/7-TFBS/Unibind/hg38_robust_UniBind_LOLA.RDS")
# subset the regionDB to contain only AR data
regionDB_ESR1 <- regionDB
listGRL <- list()
for (i in which(regionDB_ESR1$regionAnno$collection == "ESR1")){
  listGRL[[length(listGRL)+1]] <- regionDB_ESR1$regionGRL[[i]]
}
regionDB_ESR1$regionGRL <- GRangesList(listGRL)
regionDB_ESR1$regionAnno <- regionDB_ESR1$regionAnno[which(regionDB_ESR1$regionAnno$collection == "ESR1"), ]
regionDB_ESR1$collectionAnno <- regionDB_ESR1$collectionAnno[which(regionDB_ESR1$collectionAnno == "ESR1"), ]
# enrichment analysis
ESR1_prenatal_femaleBiased <- runLOLA(prenatal_femaleBiased_promoters, universe, regionDB_ESR1)
ESR1_prenatal_femaleBiased_signif <- ESR1_prenatal_femaleBiased[2,]
# extract the region responsible for the enrichment
ESR1_prenatal_femaleBiased_regions <- extractEnrichmentOverlaps(ESR1_prenatal_femaleBiased_signif, prenatal_femaleBiased_promoters, regionDB_ESR1)
# get the genes corresponding to these region responsible for the enrichment
ESR1_prenatal_femaleBiased_genes <- rownames(as.data.frame(subsetByOverlaps(prenatal_femaleBiased_promoters, ESR1_prenatal_femaleBiased_regions, ignore.strand = TRUE, type = "any")))
ESR1_prenatal_femaleBiased_genesDE <- bayesResults[which(bayesResults$ensemblID %in% ESR1_prenatal_femaleBiased_genes), ]

# Load prenatal sex-DE results
de_results <- read.table("~/Desktop/Workspace/HDBR/sexDE_prenatal_brain/data/3-DE/VoomDream_topTable_forebrain_pseudotime_sva_ind_interaction.txt", header = TRUE)
ESR1_prenatal_femaleBiased_genesDE2 <- merge(ESR1_prenatal_femaleBiased_genesDE, de_results)
```


Go enrichment analysis for the male- or female-biased genes bound by either AR or ESR1.

```{r GO_enrich, echo=FALSE, eval=FALSE}
suppressPackageStartupMessages( library(clusterProfiler) )
suppressPackageStartupMessages( library(org.Hs.eg.db) )
# create universe
bayesRes <- read.table("/Users/benoicla/Desktop/Workspace/HDBR/sexDE_prenatal_brain/data/4-postprocessing/4-Comparison_GTEx/sexDE_specificity/lineModel_prenatalVSadult_6models.txt", header = TRUE, sep = " ")
universe_bayes <- as.character(bayesRes$gene_name)
# prepare gene lists of up and down regulated genes 
clusters <- list("AR male-biased" = AR_prenatal_maleBiased_genesDE$gene_name,
                 "AR female-biased" = AR_prenatal_femaleBiased_genesDE$gene_name,
                 "ESR1 female-biased" = ESR1_prenatal_femaleBiased_genesDE$gene_name)
# do enrichment analyses
resGO_BP <- compareCluster(clusters, fun = "enrichGO", universe = universe_bayes, ont = "BP", OrgDb = org.Hs.eg.db, 
                        keyType = "SYMBOL", minGSSize = 15, maxGSSize = 500, pvalueCutoff = 0.05)
resGO_MF <- compareCluster(clusters, fun = "enrichGO", universe = universe_bayes, ont = "MF", OrgDb = org.Hs.eg.db, 
                        keyType = "SYMBOL", minGSSize = 15, maxGSSize = 500, pvalueCutoff = 0.05)
resGO_CC <- compareCluster(clusters, fun = "enrichGO", universe = universe_bayes, ont = "CC", OrgDb = org.Hs.eg.db, 
                        keyType = "SYMBOL", minGSSize = 15, maxGSSize = 500, pvalueCutoff = 0.05)
# plot function
plot_resGO <- function(resGO){
  p <- dotplot(resGO, showCategory = 20, label_format = 100)
  p <- p + theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
  p <- p + xlab("")
  p <- p + scale_color_gradient(low = "#e6e6e6", high = "#000000", trans = 'reverse')
  return(p)
}
p <- plot_resGO(resGO_BP)
p
ggsave("~/Desktop/Workspace/HDBR/sexDE_prenatal_brain/images/4-postprocessing/7-TFBS/GOenrich/TFBS_hormones_GO_BP.png", units = "cm", height = 15, width = 15)
p <- plot_resGO(resGO_MF)
p
ggsave("~/Desktop/Workspace/HDBR/sexDE_prenatal_brain/images/4-postprocessing/7-TFBS/GOenrich/TFBS_hormones_GO_MF.png", units = "cm", height = 12, width = 15)
p <- plot_resGO(resGO_CC)
p
ggsave("~/Desktop/Workspace/HDBR/sexDE_prenatal_brain/images/4-postprocessing/7-TFBS/GOenrich/TFBS_hormones_GO_CC.png", units = "cm", height = 12, width = 15)
```

```{r KEGG_enrich, echo=FALSE, eval=FALSE}
suppressPackageStartupMessages( library(AnnotationDbi) )
universe_bayes_ensembl <- as.character(bayesRes$ensemblID)
clusters_ensembl <- list("AR male-biased" = AR_prenatal_maleBiased_genesDE$ensemblID,
                         "AR female-biased" = AR_prenatal_femaleBiased_genesDE$ensemblID,
                         "ESR1 female-biased" = ESR1_prenatal_femaleBiased_genesDE$ensemblID)
# Get the Entrez gene IDs associated with the ensembl gene IDs
clusters_KEGG <- clusters_ensembl
for (listName in names(clusters_ensembl)){
  entrezID <- mapIds(org.Hs.eg.db, keys=clusters_ensembl[[listName]], column="ENTREZID", keytype="ENSEMBL", multiVals="first")
  clusters_KEGG[[listName]] <- unname(entrezID[!is.na(entrezID)])
}
universe_bayes_entrez <- unname(mapIds(org.Hs.eg.db, keys=universe_bayes_ensembl, column="ENTREZID", keytype="ENSEMBL", multiVals="first"))

resKEGG <- compareCluster(clusters_KEGG, fun = "enrichKEGG", universe = universe_bayes_entrez, organism = "hsa",
                          keyType = "ncbi-geneid", minGSSize = 10, maxGSSize = 500, pvalueCutoff = 1)
p <- plot_resGO(resKEGG)
p
ggsave("~/Desktop/Workspace/HDBR/sexDE_prenatal_brain/images/4-postprocessing/7-TFBS/GOenrich/TFBS_hormones_KEGG.png", units = "cm", height = 8, width = 12)
```


```{r DO_pathway_enrich, echo=FALSE, eval=FALSE}
resDO <- compareCluster(clusters_KEGG, fun = "enrichDO", universe = universe_bayes_entrez, ont = "DO",
                        organism = "hsa", minGSSize = 10, maxGSSize = 500, pvalueCutoff = 1)
#plot_resGO(resDO)

suppressPackageStartupMessages( library(ReactomePA) )
resPathway <- compareCluster(clusters_KEGG, fun = "enrichPathway", universe = universe_bayes_entrez, organism = "human",
                             minGSSize = 10, maxGSSize = 500, pvalueCutoff = 1)
plot_resGO(resPathway)
```



# Session Info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

