---
title: "Enrichment for transcription factor binding sites in promoter regions of shared and prenatal-specific sex-biased genes"
author: "Clara Benoit-Pilven"
date: "`r Sys.Date()`"
output: html_document
---

```{r library, echo=FALSE}
suppressPackageStartupMessages( library(rtracklayer) )
suppressPackageStartupMessages( library(data.table) )
```

```{r, setup, include=FALSE, echo=FALSE}
knitr::opts_knit$set(root.dir = '~/Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/')
```

Parameters
```{r thresholds, echo=FALSE}
threshold_qvalue <- 0.01
threshold_logFC <- 0
```

Functions
```{r function_save_bed_file, echo=FALSE}
save_bedFile <- function(data, fileName){
  bed <- data[, c("chrom", "promoter_start", "promoter_end", "gene_id", "gene_name", "strand")]
  colnames(bed) <- c('chrom', 'chromStart', 'chromEnd', 'name', 'score', 'strand')
  bed$score <- 0
  write.table(bed, file = fileName, quote = FALSE, sep = "\t", col.names = FALSE, row.names = FALSE)
}
```


# Generate bed files for Unibind analysis

## Load bayes genes

```{r load_bayes_results, echo=FALSE}
bayesResults <- read.table("data/4-GTEx/2-comparison/lineModel/lineModel_prenatalVSadult_6models_withAnnotation.txt", 
                           header = TRUE, sep = "\t")
```


## Load annotation and prepare promoters

```{r load_annotation, echo=FALSE}
gencode28_annot <- rtracklayer::import("/Users/benoicla/Desktop/Genome/Gencode_v28/gencode.v28.annotation.gtf.gz")
gencode28 <- gencode28_annot[which(elementMetadata(gencode28_annot)$type == "gene"),]
gencode28 <- data.frame(gene_id = elementMetadata(gencode28)$gene_id,
                    gene_name = elementMetadata(gencode28)$gene_name,
                    gene_type = elementMetadata(gencode28)$gene_type,
                    chr = seqnames(gencode28),
                    start = start(gencode28),
                    end = end(gencode28),
                    strand = strand(gencode28))
```

Prepare promoters.

```{r prepare_coord_prom, echo=FALSE}
promoters <- data.frame(gene_id = gencode28$gene_id, gene_name = gencode28$gene_name, chrom = gencode28$chr, strand = gencode28$strand, 
                        promoter_start = 0, promoter_end = 0)
promoters$gene_id <- sapply(strsplit(as.character(promoters$gene_id), "\\."), function(x){x[1]})
for (i in 1:dim(gencode28)[1]){
  if (gencode28$strand[i] == "+"){
    promoters$promoter_start[i] <- as.integer(gencode28$start[i]-2000-1) # -1 because bed files are 0 based
    promoters$promoter_end[i] <- as.integer(gencode28$start[i])
  } else if (gencode28$strand[i] == "-"){
    promoters$promoter_start[i] <- as.integer(gencode28$end[i]-1) # -1 because bed files are 0 based
    promoters$promoter_end[i] <- as.integer(gencode28$end[i]+2000)
  } else{
    print("Unknown strand...")
  }
}
```

## Get promoters for all genes analyzed

```{r save_bed_file_universe, echo=FALSE}
promoters_universe <- promoters[which(promoters$gene_id %in% bayesResults$ensemblID), ]
save_bedFile(promoters_universe, "data/5-postprocessing/7-TFBS/shared/universe_bayes_promoters.bed")
```


## Extract promoters of DE genes


Separate male/female-biased and shared/prenatal-specific genes (4 lists of genes).
```{r separate_DE_genes, echo=FALSE}
shared <- bayesResults[which(bayesResults$specificity == "Shared"), ]
prenatal <- bayesResults[which(bayesResults$specificity == "Prenatal"), ]
shared_maleBiased <- shared[which(shared$logFC_prenatal <= threshold_logFC), ]
shared_femaleBiased <- shared[which(shared$logFC_prenatal >= threshold_logFC), ]
prenatal_maleBiased <- prenatal[which(prenatal$logFC_prenatal <= threshold_logFC), ]
prenatal_femaleBiased <- prenatal[which(prenatal$logFC_prenatal >= threshold_logFC), ]
```

Select the promoters for the sex-biased genes.
```{r promoter_sexBiased, echo=FALSE}
promoters_shared_maleBiased <- promoters[which(promoters$gene_id %in% shared_maleBiased$ensemblID), ]
promoters_shared_femaleBiased <- promoters[which(promoters$gene_id %in% shared_femaleBiased$ensemblID), ]
promoters_prenatal_maleBiased <- promoters[which(promoters$gene_id %in% prenatal_maleBiased$ensemblID), ]
promoters_prenatal_femaleBiased <- promoters[which(promoters$gene_id %in% prenatal_femaleBiased$ensemblID), ]
```

Save these promoters as bed files.
```{r save_bed_files, echo=FALSE}
save_bedFile(promoters_shared_maleBiased, "data/5-postprocessing/7-TFBS/shared/shared_maleBiased_promoters.bed")
save_bedFile(promoters_shared_femaleBiased, "data/5-postprocessing/7-TFBS/shared/shared_femaleBiased_promoters.bed")
save_bedFile(promoters_prenatal_maleBiased, "data/5-postprocessing/7-TFBS/shared/prenatal_maleBiased_promoters.bed")
save_bedFile(promoters_prenatal_femaleBiased, "data/5-postprocessing/7-TFBS/shared/prenatal_femaleBiased_promoters.bed")
```


# Run analysis in bash

```{bash unibind_analysis, echo=FALSE, eval=FALSE}
# from /Users/benoicla/Desktop/Tools/unibind_enrichment/ directory
## analysis for shared male-biased genes
bin/UniBind_enrich.sh oneSetBg /Users/benoicla/Desktop/Tools/unibind_enrichment/data/hg38_robust_UniBind_LOLA.RDS /Users/benoicla//Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/data/5-postprocessing/7-TFBS/shared/shared_maleBiased_promoters.bed /Users/benoicla//Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/data/5-postprocessing/7-TFBS/shared/universe_bayes_promoters.bed /Users/benoicla//Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/data/5-postprocessing/7-TFBS/shared/Unibind_shared_maleBiased
## analysis for shared female-biased genes
bin/UniBind_enrich.sh oneSetBg /Users/benoicla/Desktop/Tools/unibind_enrichment/data/hg38_robust_UniBind_LOLA.RDS /Users/benoicla//Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/data/5-postprocessing/7-TFBS/shared/shared_femaleBiased_promoters.bed /Users/benoicla//Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/data/5-postprocessing/7-TFBS/shared/universe_bayes_promoters.bed /Users/benoicla//Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/data/5-postprocessing/7-TFBS/shared/Unibind_shared_femaleBiased

## analysis for prenatal male-biased genes
bin/UniBind_enrich.sh oneSetBg /Users/benoicla/Desktop/Tools/unibind_enrichment/data/hg38_robust_UniBind_LOLA.RDS /Users/benoicla//Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/data/5-postprocessing/7-TFBS/shared/prenatal_maleBiased_promoters.bed /Users/benoicla//Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/data/5-postprocessing/7-TFBS/shared/universe_bayes_promoters.bed /Users/benoicla//Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/data/5-postprocessing/7-TFBS/shared/Unibind_prenatal_maleBiased
## analysis for prenatal female-biased genes
bin/UniBind_enrich.sh oneSetBg /Users/benoicla/Desktop/Tools/unibind_enrichment/data/hg38_robust_UniBind_LOLA.RDS /Users/benoicla//Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/data/5-postprocessing/7-TFBS/shared/prenatal_femaleBiased_promoters.bed /Users/benoicla//Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/data/5-postprocessing/7-TFBS/shared/universe_bayes_promoters.bed /Users/benoicla//Desktop/Workspace/HDBR/github-repo/sexDE_prenatal_brain/data/5-postprocessing/7-TFBS/shared/Unibind_prenatal_femaleBiased
```


# Session Info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
